---
title: "A Short Overview"
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{A Short Overview}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette provides a short walk through of how to use adatutor for
classification tasks using Adaptive Boosting (AdaBoost). AdaBoost is an ensemble
learning method that improves weak classifiers by iteratively adjusting their
weights based on classification errors. Here, we apply AdaBoost to predict the
replication success of social science lab experiments using data from
@altmejd2019.

## Setup

Before we start, make sure the `adatutor` package is installed and loaded.

```{r}
#| label: load-package
# Load package
library(adatutor)
```

We will use a short version of the `altmejd` dataset (see the R Documentation
for more information). @altmejd2019 pooled the data from five large-scale
replication projects, which we will use to classify the replication outcomes of
social science lab experiment. The dataset can be loaded with:

```{r}
#| label: load-data
# Load data
data(altmejd)
```

### The Names of the Predictors and the Outcome Variables

We will use only a subset of the predictors used by @altmejd2019 -- the
*reviewer metrics*. Here are the names of the reviewer metrics:

```{r}
#| label: prednms
# Reviewer metrics (names)
prednms <- c("power.o", "effect_size.o", "n.o", "p_value.o")
```

Our outcome variable is a binary measure of replication success with the name:

```{r}
#| label: outnme
# Outcome variable (name)
outnme <- "replicate"
```

### The Variables of Interest

The predictors and outcomes are the variables of interest in our prediction task.

```{r}
#| label: voinms
# Variables of interest
voinms <- c(prednms, outnme)
```

For convenience, we abbreviate the original name of the dataset.

```{r}
#| label: shorten-df
# Shorten 'altmejd'
df <- altmejd
```

Setting a seed ensures reproducible results:

```{r}
#| label: seed
# Reproducibility
set.seed(112)
```

## Splitting the Data

To evaluate model performance, we split the data into training (70%) and test
(30%) sets.

### Simple Random Split

A standard random split:

```{r}
#| label: random-split
# Simple random split
randsplit <- partition(df, prop = 0.7)
train <- randsplit$set1
test <- randsplit$set2
```

### Stratified Split

Stratified sampling ensures proportional representation of the outcome variable:

```{r}
#| label: stratified-split
# Stratified split: the same call, told which variable to hold steady
stratsplit <- partition(df, prop = 0.7, strata = "replicate")
train <- stratsplit$set1
test <- stratsplit$set2
```

## Adaptive Boosting

The AdaBoost algorithm combines multiple base models into a strong overall model
-- the *ensemble*. Which base model is the analyst's choice, not the
algorithm's: AdaBoost is defined over any *weak learner* whose weighted error
stays below one half. That is why `adaboost()` takes a fitted learner rather
than a formula, and why its own signature mentions no trees at all:

The AdaBoost algorithm combines multiple base models into a strong overall model, which is referred to as *ensemble*. The base model is actually a free choice and originally the AdaBoost algorithm was defined over any *weak learner* whose weighted error stays below one half, but often depth-one decision trees (aka. *stumps*) are used for convenience. 

**TODO** AdaBoost is first and foremost a teaching package. That is why we mimic the signature of  


```r
adaboost(h, n_iter, eta)
```

### The Weak Learner

Here the weak learner is a classification stump, so that is what we fit first.
The tree hyperparameters belong to this call, and `adaboost()` carries them
into every boosting round: depth, the complexity parameter, the minimum split
and bucket sizes, and the splitting rule. It overrides only `xval` and
`maxsurrogate`, which are speed rather than structure.

Two settings cannot travel, and `adaboost()` refuses them rather than ignoring
them. A `prior` is recomputed from each round's observation weights -- that is
what boosting *does* -- and a `loss` matrix is not passed on, so a learner
carrying either is an error rather than a surprise.

`model = TRUE` is the one requirement. `rpart()` keeps the outcome but discards
the predictors unless you ask for them, and `adaboost()` has to refit this tree
on reweighted data, so it needs the data the tree was grown on.

```{r}
#| label: weak-learner
# Fit the weak learner: one stump
h <- rpart::rpart(
  replicate ~ .,
  data = train[, voinms],
  # Classification tree
  method = "class",
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  parms = list(split = "gini"),
  # Keep the training data, so the stump can be refitted
  model = TRUE
)
```

### Training Phase

What is left for `adaboost()` are the two hyperparameters that are genuinely
boosting's own:

- `n_iter`: Number of boosting rounds, one base model per round
- `eta`: Learning rate

Here, we develop an ensemble with 10 stumps and a learning rate of 1:

```{r}
#| label: train
#| results: hide
#| message: false
# Boosted stumps and weights
fit <- h |> adaboost(n_iter = 10, eta = 1)
```

Swapping the weak learner is now a change to the `rpart()` call above and
nothing else: a deeper tree, a different splitting rule or a different
complexity parameter reaches the ensemble without `adaboost()` gaining an
argument.

### Test Phase

Once the model is trained, we can use it to make predictions on the data in the
test set:

```{r}
#| label: predict
#| results: hide
#| message: false
# Predictions on test set
ypred <- predict(fit, newdata = test[, prednms])
```

### Model Evaluation

A confusion matrix can help us evaluate the prediction accuracy:

```{r}
#| label: confusion
# Confusion matrix
table(ypred, test[, outnme])
```

### Performance Measures

Class labels say only whether a prediction was right. The *margin*, the
weighted vote before its sign is taken, also says how sure the model was, and
every measure that reads a ranking needs that. So we score the margin:

```{r}
#| label: margin
# Margins on the test set
margin <- predict(
  fit,
  newdata = test[, prednms],
  type = "margin",
  verbose = FALSE
)
```

`assess()` computes the usual measures in one call:

```{r}
#| label: assess
# Performance measures
assess(test[, outnme], score = margin)
```

The printout shows the confusion matrix at a threshold of 0, then the measures
at that threshold. Next comes the precision among the top 3, the top 5 and the
top k scores, where k is the number of actual successes. Last are the areas
under the ROC and precision-recall curves, which use every threshold.
`?assess` lists the formulas and the other names each measure goes by (for
example, `sens` is also called recall).

## Leave-One-Group-Out Cross-Validation

A single split gives a single estimate. The five replication projects differ
in design and base rate, so we hold out one project at a time and train on the
other four: leave-one-group-out cross-validation (LOGO-CV), with the projects
as the groups.

`logo_cv()` takes the fitted ensemble, reads its settings (the formula, the
tree settings, `n_iter` and `eta`), and refits them once per project. It does
not use the fit's own data: cross-validation evaluates the procedure, not one
fitted model. The fit only carries the settings, so `verbose = FALSE` keeps it
quiet.

```{r}
#| label: logo-cv
# LOGO-CV: refit the same recipe with each project held out
res <- h |>
  adaboost(n_iter = 1000, eta = 1, verbose = FALSE) |>
  logo_cv(data = df, group = "pid")

# Results per project
res
```

Read the spread across projects, not an average. The projects differ in size
(10 to 90 studies), and they share most of their training data, so a mean and
its standard error would mislead.

`bootstrap()` resamples each held-out project to show how much its estimate
could move with different studies. `plot()` then draws the intervals as bands:

```{r}
#| label: logo-cv-bootstrap
#| fig-alt: "AUROC per replication project with bootstrapped interval bands"
# Bootstrapped intervals per project
set.seed(112)
res |>
  bootstrap(n_resample = 1000) |>
  plot()
```

To compare AdaBoost with another learner, pass a formula and a `model`
instead of a fit. Here is a logistic regression on the same predictors:

```{r}
#| label: logo-cv-logit
# LOGO-CV for a logistic regression
logit <- logo_cv(
  replicate ~ power.o + effect_size.o + n.o + p_value.o,
  data = df,
  group = "pid",
  model = "logit"
)
logit
```

## Feature Importance

AdaBoost is restricted to decision stumps here, so every tree in the ensemble
splits on exactly one variable. Each stump carries a voting weight (`a`) based
on its accuracy; grouping stumps by the variable they split on and summing
their weights gives a feature-importance score. AdaBoost sequentially
targets the hardest-to-predict observations by re-weighting the data, so a
variable that is frequently chosen -- especially by high-weight stumps --
is one that consistently helped correct the ensemble's biggest mistakes. Total
accumulated weight is therefore a direct measure of a feature's predictive contribution.

```{r}
#| label: importance
important <- gauge(fit)
head(important, 10)
```

```{r}
#| label: importance-plot
#| fig-alt: "Horizontal bar chart of AdaBoost feature importance"
plot(important, top_n = 15)
```

## References

::: {#refs}
:::
