---
title: "A Short Overview"
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{A Short Overview}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette provides a short walk through of how to use adatutor for
classification tasks using Adaptive Boosting (AdaBoost). AdaBoost is an ensemble
learning method that improves weak classifiers by iteratively adjusting their
weights based on classification errors. Here, we apply AdaBoost to predict the
replication success of social science lab experiments using data from
@altmejd2019.

## Setup

Before we start, make sure the `adatutor` package is installed and loaded.

```{r}
#| label: load-package
# Load package
library(adatutor)
```

We will use a short version of the `altmejd` dataset (see the R Documentation
for more information). @altmejd2019 pooled the data from five large-scale
replication projects, which we will use to classify the replication outcomes of
social science lab experiment. The dataset can be loaded with:

```{r}
#| label: load-data
# Load data
data(altmejd)
```

### The Names of the Predictors and the Outcome Variables

We will use only a subset of the predictors used by @altmejd2019 -- the
*reviewer metrics*. Here are the names of the reviewer metrics:

```{r}
#| label: prednms
# Reviewer metrics (names)
prednms <- c("power.o", "effect_size.o", "n.o", "p_value.o")
```

Our outcome variable is a binary measure of replication success with the name:

```{r}
#| label: outnme
# Outcome variable (name)
outnme <- "replicate"
```

### The Variables of Interest

The predictors and outcomes are the variables of interest in our prediction task.

```{r}
#| label: voinms
# Variables of interest
voinms <- c(prednms, outnme)
```

For convenience, we abbreviate the original name of the dataset.

```{r}
#| label: shorten-df
# Shorten 'altmejd'
df <- altmejd
```

Setting a seed ensures reproducible results:

```{r}
#| label: seed
# Reproducibility
set.seed(112)
```

## Splitting the Data

To evaluate model performance, we split the data into training (70%) and test
(30%) sets.

### Simple Random Split

A standard random split:

```{r}
#| label: random-split
# Simple random split
randsplit <- random_split(df, prop = 0.7)
train <- randsplit$set1
test <- randsplit$set2
```

### Stratified Split

Stratified sampling ensures proportional representation of the outcome variable:

```{r}
#| label: stratified-split
# Stratified split
stratsplit <- stratified_split(df, group = "replicate", id = "eid", prop = 0.7)
train <- stratsplit$set1
test <- stratsplit$set2
```

## Adaptive Boosting

The AdaBoost algorithm combines multiple base models into a strong overall model
-- the *ensemble*. We use classification stumps as base learners, which require
the following hyperparameter setup:

```{r}
#| label: treehypar
# Tree hyperparameter setup
treehypar <- rpart::rpart.control(
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  split = "gini")
```

### Training Phase

In addition to the tree hyperparameters, we need to set the boosting
hyperparameters. The key hyperparameters are:

- `T`: Number of base models
- `eta`: Learning rate

Here, we develop an ensemble with 100 stumps and a learning rate of 1:

```{r}
#| label: train
#| results: hide
# Boosted stumps and weights
fit <- trainAda(replicate ~ ., data = train[, voinms],
  # Boosting hyperparameter
  T = 10, eta = 1,
  # Tree hyperparameter
  treehypar = treehypar)
```

### Test Phase

Once the model is trained, we can use it to make predictions on the data in the
test set:

```{r}
#| label: predict
#| results: hide
# Predictions on test set
ypred <- testAda(fit, data = test[, prednms])
```

### Model Evaluation

A confusion matrix can help us evaluate the prediction accuracy:

```{r}
#| label: confusion
# Confusion matrix
table(ypred, test[, outnme])
```

## Feature Importance

AdaBoost is restricted to decision stumps here, so every tree in the ensemble
splits on exactly one variable. Each stump carries a voting weight (`a`) based
on its accuracy; grouping stumps by the variable they split on and summing
their weights gives a feature-importance score. AdaBoost sequentially
targets the hardest-to-predict observations by re-weighting the data, so a
variable that is frequently chosen -- especially by high-weight stumps --
is one that consistently helped correct the ensemble's biggest mistakes. Total
accumulated weight is therefore a direct, mathematically grounded measure of a
feature's predictive contribution.

```{r}
#| label: importance
important <- adaImport(fit)
head(important, 10)
```

```{r}
#| label: importance-plot
#| fig-alt: "Horizontal bar chart of AdaBoost feature importance"
plotAdaImport(important, top_n = 15)
```

*TODO*: Rename function to `calculate_importance()` or `get_importance()` 
*TODO*: Rename function to `plot_importance()`
*TODO* rename function to `train_adaboost()` and `test_adaboost()`?

## References

::: {#refs}
:::
