---
title: "Module 3: Adaptive Boosting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{module3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```

# Module 3: AdaBoost

First we fit the model to the data (training) and then test its predictive
capabilities on unseen data (testing). Since we will use the data to find
the best hyperparameters this is better thought of as validation set.

- Discuss a couple of iterations of boosting (Kunapali, p.90) 

- (Extra: Decision stumps)

## An iterative process...

Setup.

```{r}
n_trees <- 100
# Extract the number of rows
n_samples <- nrow(X_train)
# Initialize the weights
D <- rep(1, n_samples)/n_samples
# Pre-allocate the empty ensemble
H <- vector("list", n_trees)
df <- data.frame(X_train, y_train)
```


### Adaboost continued

Normalized weights.

```{r}
# Initialize the weights (normalized)
D <- rep(1, n_samples)/n_samples
```

Adaptive boosting.

```{r}
# for (tree in 1:n_trees) {
  # Train a single weak learner (h_t) with normalized(!) sample weights (D)
  h <- rpart::rpart(formula = y ~ ., data = df , method = "class",
                    weights = D, control = tree_hyperpar)
  # Make predictions with the weak learner (h_t)
  y_pred <- predict(object = h, newdata = X, type = "class")
  # y_pred <- 2*as.numeric(y_class) - 3
  wacc <- sum(D * (y == y_pred)) / sum(D)
  # Compute the training error (e_t) of the weak learner (h_t)
  # This is simply one minus the weighted accuracy, given weights D
  e <- 1 - wacc
  # Equivalent but less readable: e <- sum(D * (y != y_pred)) / sum(D)
  # Compute the the weight (a_t) of the weak learner (h_t)
  # Reduce the learner weight by a factor eta (learning rate)
  a <- 0.5 * log( (1 - e)  / e) * eta
  # Pre-Prunning or Early Stopping:
  # Do not forget to add the parameter (prunning threshold)
  # if (a < prunning_threshold) {
  #  message("Prunning threshold reached")
  #  break
  # }
  # Updates the weights
  # Increase for misclassified examples: (y == y_pred)
  # decrease for correctly classified examples: (y!=y_pred)
  m <- (y == y_pred) * 1 + (y != y_pred) * -1
  # Update the sample weights and limit the contribution of h_t by eta
  D <- D * exp(-a * m)
  # Normalize weights (probabilistic interpretation)
  D <- D/sum(D)
  # Save the weak learner and its weight
  H[[tree]] <- list("learner" = h, "learner_weight" = a,
                    "observation_weight" = D)
  # Store sample weights for visualization
}
  # Return the ensemble
# H
}
```

```{r}
# Plot
# rattle::fancyRpartPlot(h, caption = NULL)
# Sys.sleep(2)
```


## Hands dirty: Implementation

### Training 

Look under the hood of the function 

```{r}
train_adaboost
```

Tree hyperparameter settings

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 
```


### Prediction

```{r}
test_adaboost
```

Make predictions with the trained model and add them to the testing set.

```{r}
y_class <- test_adaboost(X_test, ensemble)$class
test$y_pred <- factor( (y_class + 3) / 2, labels = c("no", "yes") )
```

## Evaluation

```{r}
ROC <- AUC::roc(test$y_pred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```
