---
title: "Module 1: Splitting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{module1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```

# Module 1: Splitting 

Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")

# Outcome (classification task) 
target <- "replicated"

# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")

# Note: No missing values 
df <- altmext723[,voi]
```

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

### Simple random sampling 

```{r}
# Reproducibility
set.seed(112)

# Split proportion
sprop <- 0.7

# Number of effects 
n_effect <- nrow(df)

# Number of training samples
n_train <- round(n_effect * sprop)

# Training index
train_idx <- sample(seq_len(n_effect), n_train)

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

### Stratified sampling & imbalances

However, there are imbalances in the data set. 

- What is and why use stratified sampling - with the outcome as stratum? 
"Recall that stratified means the folds are created in such a way that the class
distribution is preserved across the folds. This also helps with imbalanced data
sets, as stratification ensures that data from all classes is represented." (p. 103)


```{r}
table(df$replicated)
```

Therefore we should do a stratified split. This means we need to partition the
data into groups based on a stratum variable - here: replicated, and then draw
a random sample from each stratum.

```{r}
# Reproducibility
set.seed(112)

# Partition the data based on the outcome variable
df_split <- split(df, df$replicated)

# Number of training samples (stratified)
n_train <- lapply(df_split, function(df) round(nrow(df) * sprop))

# Training index (stratified)
strat_idx <- mapply(function(df, n) sample(rownames(df), n), df_split, n_train)

# Training index 
train_idx <- as.numeric(unlist(strat_idx))

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

Now the training and testing set have approximately the same proportion of
replication and failure as the original outcome variable. We could also use the
`rsample` package for that.

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```


