---
title: "Module 4: Leave-Subject-Out Cross-Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{module4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```

TODO: Check if code in Module 2 is running

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 5,  eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

This will run for approximately 4 minutes.

TODO: This seems to run, but why do I always get the same performance estimates?
No matter if I use 1 tree or 1e4 trees...

```{r}
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 500, eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

## Performance estimates

```{r}
auc
```
