---
title: "Module 2: A primer on classification stumps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{module2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```

```{r, include=FALSE}
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")
target <- "replicated"
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")
df <- altmext723[,voi]
set.seed(112)
sprop <- 0.7
n_effect <- nrow(df)
n_train <- round(n_effect * sprop)
train_idx <- sample(seq_len(n_effect), n_train)
train <- df[train_idx, ] ; test  <- df[-train_idx, ] 
X_train <- train[,features] ; X_test <- test[,features] 
y_train <- train[,target] ; y_test <- test[,target]
```

# Module 2: A primer on classification stumps 

In Module 3, we show how the AdaBoost algorithm transforms classification stumps
or shallow classification trees into a strong ensemble machine learning model.
Thus, a prior study of classification trees seems mandatory. However, Module 2
is not a thorough introduction to decision trees (see, e.g., Pargent et al. xxxx
for a more complete introduction or Breiman et al. xxxx for a complete one). The
focus of this module will be entirely on the details necessary to understand
AdaBoost. We will use the `rpart` package (Ref.) for hands-on understanding and
`rpart.plot` (Ref.) for visualizing the results.

##### Classification problems

The `altmext723` data set contains measurements on the power of the original
study, its effect size, the number of observations are available, and more.
Typing `?altmext723`, provides access the full variable list as well as
additional information on each variable. Given a set of measurements, a general
goal of predictive modeling is to extract a pattern from the data that allows
accurate predictions of future outcomes. More specifically, in the `altmext723`
data set, we want to learn a *general* pattern between predictors or *features*
such as `power.o`, the power of the original study, and `replicated`, the
*outcome* variable that allows us to accurately predict whether future or unseen
replication attempts will succeed (class 1) or fail (class 2). Given the number
of classes, the scenario can be more accurately categorized as a two-class
*classification problem*. The presence of a labeled outcome that can guide the
model's learning process indicates that we are in the realm of *supervised
learning*. Note that the phrase "general pattern" is important here because it
makes it clear that the goal is to go beyond the data at hand. Put another way,
in this tutorial, and in predictive modeling in general, the goal is to create a
*predictive model* rather than a *retrodictive* one. This means we cannot just
evaluate the fit or in-sample performance of our model. We also need to consider
its out-of-sample or predictive performance. Therefore, in Module 1, we set
aside a subset of the data, the test set, to evaluate the model's predictive
ability. We will explicitly test the out-of-sample perfprmance of AdaBoost in
Module 3 and present another strategy in Module 4. But first, focus must be on
understanding how trees address the classification problem mentioned above.

<!-- Note to self: Actually, this will be a validation set, since we tune the -->
<!-- hyperparameter with these stuff. Write this problem somewhere, that the data set -->
<!-- is to small to always set aside some portion of the data. A solution will be -->
<!-- discusses with cross-validation, although this also provides just an estimate -->
<!-- that differs from the true generalization performance. -->

### Classification trees

Binary *classification trees* learn patterns by repeatedly asking (yes/no)
questions about features of the data allowing for organizing the data points
into non-overlapping groups (aka., *partitions*). An example question to split
the data could be: Is the power of the study less than 0.93002? If it is, the
study is placed in the left group; if it is not, it is placed in the right
group. Following the given data *splitting rule* ultimately results in two
non-overlapping groups; one containing studies with `power.o > 0.93002` and
another with `power.o <= 0.93002` (see Figure 1).

With a second question the classification tree could further partition the data
at a lower level. At depth two, an example question targeting the high power
studies (`power.o >= 0.93002`) could be: Is the number of observations in this
partition also greater than 100?". If it, the study is placed in the lower left
and if not in the lower right group. In this way the trees continue to grow.
That is, the three continues the grouping process until the data cannot be
divided any further or a termination criterion is reached. Termination criteria
are, for example, a minimum number of studies in the final group, or a minimum
number of studies to consider a split (see `?rpart.control` for all available
options in `rpart`).

```{r, echo = FALSE, fig.cap="Visualization of a classification stump"}
# Extract only two features
# Note: Reduces run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <-  train[,union(subfeat, target)]
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
# Create a decision tree 
h <- rpart::rpart(formula = replicated ~ ., data = subdf,  method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Classification Stump", extra = 102, 
                       digits = 5, box.palette="BuGy")
```

#### Classification stumps

Binary *classification stumps* shorten the repetetive process of classification
trees by limiting themselves to a single (yes/no) question, such as `power.o <
0.93002`?. Consequently, they partition the data into two final groups,
terminating at a maximum depth of one (see Figure 1 and 2). To compare the
results of the partitioning process to those of a Bernoulli-logit generalized
linear model see the Appendix. Calling the starting point at the top in Figure 1
(left) the root of the tree or its *root node* and the resulting group at the
terminal point its *leave nodes*, a classification stump can also be categorized
as a tree with one root and two leave nodes (see, e.g. ref.).

```{r, echo = FALSE, fig.cap="Decision boundary of a classification stump"}
# Number of classes
k <- length(unique(subdf$replicated))
# Range(s) as list
rng <- lapply(subdf[,subfeat], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Stump")
points(grid, col = grid_colors, pch = ".")
points(subdf[,subfeat], col = symbol_colors, pch = as.integer(y_train) + 15)
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)
```


The final decision on whether studies with a certain feature, such as high
power, are considered successful or not is usually made by majority vote based
on the frequency of labeled studies in the leave node of the tree. For example,
in the classification stump in Figure 1, there are more failed replication
attempts for studies in the left branch, so the overall classification for
studies with the feature `power.o < 0.93002` is "failure". In the right branch,
there are more successes, so studies with `power.o >= 0.93002` are considered
successful. To verify this mechanically, use the code in Code Chunk 1.

```{r Code Chunk 1}
# Code Chunk 1
# Define the condition for the split
splitcond <- train$power.o < 0.93002
# Studies with df$power.o < 0.93002 
table(train$replicated[splitcond])
# Studies with df$power.o >= 0.93002 
table(train$replicated[!splitcond])
```

In the first line of the code, we define the split boundary using the base R
relational operator `<`. This results in a logical vector whose `TRUE`, `FALSE`
values indicate the entries in the variable `power.o` that are greater than
0.93002. We then create two contingency tables for the factor variable
`replicated`, selecting only those cases that either meet the condition or do
not (see `?"!"` for more information on base R's logical negotiation operator).
The output shows the number of successes and failures under each condition,
which can be added together to reproduce the numbers of the classification stump
in Figure 1.

### Fit and visuaize a classification stump

There are several ways to fit and visualize classification trees. In the
tutorial, we will use the `rpart` package for fitting and the `rpart.plot`
package for visualizing the trees. Specifically, we will use the functions
`rpart()`, `rpart.control()` and `rpart.plot()`. Each function is outlined in
turn. For more information, however, see the R documentation, e.g.
`help("rpart", package = "rpart")` or `?rpart::rpart`.

```{r Code 2}
# Code Chunk 2
# Hyperparameter setup for a classification stump
hyptree <- rpart::rpart.control(
  maxdepth = 1, # Maximum depth
  split = "gini") # Gini impurity (default)
# Fit a classification stump
cstump <- rpart::rpart(formula = replicated ~ ., 
                       data = train[,union(features, target)],
                       method = "class", control = hyptree)
```

In Code Chunk 2, we first focus on `rpart()`. "rpart" is short for "recursive
partitioning". "Recursive partitioning" refers to the process of repeatedly
putting the data into non-overlapping groups, which we mentioned above. The
function takes several arguments (see `?rpart::rpart`), but we focus mainly on
`formula`, `data` and `control`. The `method` argument with the value `class`
is skipped, because we use it just to make explicit that we deal with a
classification problem. We start with passing a formula to `rpart()` that has
the general form: `target ~ features`. One could read this as "the outcome or
target (replicated) is a function of the features or predictors. All variables,
the features and the target, must be part of the data frame that we pass as
value to `rpart` with the `data` argument. The dot `.` in the formula, tells
`rpart()` to consider all variables in the data set, despite the one on the left
hand side, here `replicated`.

The `rpart.control()` function allows to specify the tree hyperparamters. In
general, *hyperparameter* are external parameter that we as the user set
*before* fitting the model and which influence the model's complexity or
learning process. Psychologists usually learn about hyperparameters in the
context of (non-Bayesian, exploratory) factor analysis, for example, where they
have to decide on a number of factors to retain before being able to evaluate a
particular solution. We characterized stumps as  primitive classification trees
with maximum depth equal to one. Consequently, we must limit the complexity of
the model using `maxdepth = 1` to avoid the tree growing beyond a stump. Put
another way, the constraint limits the number of data partitions, There is one
group with studies `power.o < 0.93002` and one with studies `power.o >= 0.93002`.

```{r, eval=FALSE}
# Code Chunk 3
# Visualize a classification stump
rpart.plot::rpart.plot(cstump, main = "Classification Stump", extra = 102,
                       digits = 5)
```

We show how to visualize a fitted classification stump in Code Chunk 3 using the
`rpart.plot()` function from the `rpart.plot` package. The code can also be used
to reproduce Figure 2. The `rpart.plot()` function takes an `rpart` model
object, such as the fitted classification stump `cstump`, and creates a
flow-chart like(?) plot (more precise) of it. There are several arguments to
customize the appearance of the plot (see `?rpart.plot::rpart.plot` for
available options). However, if a custom plot is required that goes beyond the
capabilities of `rpart.plot()`, we recommend switching to `rpart.plot`'s `prp()`
function. This function allows for fine-tuning the appearance of any decision
tree plot.

If working with (tree) hyperparameters still seems unfamiliar or if the goal is
to solidify and extend understanding of the discussed concepts, we recommend
doing Additional Exercise 1, visualizing the results using Code Chunk 2.

#### Additional Exercise 1 TODO

Access the `rpart.control()` documentation with `?rpart.control` and get an
overview over all hyperparameters of the `rpart` package. Increase the maximum
depth of the stump to two and three to create shallow classification trees.
Then, set the minimum number of observations in the terminal node to 1. How will
these setting influence the shape of the tree. Think about the solution first,
before running the code. (Note to self: Do the results make sense? Nope.)

### Gini impurity

Binary classification stumps learn by asking a single (yes/no) questions about
a data feature, that allows them to organize the data into two partitions. This
characterization, however, lacks a clear answer on how a particular data feature
is chosen for the split. So the question is: why `power.o > 0.93002`?

There are multiple strategies to an optimal splitting rule (see, e.g. Ref). The
strategy used in the classic AdaBoost algorithm can be found the belt of the
trees that are boosted. be traced back to Breiman and Colleagues CART algorithm.

is to split the data in such
a way that they minimize the misclassification risk (Ref. "The best splits would
then be those that most reduced the estimated misclassification rate, p.94)"

The general idea: Node impurity

Metaphorically speaking, the goal is to provide a clear picture for an
uncontroversial decision process. Technically, this is done by increasing the
homogeneity of the labels at a particular node. The more homogeneous the node,
that is, the less diverse the mix of examples labeled as "success" or "failure"
the purer the node and the less controversial the final decision for "success"
or "failure" in the majority vote. Put another way, by increasing the
homogeneity we reduce the node *impurity*. In this logic an optimum is reached
if the examples in the node belong to the same class (e.g., success). Thus, a
uniform class distribution implies the minimum or zero impurity, (maximal
purity). To stick with the metaphor, this is a clear picture for the decision
process. In contrast, maximal impurity, the blurred picture, is obtained by a
fifty-fifty split. In a two class classification problem the maximum impurity
can (thus) be one half. As a consequence, if classification is easier when the
impurity is low, by minimizing the impurity we reduce the risk
misclassification. The *Gini index* "Gini index of diversity as a measure of
node impurity; i.e (p.38)". The final measure of optimizing node impurity as by
the Gini index is refered to as  *Gini impurity*. The Gini impurity considers
how often we would misclassify a randomly chosen example from our data set if we
were to guess its class label based on the class distribution in that group.
Check: Applied to our example, this means that the Gini impurity measures how
often we are expected to misclassify a replication study's outcome (success or
failure) if we were to randomly assign a label (success or failure) based on the
distribution of successes and failures within our group of replication studies
-- split according to a specific power threshold. In our example the threshold is power.o >= 0.93.

#### Why `power.o > 0.93`?

When you're building a decision tree, you calculate the Gini impurity for each
possible split (question) you could make on the data. The question that leads to
the lowest Gini impurity (the purest resulting groups) is the one you choose.

There are only continuous features in the data set. The power of the original
study is one of them. For continuous features, the split point is chosen by
sorting the individual values and then calculating the Gini impurity for each
mid-point between adjacent values.

```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
```
We want to reproduce the results in the graphic. Therefore we will calculate
the Gini impurity for a split with the criterion `power.o < 0.9288`.


$$I_G(D) = 1 - \sum_{j=1}^K p^2_i$$

Since we have data with examples from 2 classes the formula simplifies as

$$I_G(D) = 1 - \sum_{j=1}^2 p^2_i =  1 - (p_1^2 + p_2^2)$$

where we calculate as $p_i = \frac{f_i}{n_i}$.

Weighted Gini impurity given feature X for i = 1,2 partitions.

$$I_{G\mid X}(D) = \sum_{i=1}^2 \frac{n_i}{n} I_G(D_i) = \frac{n_1}{n} I_G(D_1) + \frac{n_2}{n} I_G(D_2)$$

```{r}
# Example above:
constant <- adjacent_means[69] # 0.9300241
N_total <- length(y_train) # N_total: 107

# Left leaf (power.o < 0.9300241: yes)
criterion_left <- df$power.o < constant
N_left <- length(df$y_train[criterion_left]) # 69 total
left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1) # 19 successes
left_failure <- N_left - left_success # 50 failures

# Right leaf (power.o < 0.9300241: no)
criterion_right <- df$power.o >= constant
N_right <- length(df$y_train[criterion_right]) # 38 total
right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) # 29 successes
right_failure <- N_right - right_success # failures
```


```{r}
# Individual Gini impurities
gini_left <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
gini_right <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
```

Gini left: The Gini impurity suggests that if we were to randomly pick a
replication from our sample where the original had a power of less than 0.93 and
say that it would fail to replicate, there is approximately a 39.91% chance that
we would be wrong.

Gini right: The Gini impurity suggests that if we were to randomly pick a
replication from our sample where the original had a power of greater than 0.93 and
say that it would successfully replicate, there is approximately a 36.15% chance
that we would be wrong.

We finally weight the results to get the overall weighted Gini impurity
considering the different class distributions.

```{r}
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) * gini_left + (N_right / N_total) * gini_right
round(gini_impurity, digits = 5)
```


G = 0 indicates a perfectly pure node (all elements belong to the same class).
G = 0.5 indicates maximum impurity (elements are evenly distributed among all classes).

Estimating the Gini impurity

Gini impurity = 1 - (probability of the most frequent class)^2 - (probability of
the second-most frequent class)^2 - ... (and so on for all classes)


Gini Impurity in Action

Group 1 (e.g., power > 0.8): What's the Gini impurity of this group in terms of
replication success vs. failure labels? Are the studies in this group mostly
successful replications, mostly failures, or a mix?

Group 2 (e.g., power ≤ 0.8): Calculate the Gini impurity for this group as well.

Choosing the best Split: The algorithm selects the power threshold split that
results in the lowest overall Gini impurity in the new groups. This split
maximizes class purity, meaning it clearly separates studies more likely to
succeed in replication from those more likely to fail.






For example, lets say we have



1. step gini_d1 and gini_d2: You're correctly calculating the standard Gini
impurity for the left and right groups separately using the formula: 1 -
(probability of class 1)^2 - (probability of class 2)^2 - ...

2. step (N_left / N_total) and (N_right / N_total):  These represent the weights
assigned to each group's Gini impurity, based on the proportion of examples in
each group relative to the total.

3. step: gini_impurity:  The final line combines the weighted Gini impurities of both
groups to calculate the overall weighted Gini impurity for the potential split.

TODO: Calculations seem to be different from, e.g.: https://machinelearningmastery.com/classification-and-regression-trees-for-machine-learning/

Additional exercise: We leave it as an additional exercise to the reader to
verify that the result has the lowest Gini impurity among all splits (a) within
the particular features and (b) overall.


Important: For a two class classification problem.

```{r}
# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01)  # Values for fraction of class k
y <- 1 - (x^2) - (1-x)^2      # Gini Impurity calculation

# Create the plot
plot(x, y, type = "l",  # Specify line plot
     xlab = "Fraction of Class k (pk)",
     ylab = "Impurity",
     main = "Gini Impurity")

mtext("2 class classification problem")
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1))

# Optional: Add gridlines
abline(v = seq(0, 1, by = 0.1), h = seq(0, 1, by = 0.1), col = "lightgray")

```



# Control calculations






```{r, eval=FALSE}
subdf <- df[, c("power.o", "n.o", "y_train")]
# Hyperparameter setup
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  maxdepth = 3,
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion
  split = "gini") # or ID3: split = "information"
  # Create a decision tree
h <- rpart::rpart(formula = y_train ~ ., data = subdf , method = "class",
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Classification Tree", extra = 102,
                       digits = 5, box.palette="BuGy")

# With minbucket = 4 i can reproduce the following values.
# Check: Does this make sense? Maybe ignore the decision tree and just use
# decision stump. Or describe that the last split is nonsensical.

table(subdf$y_train[subdf$power.o < 0.93002 & subdf$n.o >= 24])
table(subdf$y_train[subdf$power.o < 0.93002 & subdf$n.o < 24])
table(subdf$y_train[subdf$power.o >= 0.93002 & subdf$power.o >= 1])
table(subdf$y_train[subdf$power.o >= 0.93002 & subdf$power.o < 1])
```

```{r, eval=FALSE}
# Number of classes
k <- length(unique(subdf$y_train))
# Range(s) as list
rng <- lapply(subdf[-length(subdf)], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f)
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq),
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Tree")
points(grid, col = grid_colors, pch = ".")
points(subdf[-length(subdf)], col = symbol_colors, pch = as.integer(y_train) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2,
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16),
     col = c(col2, col1), cex=0.7)

```

Decision stumps simplify this repetitive partitioning process by asking only a
single question about some feature of the data. For example, is the power of the
original study greater than . Classification stump is a classification tree of
depth 1.


One question that is left unanswered now, is how do we get to a particular
question that allows us to parition the data in the way we've seen?

- Splitting criterion
- Focus on CART with Gini impurity

A selection of most important tree hyperparameters for our AdaBoost
implementation.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  maxdepth = 1, # Boost stumps
  # Original CART with gini impurity as splitting criterion
  split = "gini") # or ID3: split = "information"
```

Fit a decision stump to the training data and visualize the result.

```{r}
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class",
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump", extra = 102, digits = 5)
```

- What do we see?
  - Why `digits =5`?

- Why this particular split?
  - Gini impurity
  - Sensitivity to observation weights (Assumption: equal)

Control calculations

```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
gini_impurity <- vector(length = length(adjacent_means))
  for(i in 1:length(adjacent_means) ) {
  constant <- adjacent_means[i]
  N_total <- length(y_train)
  # Left leaf
  criterion_left <- df$power.o < constant
  N_left <- length(df$y_train[criterion_left])
  left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- df$power.o >= constant
  N_right <- length(df$y_train[criterion_right])
  right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
  gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
  # Impurity
  gini_impurity[i] <- (N_left / N_total) * gini_d1 + (N_right / N_total) * gini_d2
  }
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjacent_means[which.min(gini_impurity)]
```


#### Sensitivity to observation weights

Normalized initialization weights. Should not change anything to before.

```{r}
# Equal weight to each observation
D <- rep(1, n_samples)/n_samples
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class",
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

Normalized random observation weights for demonstration. Changes result.

```{r}
# Reproducibility
set.seed(113)
# Random draws from Uniform(0,1)
D_unorm <- runif(107)
# Normalize (sum: 1)
D <- D_unorm/sum(D_unorm)
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class",
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

##### Additional exercises

`rpart` offers a number of hyperparameters that affect the outcome of the
results. We leave it to the reader as an additional exercise to explore how each
of them influences the result. For our purposes, we will focus only on the
maximum depth and the splitting criterion.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree (root node: depth = 0)
  # Important: We usually boost stumps (depth = 1) or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted
  minsplit = 20,
  # Minimum n° of obs in any terminal <leaf> node
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations
  xval = 10,
  # Original CART with gini impurity as splitting criterion
  split = "gini" # or ID3: split = "information"
  )
```

##### Additional exercise 1

We created a decision stump using the gini impurity as splitting criterion.
Adapt the code above to fit a shallow tree of depth two and three.

##### Additional exercise 2

Despite the hyperparameters we already mentioned there are more. See the
`?rpart::rpart.control` documentation and explore how changing different
arguments influence the results.


