---
title: "Supplemental Materials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{supplemental}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```

# Learning the weights across iterations (Daniel)

## Data

```{r}
# Load the data
data("altmext723")
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")
# Outcome (classification task) 
target <- "replicated"
# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")
# Note: No missing values 
df <- altmext723[,voi]
# Reproducibility
set.seed(112)
# Partition the data based on the outcome variable
df_split <- split(df, df$replicated)
# Split proportion
sprop <- 0.7
# Number of training samples (stratified)
n_train <- lapply(df_split, function(df) round(nrow(df) * sprop))
# Training index (stratified)
strat_idx <- mapply(function(df, n) sample(rownames(df), n), df_split, n_train)
# Training index 
train_idx <- as.numeric(unlist(strat_idx))
# Training set 
train <- df[train_idx, ]
# Feature matrix
X_train <- train[,features] 
# Target vector
y_train <- train[,target]
# Hyperparameter settings (tree) 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
# Fit the model to the data
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree)
```

Visualize the results.

```{r}
a <- sapply(ensemble, "[[", "learner_weight")
D <- t(sapply(ensemble, "[[", "observation_weight"))
plot(a, type = "l", ylim = c(0, .04), ylab = "Weights", xlab = "Trees", col = "red")

for(i in 1:ncol(D)) 
  lines(1:nrow(D), D[,i] + runif(1, -.00014, +.00014), col = adjustcolor(1, alpha.f = 0.1))

# DANIEL: why do the weights of the trees have these regular peaks for the sequential updates?
```
 
Daniel: Das Updating der Gewichtung der Beobachtungen pro Tree könnte man auch
illustrieren durch eine Art Heat Map (x = Trees, y = Observations, degree of
gray = weighting). Direkt untendrunter könnte man mit einer ähnlichen
graphischen Matrix die correct predictions als black/white Heatmap illustrieren.

# Decision Boundaries 

Visualize decision boundaries: 
- https://rpubs.com/ZheWangDataAnalytics/DecisionBoundary
- https://quantifyinghealth.com/plot-logistic-regression-decision-boundary-in-r/
- https://www.spsanderson.com/steveondata/posts/2023-09-11/index.html

## Data

```{r}
data("altmext723")
# Variables of interests
voi <- c("power.o", "effect_size.o", "replicated")
# Predictors 
features <- c("power.o", "effect_size.o")
# Outcome
target <- "replicated"
# Subset of the data  
d <- altmext723[, voi]
# Feature matrix
X <- d[,features]
# Outcome vector
y_obs <- d[,target]
```

## Bernoulli-logit GLM  

Prepare the visualization.

```{r}
# Number of classes
k <- length(unique(y_obs))
# Range(s) as list
rng <- lapply(X, range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$effect_size.o[1], rng$effect_size.o[2],
              length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "effect_size.o" = x2_seq)
# Fit a logitstic regression model to the data 
mdl <- glm(replicated ~., data = d, family=binomial(link='logit'))
# Change class attribute
class(mdl) <- c("lr", class(mdl))
# Predicted probability of replication success
y_prob <- predict(mdl, grid, type = "response")
# If p > 0.5 "replicated", else "not replicated"
y_pred <- factor(as.integer(y_prob > 0.5))
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "no" means "not replicated", "yes" means "replicated"
symbol_colors <- ifelse(y_obs == "failuire", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == 0, col1_transp, col2_transp) 
# Plot the features
```

Visualize.

```{r}
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
     xlab = "Power", ylab = "Effect Size", main = "Decision Boundary")
mtext("Bernoulli Logit Generalized Linear Model")
points(grid, col = grid_colors, pch = ".")
points(X, col = symbol_colors, pch = as.integer(y_obs) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
        col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("replicated", "not replicated") , pch = c(17,16), 
       col = c(col2, col1), cex=0.7)
```

## Random Forest

Prepare the visualization.

```{r}
# Number of classes
k <- length(unique(y_obs))
# Range(s) as list
rng <- lapply(X, range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$effect_size.o[1], rng$effect_size.o[2], 
              length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "effect_size.o" = x2_seq)
# Fit the model to the data
mdl <- randomForest::randomForest(replicated ~ ., data=d)
# Predict grid
y_pred <- predict(mdl, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_obs == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
```

Visualize.

```{r}

plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
     xlab = "Power", ylab = "Effect Size", main = "Decision Boundary")
mtext("Random Forest Algorithm")
points(grid, col = grid_colors, pch = ".")
points(X, col = symbol_colors, pch = as.integer(y_obs) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
        col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("replicated", "not replicated") , pch = c(17,16), 
       col = c(col2, col1), cex=0.7)
```

## ADAboost

Prepare the visualization.

```{r}
# Make 0/1 factor into something ADA can handle (-1/1) 
y_obs_ml <- 2 * as.numeric(y_obs) - 3
# Number of classes
k <- length(unique(y_obs))
# Range(s) as list
rng <- lapply(X, range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$effect_size.o[1], rng$effect_size.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "effect_size.o" = x2_seq)
# Make grid
# Tree hyperparameter setting 
tree_hyperpar <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 3,
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 5, 
  # Minimum n° of obs in any terminal <leaf> node 
  # Default: round(minsplit/3)
  minbucket = round(5 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
# Fit ADAboost
estimators <- train_adaboost(X, y_obs_ml, 
                           # Number of trees
                           n_trees = 5e2, 
                           eta = 1, 
                           tree_hyperpar)
# Predict ADAboost
y_pred_ml <- test_adaboost(grid, estimators)
# Make prediction 0/1 instead of -1/1 
y_pred <- as.factor((y_pred_ml$class + 1) / 2)
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_obs == "failure", col1, col2)
grid_colors <- ifelse(y_pred == 0, col1_transp, col2_transp)
```

Visualize.

```{r}
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
     xlab = "Power", ylab = "Effect Size", main = "Decision Boundary")
mtext("Adaptive Boosting Algorithm")
points(grid, col = grid_colors, pch = ".")
# points(grid, col = as.integer(y_pred) + 6L, pch = ".")
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# For the black line 
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
        col = "black", levels = (1:(k-1))+.5)
points(X, col = symbol_colors, pch = as.integer(y_obs) + 15)
legend("topleft", c("replicated", "not replicated") , pch = c(17,16), 
       col = c(col2, col1), cex=0.7)
```
