---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```

# Utility Functions

Majority vote

```{r}
calc_mode <- function(x) {
   ux <- unique(x)
   ux[ which.max( tabulate(match(x, ux)) ) ]
}
```

Performance evaluation

```{r}
calculate_f1 <- function(confusion) {
  if(length(confusion) > 2) {
    TP <- confusion[1, 1]  
    FP <- confusion[1, 2]  # False Positives
    FN <- confusion[2, 1]  # False Negatives
    precision <- TP / (TP + FP)
    recall <- TP / (TP + FN)
    # F1 Score
    return(2 * (precision * recall) / (precision + recall))
  } else {
    return(NA)
  }
}
```

## Dataset

Full dataset from Altmejd and colleagues plus the SSRP data

```{r}
# Load the data
# data("altfull723") Switched after discussion with Daniel
# data("altmext723")
load("altmext723.rda")

# Project ID
# altfull723$pid <- sub("\\..*", "", as.character(altfull723$id))
# Rename ID to "eid", i.e., "Effect ID"
# TODO: Change in the dataset
# names(altfull723)[1] <- "eid"
# altfull723$log_p.o <- log(altfull723$p_value.o)
# Note: It seems that adding the log_p value reduces predictive accuracy

# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")

# Target 
target <- "replicated"

# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")

# Note: No missing values 
df <- altmext723[,voi]
```

## Training and testing split 

Let's start by randomly splitting the data in a training and testing set (70/30).

```{r}
# Number of training samples
n_train <- round( nrow(df) * 0.7)

# Training index
set.seed(112)
trainidx <- sample( seq_len(n_train), n_train )

# Training and testing set 
train <- df[trainidx, ]
test  <- df[-trainidx, ] 

# Feature matrices
X_train <- train[,features] 
y_train <- train[,target]

# Target vectors 
X_test <- test[,features] 
y_test <- test[,target]
```

## Predictions

First we fit the model to the data (training) and then test its predictive
capabilities on unseen data (testing). Since we will use the data to find
the best hyperparameters this is better thought of as validation set.

TODO: We need a true validation set! E.g. the Many Labs 2 data

### Training 

Define the function 

```{r}
# X11()
train_adaboost <- function(X, y, eta, n_trees, tree_hyperpar) {
  # Extract the number of rows in training set
  n_samples <- nrow(X)
  
  # Initialize the weights with equal weighting for all observations 
  D <- rep(1, n_samples)/n_samples
  
  # Pre-allocate the empty ensemble (this will include the trained learners)
  H <- vector("list", n_trees) 
  
  # Pre-allocate the data set (only for convenience)
  df <- data.frame(X, y)

  # Sequentially build the homogeneous ensemble 
  for (tree in 1:n_trees) {
    # Train a single weak learner (h_t) using weights (D) for the observations.
    # These  normalized(!) weights are sequentially optimized across iterations/trees.
    h <- rpart::rpart(formula = y ~ ., data = df , method = "class", 
                      weights = D, control = tree_hyperpar)
    
    # Get the predictions of the weak learner (h_t) for the training set
    y_pred <- predict(object = h, newdata = X, type = "class")
    
    # Compute the weighted accuracy of the weak learner (h_t) 
    wacc <- sum(D * (y == y_pred)) / sum(D) 
    
    # Compute the training error (e_t), which is one minus the weighted accuracy
    e <- 1 - wacc 
    
    # Compute the the weight/contribution (a_t) of the weak learner (h_t) 
    # In doing so, the weight is reduced by a factor eta (learning rate)
    a <- 0.5 * log( (1 - e)  / e) * eta
    
    # Pre-Prunning or Early Stopping:
    # Do not forget to add the parameter (prunning threshold)
    # if (a < prunning_threshold) {
    #  message("Prunning threshold reached")
    #  break
    # }
    
    # Update the weights for the next weak learner (h_[t+1])
    # --> Increase the weights of misclassified observations: (y == y_pred)
    # --> Decrease the weights of correctly classified observations: (y!=y_pred)
    m <- (y == y_pred) * 1 + (y != y_pred) * -1 
    
    # Update the weights of the observations for the next lerner (h_[t+1])
    D <- D * exp(-a * m)
    
    # Normalize the weights so they have a sum of one (probabilistic interpretation) 
    D <- D/sum(D)

    # Save the current weak learner (h_t) and its weight  
    H[[tree]] <- list("learner" = h, "learner_weight" = a, "observation_weight" = D)
    
    # H[[t]] <- list("weight" = a, "weaklearner" = h, "sampleweight" = D)
    # Store sample weights for visualization
    # cat(t,"weighted accuracy:", wacc, "\n")
    # Plot 
    # rattle::fancyRpartPlot(h, caption = NULL)
    # Sys.sleep(2)
  } 
  # Return the ensemble of trained weak learners
  H
}
```

Tree hyperparameter settings

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 100, # Number of learner
                           eta = 1e-1, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 

# illustration of learning the weights across iterations:
a <- sapply(ensemble, "[[", "learner_weight")
D <- t(sapply(ensemble, "[[", "observation_weight"))
plot(a, type = "l", ylim = c(0, .04), ylab = "Weights", xlab = "Trees", col = "red")
for(i in 1:ncol(D)) 
  lines(1:nrow(D), D[,i] + runif(1, -.00014, +.00014), col = adjustcolor(1, alpha = 0.1))

# DANIEL: why do the weights of the trees have these regular peaks for the sequential updates?
```

### Prediction

```{r}
test_adaboost <- function(X, ensemble) {
  pred <- rep(0, nrow(X))
  
  # Iterate across ensemble of trained weak learners
  for (tree in ensemble) {
    h <- tree$learner
    a <- tree$learner_weight
    
    # Get predictions for new data set (X) as integer 
    y_pred <- predict(h, X, type = "vector")
    
    # Transform from 1/2 coding to -1/1
    y_num <- 2 * y_pred - 3
    
    # DANIEL: Shouldn't the next parameter be called "a" ? And why -1/+1 coding and not 0/1 coding?!
    
    # Eta: learning rate (relative contribution/weight of each weak learner)
    pred <- pred + a * y_num
    
    # pred <- pred + a * eta * (2*as.numeric(y_class)-3)
    # Outcome: y
  }
  list("class" = sign(pred), "raw" = pred)
}
```

Make predictions with the trained model

```{r}
test$ypred <- test_adaboost(X_test, ensemble)$class

```

## Evaluation

```{r}
ROC <- AUC::roc(test$ypred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```

# Cross Validation

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 1e3,  eta = 1e-1, hyptree)
    
    # Make predictions for the validation set 
    ypred <- test_adaboost(test[,features], ensemble)$class
    
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(ypred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

This will run for approximately 4 minutes.

## Performance estimates

```{r}
auc
```

# Hyperparameter Optimization (TODO)

```{r}
# TODO: Increase n_trees 1e2:1e5 and reduce eta 1e-2 : 1e-5
hgrid <- expand.grid(
  
  # Tree-specific 
  maxdepth = 1:3, # Interaction depth
  
  # Ensemble specific
  n_tree = c(10, 100, 1000), # Number of trees
  eta = c(1, .1, .001, .0001), # Learning rate
  
  # Container for fold performance
  ee = NA, ml1 = NA, ml3 = NA, rpp = NA,  ssrp = NA, 
  auc.mean = NA, runtime = NA
) # Average performance

# Number of combinations 
hgrid_rows <- nrow(hgrid)
```

Hyperparameter optimization using Leave-Project-Out Cross-validation.

```{r}
# Partition the data by "Project ID"
df_split <- split(df[,voi], df$pid) 

# Setup a progress bar
pb <- txtProgressBar(0, hgrid_rows, initial = 0, style = 3) 

# For each entry in the hypergrid, do...
for(i in seq_len(hgrid_rows)) {
  
  # Update the progress bar
  setTxtProgressBar(pb, i) 
  
  # Track the iteration time (time-accuracy trade-off)
  runtime <- system.time({
    
    # Leave-One-Project-Out Cross-Validation
    for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
      
      # Define the testing set
      test <- df_split[[project]] # TODO: Inconsistent (partial matching vs explicit)
      
      # Define the training set
      train <- df[!(df$pid %in% project), ]
      
      # Fit the model to the data to get the ensemble
      ensemble <- train_adaboost(train[,features], train[,target],
                                 n_trees = hgrid$n_tree[i],  
                                 eta = hgrid$eta[i], 
                                 tree_hyperpar = rpart::rpart.control(
                                   maxdepth = hgrid$maxdepth[i]))
      
      # Make predictions on validation set 
      ypred <- test_adaboost(test[,features], ensemble)$class
      
      # Store the performance evaluation (level: effects)
      hgrid[[project]][i] <- AUC::auc( AUC::roc(ypred, test$replicated)) }
    
    # Calculate the average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i,4:8]))
  }) 
  # TODO: Inconsistent (partial matching vs explicit)
  hgrid$runtime[i] <- runtime[["elapsed"]]
}
# Close connection 
close(pb)
```

Performance evaluation sorted by highest AUC.

```{r}
# Note: EE is highly anomalous!
hgrid[order(-hgrid$auc.mean), ] 
```

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1.

# Optimized Ensemble 

If we do not stack the results, we should at least fit the model with the
hyperparameters that had the best average performance (here AUC) across all
replication projects. However, we don't get a true sense of the model's
out-of-sample capabilities because we don't have another project it hasn't seen
yet.

```{r}
# Define the testing set
test <- df_split[[project]] # TODO: Maybe inconsistent

# Define the training set
train <- df[!(df$pid %in% project), ]

# Fit the model to the data to get the ensemble
ensemble <- train_adaboost(train[,features], train[,target], 
                           n_trees = 1e4,  eta = 1e-3, 
                           tree_hyperpar = rpart::rpart.control(
                                 maxdepth = 3))

# Make predictions on validation set 
ypred <- test_adaboost(test[,features], ensemble)$class
```
