---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

- Why Boosting?

- Why AdaBoost?

- What is Boosting 

- What means adaptive? (see, e.g. Kunapuli, 4.2.1)

- Comparison to prediction markets: lay person (boosting) vs experts (random forest)


# Approach (hands dirty)

- Describe the approach taken
  - Why implementing it yourself? 
    - Understanding
    - The role of hyperparameters!
    - Hyperparameters and tuning! 

- Note Florian Pargent et al.'s paper (supplement not complement)
  - For basic terminology (e.g., bias-variance trade-off, classification, AUC)
  - For the right mindset (e.g., predictions in psychology)
  - and other things

# Software (maybe together with Approach)

- Describe the software used

```{r setup}
library(adatutoR)
```

- Describe the data set, i.e. data sets used


```{r}
# Load the data
data("altmext723")
```

- TODO: Document the data set in R so that it is accessible with `?altmext723`



# Module 1: Splitting 


Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")

# Outcome (classification task) 
target <- "replicated"

# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")

# Note: No missing values 
df <- altmext723[,voi]
```

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

### Simple random sampling 

```{r}
# Reproducibility
set.seed(112)

# Split proportion
sprop <- 0.7

# Number of effects 
n_effect <- nrow(df)

# Number of training samples
n_train <- round(n_effect * sprop)

# Training index
train_idx <- sample(seq_len(n_effect), n_train)

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

### Stratified sampling & imbalances

However, there are imbalances in the data set. 

- What is and why use stratified sampling - with the outcome as stratum? 
"Recall that stratified means the folds are created in such a way that the class
distribution is preserved across the folds. This also helps with imbalanced data
sets, as stratification ensures that data from all classes is represented." (p. 103)


```{r}
table(df$replicated)
```

Therefore we should do a stratified split. This means we need to partition the
data into groups based on a stratum variable - here: replicated, and then draw
a random sample from each stratum.

```{r}
# Reproducibility
set.seed(112)

# Partition the data based on the outcome variable
df_split <- split(df, df$replicated)

# Number of training samples (stratified)
n_train <- lapply(df_split, function(df) round(nrow(df) * sprop))

# Training index (stratified)
strat_idx <- mapply(function(df, n) sample(rownames(df), n), df_split, n_train)

# Training index 
train_idx <- as.numeric(unlist(strat_idx))

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

Now the training and testing set have approximately the same proportion of
replication and failure as the original outcome variable. We could also use the
`rsample` package for that.

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```

# Module 2: AdaBoost

First we fit the model to the data (training) and then test its predictive
capabilities on unseen data (testing). Since we will use the data to find
the best hyperparameters this is better thought of as validation set.

- Discuss a couple of iterations of boosting (Kunapali, p.90) 

- (Extra: Decision stumps)

## An iterative process...

Setup.

```{r}
n_trees <- 100
# Extract the number of rows
n_samples <- nrow(X_train)
# Initialize the weights
D <- rep(1, n_samples)/n_samples
# Pre-allocate the empty ensemble
H <- vector("list", n_trees)
df <- data.frame(X_train, y_train)
```

### Digression: Classification stumps and shallow trees 

Crash course on classification trees to understand stumps and shallow trees
using `rpart`.

- What are decision trees?
- What are decision stumps?
- Splitting criterion (CART, Breiman)?

A selection of most important tree hyperparameters for our AdaBoost
implementation.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, # Boost stumps
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
```

Fit a decision stump to the training data and visualize the result.

```{r}
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump", extra = 102, digits = 5)
```


- Why this particular split?
  - Gini impurity
  - Sensitivity to observation weights (Assumption: equal) 

#### Gini impurity

- Interesting, especially for continuous features 
- Calculation example(?)

The split point is chosen by sorting the individual values of a given feature
and then calculating the Gini impurity for each mid-point between adjacent values.

For example, lets say we have

```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
# Example above:
constant <- adjacent_means[69] # 0.9286641
N_total <- length(y_train) # N_total: 107
# Left leaf (power.o < 0.9286641: yes)
criterion_left <- df$power.o < constant
N_left <- length(df$y_train[criterion_left]) # 69 total
left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1) # 19 successes
left_failure <- N_left - left_success # 50 failures
# Right leaf (power.o < 0.9286641: no)
criterion_right <- df$power.o >= constant
N_right <- length(df$y_train[criterion_right]) # 38 total
right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) # 29 sucesses
right_failure <- N_right - right_success # failures
# Gini index
gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
# Gini impurity
gini_impurity <- (N_left / N_total) * gini_d1 + (N_right / N_total) * gini_d2
round(gini_impurity, digits = 5)
```

# Control calculations 

```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
gini_impurity <- vector(length = length(adjacent_means))
  for(i in 1:length(adjacent_means) ) {
  constant <- adjacent_means[i] 
  N_total <- length(y_train) 
  # Left leaf 
  criterion_left <- df$power.o < constant
  N_left <- length(df$y_train[criterion_left])
  left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1)
  left_failure <- N_left - left_success 
  # Right leaf
  criterion_right <- df$power.o >= constant
  N_right <- length(df$y_train[criterion_right]) 
  right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) 
  right_failure <- N_right - right_success 
  # Gini index
  gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
  gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
  # Impurity
  gini_impurity[i] <- (N_left / N_total) * gini_d1 + (N_right / N_total) * gini_d2
  }
# Position of smallest impurity 
which.min(gini_impurity)
# Outcome 
adjacent_means[which.min(gini_impurity)]
```


#### Sensitivity to observation weights

Normalized initialization weights. Should not change anything to before.

```{r}
# Equal weight to each observation
D <- rep(1, n_samples)/n_samples
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

Normalized random observation weights for demonstration. Changes result.

```{r}
# Reproducibility
set.seed(113)
# Random draws from Uniform(0,1)
D_unorm <- runif(107) 
# Normalize (sum: 1)
D <- D_unorm/sum(D_unorm)
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

##### Additional exercises

`rpart` offers a number of hyperparameters that affect the outcome of the
results. We leave it to the reader as an additional exercise to explore how each
of them influences the result. For our purposes, we will focus only on the
maximum depth and the splitting criterion.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree (root node: depth = 0)
  # Important: We usually boost stumps (depth = 1) or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini" # or ID3: split = "information"
  )
```

##### Additional exercise 1

We created a decision stump using the gini impurity as splitting criterion.
Adapt the code above to fit a shallow tree of depth two and three.

##### Additional exercise 2

Despite the hyperparameters we already mentioned there are more. See the
`?rpart::rpart.control` documentation and explore how changing different
arguments influence the results.


### Adaboost continued

Normalized weights.

```{r}
# Initialize the weights (normalized)
D <- rep(1, n_samples)/n_samples
```

Adaptive boosting.

```{r}
# for (tree in 1:n_trees) {
  # Train a single weak learner (h_t) with normalized(!) sample weights (D)
  h <- rpart::rpart(formula = y ~ ., data = df , method = "class",
                    weights = D, control = tree_hyperpar)
  # Make predictions with the weak learner (h_t)
  y_pred <- predict(object = h, newdata = X, type = "class")
  # y_pred <- 2*as.numeric(y_class) - 3
  wacc <- sum(D * (y == y_pred)) / sum(D)
  # Compute the training error (e_t) of the weak learner (h_t)
  # This is simply one minus the weighted accuracy, given weights D
  e <- 1 - wacc
  # Equivalent but less readable: e <- sum(D * (y != y_pred)) / sum(D)
  # Compute the the weight (a_t) of the weak learner (h_t)
  # Reduce the learner weight by a factor eta (learning rate)
  a <- 0.5 * log( (1 - e)  / e) * eta
  # Pre-Prunning or Early Stopping:
  # Do not forget to add the parameter (prunning threshold)
  # if (a < prunning_threshold) {
  #  message("Prunning threshold reached")
  #  break
  # }
  # Updates the weights
  # Increase for misclassified examples: (y == y_pred)
  # decrease for correctly classified examples: (y!=y_pred)
  m <- (y == y_pred) * 1 + (y != y_pred) * -1
  # Update the sample weights and limit the contribution of h_t by eta
  D <- D * exp(-a * m)
  # Normalize weights (probabilistic interpretation)
  D <- D/sum(D)
  # Save the weak learner and its weight
  H[[tree]] <- list("learner" = h, "learner_weight" = a,
                    "observation_weight" = D)
  # Store sample weights for visualization
}
  # Return the ensemble
# H
}
```

```{r}
# Plot
# rattle::fancyRpartPlot(h, caption = NULL)
# Sys.sleep(2)
```


## Hands dirty: Implementation

### Training 

Look under the hood of the function 

```{r}
train_adaboost
```

Tree hyperparameter settings

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 
```


### Prediction

```{r}
test_adaboost
```

Make predictions with the trained model and add them to the testing set.

```{r}
y_class <- test_adaboost(X_test, ensemble)$class
test$y_pred <- factor( (y_class + 3) / 2, labels = c("no", "yes") )
```

## Evaluation

```{r}
ROC <- AUC::roc(test$y_pred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```


# Module 3: Leave-Project-Out Cross-Validation

TODO: Check if code in Module 2 is running

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 5,  eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

This will run for approximately 4 minutes.

TODO: This seems to run, but why do I always get the same performance estimates?
No matter if I use 1 tree or 1e4 trees...

```{r}
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 500, eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

## Performance estimates

```{r}
auc
```

# Module 4: Hyperparameter Optimization using LPO-CV

TODO: Check if code in Module 3 is running

```{r}
# TODO: Increase n_trees 1e2:1e5 and reduce eta 1e-2 : 1e-5
hgrid <- expand.grid(
  # Tree-specific 
  maxdepth = 1:3, # Interaction depth
  # Ensemble specific
  n_tree = c(1e1, 1e2), # Number of trees
  eta = c(1, 0.1), # Learning rate
  # Container for fold performance
  ee = NA, ml1 = NA, ml3 = NA, rpp = NA,  ssrp = NA, 
  auc.mean = NA, runtime = NA) # Average performance
# Number of combinations 
hgrid_rows <- nrow(hgrid)
```

Hyperparameter optimization using Leave-Project-Out Cross-validation.

```{r}
# Partition the data by "Project ID"
df_split <- split(df[,voi], df$pid) 
# Setup a progress bar
pb <- txtProgressBar(0, hgrid_rows, initial = 0, style = 3) 
# For each entry in the hypergrid, do...
for(i in seq_len(hgrid_rows)) {
  # Progress bar
  setTxtProgressBar(pb, i) 
  # Track the iteration time (time-accuracy trade-off)
  runtime <- system.time({
    # Leave-One-Project-Out Cross-Validation
    for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
      # Define the testing set
      test <- df_split[[project]] # TODO: Inconsistent (partial matching vs explicit)
      # Define the training set
      train <- df[!(df$pid %in% project), ]
      # Fit the model to the data to get the ensemble
      ensemble <- train_adaboost(train[,features], train[,target],
                                 n_trees = hgrid$n_tree[i],  
                                 eta = hgrid$eta[i], 
                                 tree_hyperpar = rpart::rpart.control(
                                   maxdepth = hgrid$maxdepth[i]))
      # Make predictions on validation set 
      y_pred <- test_adaboost(test[,features], ensemble)$class
      # Store the performance evaluation (level: effects)
      hgrid[[project]][i] <- AUC::auc( AUC::roc(y_pred, test$replicated)) }
    # Calculate the average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i,4:8]))
  }) 
  # TODO: Inconsistent (partial matching vs explicit)
  hgrid$runtime[i] <- runtime[["elapsed"]]
}
# Close connection 
close(pb)
```

Performance evaluation sorted by highest AUC.

```{r}
# Note: EE is highly anomalous!
hgrid[order(-hgrid$auc.mean), ] 
```

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1.

## Optimized Ensemble 

If we do not stack the results, we should at least fit the model with the
hyperparameters that had the best average performance (here AUC) across all
replication projects. However, we don't get a true sense of the model's
out-of-sample capabilities because we don't have another project it hasn't seen
yet.

```{r}
# Define the testing set
test <- df_split[[project]] # TODO: Maybe inconsistent
# Define the training set
train <- df[!(df$pid %in% project), ]
# Fit the model to the data to get the ensemble
ensemble <- train_adaboost(train[,features], train[,target], 
                           n_trees = 1e4,  eta = 1e-3, 
                           tree_hyperpar = rpart::rpart.control(
                                 maxdepth = 3))
# Make predictions on validation set 
y_pred <- test_adaboost(test[,features], ensemble)$class
```

# Module 5: Stacking (?) 

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1. 
