---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

- Why Boosting?

- Why AdaBoost?

- What is Boosting 

- What means adaptive? (see, e.g. Kunapuli, 4.2.1)

- Comparison to prediction markets: lay person (boosting) vs experts (random forest)


# Approach (hands dirty)

- Describe the approach taken
  - Why implementing it yourself? 
    - Understanding
    - The role of hyperparameters!
    - Hyperparameters and tuning! 

- Note Florian Pargent et al.'s paper (supplement not complement)
  - For basic terminology (e.g., bias-variance trade-off, classification, AUC)
  - For the right mindset (e.g., predictions in psychology)
  - and other things

# Software (maybe together with Approach)

- Describe the software used

```{r setup}
library(adatutoR)
```

- Describe the data set, i.e. data sets used


```{r}
# Load the data
data("altmext723")
```

- TODO: Document the data set in R so that it is accessible with `?altmext723`



# Module 1: Splitting 


Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")

# Outcome (classification task) 
target <- "replicated"

# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")

# Note: No missing values 
df <- altmext723[,voi]
```

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

### Simple random sampling 

```{r}
# Reproducibility
set.seed(112)

# Split proportion
sprop <- 0.7

# Number of effects 
n_effect <- nrow(df)

# Number of training samples
n_train <- round(n_effect * sprop)

# Training index
train_idx <- sample(seq_len(n_effect), n_train)

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

### Stratified sampling & imbalances

However, there are imbalances in the data set. 

- What is and why use stratified sampling - with the outcome as stratum? 
"Recall that stratified means the folds are created in such a way that the class
distribution is preserved across the folds. This also helps with imbalanced data
sets, as stratification ensures that data from all classes is represented." (p. 103)


```{r}
table(df$replicated)
```

Therefore we should do a stratified split. This means we need to partition the
data into groups based on a stratum variable - here: replicated, and then draw
a random sample from each stratum.

```{r}
# Reproducibility
set.seed(112)

# Partition the data based on the outcome variable
df_split <- split(df, df$replicated)

# Number of training samples (stratified)
n_train <- lapply(df_split, function(df) round(nrow(df) * sprop))

# Training index (stratified)
strat_idx <- mapply(function(df, n) sample(rownames(df), n), df_split, n_train)

# Training index 
train_idx <- as.numeric(unlist(strat_idx))

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

Now the training and testing set have approximately the same proportion of
replication and failure as the original outcome variable. We could also use the
`rsample` package for that.

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```

# Module 2: A primer on classification stumps 

In Module 3, we will show how the AdaBoost algorithm transforms classification
stumps or shallow classification trees into a strong ensemble machine learning
model. Thus, a prior study of classification trees in Module 2 seems mandatory.
However, Module 2 is not a thorough introduction to decision trees (see, e.g.,
Breiman et al. for a complete introduction). The focus will be on the details
necessary to understand AdaBoost. We will use the `rpart` package (Ref.) for
hands-on understanding and `rpart.plot` (Ref.) for visualizing the results.

##### Classification problems

The `altmext723` data set contains measurements on the power of the original
study, its effect size, there is information about the number of observations
available, and so on. Typing `?altmext723`, we can access the full variable list
as well as additional information.

With these measurements, a general goal of predictive modeling is to extract a
pattern between features (e.g., power of the original study) and the outcome
(e.g., replication success) that allows to accurately predict whether future
replication attempts will be successful (class 1) or fail (class 2). Given the
number of classes of the target variable, `replicated` the present scenario can
be called a two-class *classification problem*.

In this tutorial, however, we will not wait for future data to evaluate
performance. This is why, in Module 1, we already set aside a subset of the
data, the test set, to evaluate the predictive capabilities of the later-built
machine learner. The final evaluation is part of Module xxx later in this
tutorial (see Module xxx and). In the next section, we will first focus on how
classification trees address the classification problem outlined above.

Note: Actually, this will be a validation set, since we tune the hyperparameter
with these stuff. Write this problem somewhere, that the data set is to
small to always set aside some portion of the data. A solution will be discusses
with cross-validation, although this also provides just an estimate that differs
from the true generalization performance.

### Classification trees

Binary *classification trees* learn by repeatedly asking (yes/no) questions
about features of the data that allow them to organize the data points into
non-overlapping groups better *partitions*. An example for a first question
could be: Is the power of the study less than 0.93? If it is, the study is
placed in the left group; if it is not, it is placed in the right group.
Following the given data *splitting rule* ultimately results in two
non-overlapping groups, one containing studies with `power.o > 0.9` and another
with `power.o <= 0.9` (see Figure 1, left).

With a second question, targeting all the high power studies (`power.o >= 0.9`),
for instance, the tree could further partition the data (now at depth two): Is
the number of observations also greater than 100 (`n.o >= 100`)?" In this way
the trees go on continue to grow; that is, the grouping process continues, until
the data cannot be divided any further or a termination criterion is reached.
Termination criteria can be, e.g., a minimum number of 100 studies in the final
group, or a minimum number of 100 studies to consider a split (see
`?rpart.control` for available options in `rpart`).

```{r}
# Extract only two features
# Note: Reduces run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <-  train[,union(subfeat, target)]

# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"

# Create a decision tree 
h <- rpart::rpart(formula = replicated ~ ., data = subdf,  method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Classification Stump", extra = 102, 
                       digits = 5, box.palette="BuGy")

# Number of classes
k <- length(unique(subdf$replicated))

# Range(s) as list
rng <- lapply(subdf[,subfeat], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Stump")
points(grid, col = grid_colors, pch = ".")
points(subdf[,subfeat], col = symbol_colors, pch = as.integer(y_train) + 15)
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)
```

To compare the results, to a GLM see the Appendix!

```{r}
# Maybe as comparison the decision boundaries from a GLM?
mdl <- glm(replicated ~., data = subdf, family=binomial(link='logit'))
# Change class attribute
class(mdl) <- c("lr", class(mdl))
# Predicted probability of replication success
y_prob <- predict(mdl, grid, type = "response")
# If p > 0.5 "replicated", else "not replicated"
y_pred <- factor(as.integer(y_prob > 0.5))
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "no" means "not replicated", "yes" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == 0, col1_transp, col2_transp) 
# Plot the features
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
     xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Bernoulli Logit Generalized Linear Model")
points(grid, col = grid_colors, pch = ".")
points(subdf[,subfeat], col = symbol_colors, pch = as.integer(y_train) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
        col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
       col = c(col2, col1), cex=0.7)
```

#### Classification stumps

Binary *classification stumps* shorten the repetitive process of classification
trees by limiting themselves to a single (yes/no) question (see Figure 1, left).
Consequently, they partition the data into two final groups, terminating at a
maximum depth of one (see Figure 1, left side & center, and Code xxx). Calling
the starting point in the graph of the tree its root or *root node* and the
resulting group at the terminal point its *leave node*, a classification stump
can be categorized as a tree with one root and two leave nodes (see, e.g. ref.).

The final decision on whether studies with a certain feature, such as high
power, are considered successful or not is made by the classification stump in a
majority vote based on the frequency of labeled studies in the leave node. For
example, in the classification stump in Figure 1, there are more failed
replication attempts for studies in the left branch, so the overall
classification for studies with the feature `power.o < 0.93` is "failure". In
the right branch, there are more successes, so studies with `power.o >= 0.93`
are considered successful. To verify this mechanically, we can use the code in
Code Chunk 1.

TODO

```{r Code 1}
split_criterion <- train$power.o < 0.92866
# Studies with df$power.o < 0.92866 
table(train$replicated[split_criterion])
# Studies with df$power.o >= 0.92866 
table(train$replicated[!split_criterion])
```

### Fit and visuaize a classification stump

There are several ways to fit and visualize classification trees. In the 
tutorial, we will use the `rpart` package for fitting and the `rpart.plot`
package for visualizing the trees. Specifically, we will use the functions
`rpart()`, `rpart.control()` and `rpart.plot()`. Each function is outlined in
turn. For more information, see the R documentation, e.g. 
`help("rpart", package = "rpart")` or `?rpart::rpart`.

```{r Code 2}
# Hyperparameter setup for a classification stump
hyptree <- rpart::rpart.control(
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  split = "gini") 

# Fit a classification stump
cstump <- rpart::rpart(formula = replicated ~ ., 
                       data = train[,union(features, target)],
                       method = "class", control = hyptree)
```

In Code Chunk 2, we first focus on `rpart()`. "rpart" is short for "recursive
partitioning". "Recursive partitioning refers to the process of repeatedly
putting the data into non-overlapping groups, which we mentioned above. The
function takes several arguments (see `?rpart::rpart`), but we will focus mainly
on `formula`, `data` and `control`. The `method` argument with the value `class`
just makes explicit that we deal with a classification problem.

TODO

First, we pass a formula to `rpart()` that has the general form: `target ~
features`. One could read this as "the outcome or target (replicated) is a
function of the features or predictors. All variables, the features and the
target, must be part of the data frame that we pass as value to `rpart` with the
`data` argument. The dot `.` in the formula, tells `rpart()` to consider all
variables in the data set, despite the one on the left hand side, `replicated`.

The `rpart.control()` function allows to specify the tree hyperparamters. In
general, *hyperparameter* are external parameter that we as the user set
*before* fitting the model and which influence the model's complexity or it's
learning process. Psychologists usually learn about hyperparameters in the
context of (non-Bayesian, exploratory) factor analysis, for example, where they
have to decide on a number of factors to retain before being able to evaluate a
particular solution. We characterized stumps as primitive classification trees
with maximum depth equal to one. Consequently, we should limit the complexity of
the model using `maxdepth = 1` to avoid the tree growing beyond a stump. Put
another way, this constraint limits the number of data partitions: There is one
group with studies `power.o < 0.93` and one with studies `power.o >= 0.93`. 

```{r}
# Visualize a classification stump
rpart.plot::rpart.plot(cstump, main = "Classification Stump", extra = 102,
                       digits = 5)
```

We demonstrate how to visualize a fitted classification stump using the The
`rpart.plot()` function from `rpart.plot` in Code Chunk 2. The code can also be
used to reproduce Figure 2. The `rpart.plot()` function takes an `rpart` model
object, such as the fitted classification stump `cstump`, and creates a
flow-char like(?) plot (more precise) of it. There are several arguments to
customize the appearance of the plot (for a full list, see
`?rpart.plot::rpart.plot`). If a more fine-grained customization of the
appearance is required, we recommend switching to the `rpart.plot`'s `prp()`
function.

If working with (tree) hyperparameters still seems unfamiliar or if the goal is
to solidify and extend understanding of the discussed concepts, we recommend
doing Additional Exercise 1, visualizing the results using Code Chunk 2.

#### Additional exercise 1

TODO

Access the `rpart.control()` documentation with `?rpart.control` and get an
overview over all hyperparameters of the `rpart` package. Increase the maximum
depth of the stump to two and three to create shallow classification trees.
Then, set the minimum number of observations in the terminal node to 1. How will
these setting influence the shape of the tree. Think about the solution first,
before running the code. (Are the results sensical give your general knowledge
about meta-science

### Gini impurity

Binary *classification stumps* learn by asking a single (yes/no) questions about
some feature of the data, that allows them to organize the data into two
partitions. This characterization, however, lacks a clear answer on how a data
feature and the split in particular is chosen: Why `power.o > 0.98266`?

There are multiple strategies to an optimal splitting rule (see, e.g.). The
strategy in Breimans classic CART algorithm which is part of AdaBoost by Freud
and Shapire which we'll discuss later is to split the data in such a way that
they minimize the risk of misclassification ().

"The best splits would then be those that most reduced the estimated
misclassification rate (p.94)"

As judged by *Gini impurity*. Cite!
Skipping a ton of techical details, one option to do that is the Gini impurity.

"Gini index of diversity as a measure of node impurity; i.e (p.38)"

The Gini impurity considers how often we would misclassify a randomly chosen
example from our data set if we were to guess its class label based on the class
distribution in that group. (ref.).

Check: Applied to our example, this means that the Gini impurity measures how
often we are expexted to misclassify a replication study's outcome (success or
failure) if we were to randomly assign a label (success or failure) based on the
distribution of successes and failures within our group of replication studies
-- split according to a specific power threshold (e.g. power.o >= 0.93).

Say: Gini impurity determines how our model learns (see above).

So finding the feature or split with the lowest Gini impurity helps us
minimizing this misclassification risk.

#### Why `power.o > 0.93`?

When you're building a decision tree, you calculate the Gini impurity for each
possible split (question) you could make on the data. The question that leads to
the lowest Gini impurity (the purest resulting groups) is the one you choose.

There are only continuous features in the data set. The power of the original
study is one of them. For continuous features, the split point is chosen by
sorting the individual values and then calculating the Gini impurity for each
mid-point between adjacent values.

```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
```
We want to reproduce the results in the graphic. Therefore we will calculate
the Gini impurity for a split with the criterion `power.o < 0.9288`.


$$I_G(D) = 1 - \sum_{j=1}^K p^2_i$$

Since we have data with examples from 2 classes the formula simplifies as

$$I_G(D) = 1 - \sum_{j=1}^2 p^2_i =  1 - (p_1^2 + p_2^2)$$

where we calculate as $p_i = \frac{f_i}{n_i}$.

Weighted Gini impurity given feature X for i = 1,2 partitions.

$$I_{G\mid X}(D) = \sum_{i=1}^2 \frac{n_i}{n} I_G(D_i) = \frac{n_1}{n} I_G(D_1) + \frac{n_2}{n} I_G(D_2)$$

```{r}
# Example above:
constant <- adjacent_means[69] # 0.9286641
N_total <- length(y_train) # N_total: 107

# Left leaf (power.o < 0.9286641: yes)
criterion_left <- df$power.o < constant
N_left <- length(df$y_train[criterion_left]) # 69 total
left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1) # 19 successes
left_failure <- N_left - left_success # 50 failures

# Right leaf (power.o < 0.9286641: no)
criterion_right <- df$power.o >= constant
N_right <- length(df$y_train[criterion_right]) # 38 total
right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) # 29 successes
right_failure <- N_right - right_success # failures
```


```{r}
# Individual Gini impurities
gini_left <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
gini_right <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
```

Gini left: The Gini impurity suggests that if we were to randomly pick a
replication from our sample where the original had a power of less than 0.93 and
say that it would fail to replicate, there is approximately a 39.91% chance that
we would be wrong.

Gini right: The Gini impurity suggests that if we were to randomly pick a
replication from our sample where the original had a power of greater than 0.93 and
say that it would successfully replicate, there is approximately a 36.15% chance
that we would be wrong.

We finally weight the results to get the overall weighted Gini impurity
considering the different class distributions.

```{r}
# Overall weighted Gini impurity 
gini_impurity <- (N_left / N_total) * gini_left + (N_right / N_total) * gini_right
round(gini_impurity, digits = 5)
```


G = 0 indicates a perfectly pure node (all elements belong to the same class).
G = 0.5 indicates maximum impurity (elements are evenly distributed among all classes).

Estimating the Gini impurity

Gini impurity = 1 - (probability of the most frequent class)^2 - (probability of
the second-most frequent class)^2 - ... (and so on for all classes)


Gini Impurity in Action

Group 1 (e.g., power > 0.8): What's the Gini impurity of this group in terms of
replication success vs. failure labels? Are the studies in this group mostly
successful replications, mostly failures, or a mix?

Group 2 (e.g., power ≤ 0.8): Calculate the Gini impurity for this group as well.
        
Choosing the best Split: The algorithm selects the power threshold split that
results in the lowest overall Gini impurity in the new groups. This split
maximizes class purity, meaning it clearly separates studies more likely to
succeed in replication from those more likely to fail.






For example, lets say we have



1. step gini_d1 and gini_d2: You're correctly calculating the standard Gini
impurity for the left and right groups separately using the formula: 1 -
(probability of class 1)^2 - (probability of class 2)^2 - ...

2. step (N_left / N_total) and (N_right / N_total):  These represent the weights
assigned to each group's Gini impurity, based on the proportion of examples in
each group relative to the total.

3. step: gini_impurity:  The final line combines the weighted Gini impurities of both
groups to calculate the overall weighted Gini impurity for the potential split.

TODO: Calculations seem to be different from, e.g.: https://machinelearningmastery.com/classification-and-regression-trees-for-machine-learning/

Additional exercise: We leave it as an additional exercise to the reader to
verify that the result has the lowest Gini impurity among all splits (a) within
the particular features and (b) overall.


Important: For a two class classification problem.

```{r}
# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01)  # Values for fraction of class k
y <- 1 - (x^2) - (1-x)^2      # Gini Impurity calculation

# Create the plot
plot(x, y, type = "l",  # Specify line plot
     xlab = "Fraction of Class k (pk)", 
     ylab = "Impurity",
     main = "Gini Impurity")

mtext("2 class classification problem")
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1)) 

# Optional: Add gridlines
abline(v = seq(0, 1, by = 0.1), h = seq(0, 1, by = 0.1), col = "lightgray") 
 
```



# Control calculations 






```{r}
subdf <- df[, c("power.o", "n.o", "y_train")]
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 3, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
  # Create a decision tree 
h <- rpart::rpart(formula = y_train ~ ., data = subdf , method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Classification Tree", extra = 102, 
                       digits = 5, box.palette="BuGy")

# With minbucket = 4 i can reproduce the following values. 
# Check: Does this make sense? Maybe ignore the decision tree and just use
# decision stump. Or describe that the last split is nonsensical.

table(subdf$y_train[subdf$power.o < 0.92866 & subdf$n.o >= 24])
table(subdf$y_train[subdf$power.o < 0.92866 & subdf$n.o < 24])
table(subdf$y_train[subdf$power.o >= 0.92866 & subdf$power.o >= 1])
table(subdf$y_train[subdf$power.o >= 0.92866 & subdf$power.o < 1])
```

```{r}
# Number of classes
k <- length(unique(subdf$y_train))
# Range(s) as list
rng <- lapply(subdf[-length(subdf)], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Tree")
points(grid, col = grid_colors, pch = ".")
points(subdf[-length(subdf)], col = symbol_colors, pch = as.integer(y_train) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)

```

Decision stumps simplify this repetitive partitioning process by asking only a
single question about some feature of the data. For example, is the power of the
original study greater than . Classification stump is a classification tree of
depth 1.


One question that is left unanswered now, is how do we get to a particular
question that allows us to parition the data in the way we've seen?

- Splitting criterion 
- Focus on CART with Gini impurity

A selection of most important tree hyperparameters for our AdaBoost
implementation.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, # Boost stumps
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
```

Fit a decision stump to the training data and visualize the result.

```{r}
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump", extra = 102, digits = 5)
```

- What do we see?
  - Why `digits =5`?

- Why this particular split?
  - Gini impurity
  - Sensitivity to observation weights (Assumption: equal) 

Control calculations

```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
gini_impurity <- vector(length = length(adjacent_means))
  for(i in 1:length(adjacent_means) ) {
  constant <- adjacent_means[i] 
  N_total <- length(y_train) 
  # Left leaf 
  criterion_left <- df$power.o < constant
  N_left <- length(df$y_train[criterion_left])
  left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1)
  left_failure <- N_left - left_success 
  # Right leaf
  criterion_right <- df$power.o >= constant
  N_right <- length(df$y_train[criterion_right]) 
  right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) 
  right_failure <- N_right - right_success 
  # Gini index
  gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
  gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
  # Impurity
  gini_impurity[i] <- (N_left / N_total) * gini_d1 + (N_right / N_total) * gini_d2
  }
# Position of smallest impurity 
which.min(gini_impurity)
# Outcome 
adjacent_means[which.min(gini_impurity)]
```


#### Sensitivity to observation weights

Normalized initialization weights. Should not change anything to before.

```{r}
# Equal weight to each observation
D <- rep(1, n_samples)/n_samples
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

Normalized random observation weights for demonstration. Changes result.

```{r}
# Reproducibility
set.seed(113)
# Random draws from Uniform(0,1)
D_unorm <- runif(107) 
# Normalize (sum: 1)
D <- D_unorm/sum(D_unorm)
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

##### Additional exercises

`rpart` offers a number of hyperparameters that affect the outcome of the
results. We leave it to the reader as an additional exercise to explore how each
of them influences the result. For our purposes, we will focus only on the
maximum depth and the splitting criterion.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree (root node: depth = 0)
  # Important: We usually boost stumps (depth = 1) or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini" # or ID3: split = "information"
  )
```

##### Additional exercise 1

We created a decision stump using the gini impurity as splitting criterion.
Adapt the code above to fit a shallow tree of depth two and three.

##### Additional exercise 2

Despite the hyperparameters we already mentioned there are more. See the
`?rpart::rpart.control` documentation and explore how changing different
arguments influence the results.




# Module 3: AdaBoost

First we fit the model to the data (training) and then test its predictive
capabilities on unseen data (testing). Since we will use the data to find
the best hyperparameters this is better thought of as validation set.

- Discuss a couple of iterations of boosting (Kunapali, p.90) 

- (Extra: Decision stumps)

## An iterative process...

Setup.

```{r}
n_trees <- 100
# Extract the number of rows
n_samples <- nrow(X_train)
# Initialize the weights
D <- rep(1, n_samples)/n_samples
# Pre-allocate the empty ensemble
H <- vector("list", n_trees)
df <- data.frame(X_train, y_train)
```


### Adaboost continued

Normalized weights.

```{r}
# Initialize the weights (normalized)
D <- rep(1, n_samples)/n_samples
```

Adaptive boosting.

```{r}
# for (tree in 1:n_trees) {
  # Train a single weak learner (h_t) with normalized(!) sample weights (D)
  h <- rpart::rpart(formula = y ~ ., data = df , method = "class",
                    weights = D, control = tree_hyperpar)
  # Make predictions with the weak learner (h_t)
  y_pred <- predict(object = h, newdata = X, type = "class")
  # y_pred <- 2*as.numeric(y_class) - 3
  wacc <- sum(D * (y == y_pred)) / sum(D)
  # Compute the training error (e_t) of the weak learner (h_t)
  # This is simply one minus the weighted accuracy, given weights D
  e <- 1 - wacc
  # Equivalent but less readable: e <- sum(D * (y != y_pred)) / sum(D)
  # Compute the the weight (a_t) of the weak learner (h_t)
  # Reduce the learner weight by a factor eta (learning rate)
  a <- 0.5 * log( (1 - e)  / e) * eta
  # Pre-Prunning or Early Stopping:
  # Do not forget to add the parameter (prunning threshold)
  # if (a < prunning_threshold) {
  #  message("Prunning threshold reached")
  #  break
  # }
  # Updates the weights
  # Increase for misclassified examples: (y == y_pred)
  # decrease for correctly classified examples: (y!=y_pred)
  m <- (y == y_pred) * 1 + (y != y_pred) * -1
  # Update the sample weights and limit the contribution of h_t by eta
  D <- D * exp(-a * m)
  # Normalize weights (probabilistic interpretation)
  D <- D/sum(D)
  # Save the weak learner and its weight
  H[[tree]] <- list("learner" = h, "learner_weight" = a,
                    "observation_weight" = D)
  # Store sample weights for visualization
}
  # Return the ensemble
# H
}
```

```{r}
# Plot
# rattle::fancyRpartPlot(h, caption = NULL)
# Sys.sleep(2)
```


## Hands dirty: Implementation

### Training 

Look under the hood of the function 

```{r}
train_adaboost
```

Tree hyperparameter settings

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 
```


### Prediction

```{r}
test_adaboost
```

Make predictions with the trained model and add them to the testing set.

```{r}
y_class <- test_adaboost(X_test, ensemble)$class
test$y_pred <- factor( (y_class + 3) / 2, labels = c("no", "yes") )
```

## Evaluation

```{r}
ROC <- AUC::roc(test$y_pred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```


# Module 3: Leave-Project-Out Cross-Validation

TODO: Check if code in Module 2 is running

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 5,  eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

This will run for approximately 4 minutes.

TODO: This seems to run, but why do I always get the same performance estimates?
No matter if I use 1 tree or 1e4 trees...

```{r}
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 500, eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

## Performance estimates

```{r}
auc
```

# Module 4: Hyperparameter Optimization using LPO-CV

TODO: Check if code in Module 3 is running

```{r}
# TODO: Increase n_trees 1e2:1e5 and reduce eta 1e-2 : 1e-5
hgrid <- expand.grid(
  # Tree-specific 
  maxdepth = 1:3, # Interaction depth
  # Ensemble specific
  n_tree = c(1e1, 1e2), # Number of trees
  eta = c(1, 0.1), # Learning rate
  # Container for fold performance
  ee = NA, ml1 = NA, ml3 = NA, rpp = NA,  ssrp = NA, 
  auc.mean = NA, runtime = NA) # Average performance
# Number of combinations 
hgrid_rows <- nrow(hgrid)
```

Hyperparameter optimization using Leave-Project-Out Cross-validation.

```{r}
# Partition the data by "Project ID"
df_split <- split(df[,voi], df$pid) 
# Setup a progress bar
pb <- txtProgressBar(0, hgrid_rows, initial = 0, style = 3) 
# For each entry in the hypergrid, do...
for(i in seq_len(hgrid_rows)) {
  # Progress bar
  setTxtProgressBar(pb, i) 
  # Track the iteration time (time-accuracy trade-off)
  runtime <- system.time({
    # Leave-One-Project-Out Cross-Validation
    for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
      # Define the testing set
      test <- df_split[[project]] # TODO: Inconsistent (partial matching vs explicit)
      # Define the training set
      train <- df[!(df$pid %in% project), ]
      # Fit the model to the data to get the ensemble
      ensemble <- train_adaboost(train[,features], train[,target],
                                 n_trees = hgrid$n_tree[i],  
                                 eta = hgrid$eta[i], 
                                 tree_hyperpar = rpart::rpart.control(
                                   maxdepth = hgrid$maxdepth[i]))
      # Make predictions on validation set 
      y_pred <- test_adaboost(test[,features], ensemble)$class
      # Store the performance evaluation (level: effects)
      hgrid[[project]][i] <- AUC::auc( AUC::roc(y_pred, test$replicated)) }
    # Calculate the average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i,4:8]))
  }) 
  # TODO: Inconsistent (partial matching vs explicit)
  hgrid$runtime[i] <- runtime[["elapsed"]]
}
# Close connection 
close(pb)
```

Performance evaluation sorted by highest AUC.

```{r}
# Note: EE is highly anomalous!
hgrid[order(-hgrid$auc.mean), ] 
```

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1.

## Optimized Ensemble 

If we do not stack the results, we should at least fit the model with the
hyperparameters that had the best average performance (here AUC) across all
replication projects. However, we don't get a true sense of the model's
out-of-sample capabilities because we don't have another project it hasn't seen
yet.

```{r}
# Define the testing set
test <- df_split[[project]] # TODO: Maybe inconsistent
# Define the training set
train <- df[!(df$pid %in% project), ]
# Fit the model to the data to get the ensemble
ensemble <- train_adaboost(train[,features], train[,target], 
                           n_trees = 1e4,  eta = 1e-3, 
                           tree_hyperpar = rpart::rpart.control(
                                 maxdepth = 3))
# Make predictions on validation set 
y_pred <- test_adaboost(test[,features], ensemble)$class
```

# Module 5: Stacking (?) 

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1. 
