---
title: "Code Along Script For The Tutorial"
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Code Along Script For The Tutorial}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Prerequisites

#### Listing 1: Code to install the adatutor package form GitHub, load and attach it.

```{r}
# Load package
library(remotes)

# Install package from GitHub
# install_github("sbissantz/adatutor")

# Load package
library(adatutor)
```

# Empirical Example

#### Listing 2: Code to load the dataset.

```{r}
# Load data
data(altmejd)
```

#### Listing 3: Code to prepare the data.

```{r}
# Reviewer metrics (names)
prednms <- c("power.o", "effect_size.o", "n.o", "p_value.o")

# Outcome variable (name)
outnme <- "replicate"

# Variables of interest
voinms <- c(prednms, outnme)

# Shorten 'altmejd'
df <- altmejd
```

# Module 1: Data Splitting

#### Listing 4: Code to do a 70/30 simple random split.

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.7

# Number of effects
n_fx <- nrow(df)

# Number of training samples
n_train <- round(n_fx * prop)

# Training index
train_idx <- sample(seq_len(n_fx), n_train)

# Training set
train <- df[train_idx, ]

# Test set (incl. validation)
testvalid <- df[-train_idx, ]
```

#### Listing 5: Code to complete a 70/15/15 simple random split

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.5

# Number of effects
n_fx <- nrow(testvalid)

# Number of validation samples
n_valid <- round(n_fx * prop)

# Training index
valid_idx <- sample(seq_len(n_fx), n_valid)

# Validation set
valid <- testvalid[valid_idx, ]

# Test set
test <- testvalid[-valid_idx, ]
```

#### Listing 6: Code to check for imbalances in the outcome variable of the original dataset and the test set.

```{r}
# Frequency tables
tbl_orig <- table(df$replicate)
tbl_testvalid <- table(testvalid$replicate)

# Relative frequencies
proportions(tbl_orig)
proportions(tbl_testvalid)
```

#### Listing 7: Code to do a 70/30 stratified split.

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.7

# Partition data based on outcome
set_split <- split(df, df$replicate)

# Number of validation samples
n_train <- lapply(set_split, function(set) round(nrow(set) * prop))

# Training index
strat_idx <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_train
)

# Training index
train_idx <- df$eid %in% unlist(strat_idx)

# Training set
train <- df[train_idx, ]

# Test set (incl. validation)
testvalid <- df[!train_idx, ]
```

#### Listing 8: Code to complete a 70/15/15 stratified split

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.5

# Partition data based on outcome
set_split <- split(testvalid, testvalid$replicate)

# Numbers of validation samples
n_valid <- lapply(set_split, function(set) round(nrow(set) * prop))

# Validation index
strat_idx <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_valid
)
valid_idx <- testvalid$eid %in% unlist(strat_idx)

# Validation set
valid <- testvalid[valid_idx, ]

# Test set
test <- testvalid[!valid_idx, ]
```

#### Listing 9: Code to do a 70/30 random and stratified split, with the pre-implemented functions of the adatutor package.

```{r}
# For reproducibility
set.seed(112)

# Simple random split
randsplit <- random_split(df, prop = 0.7)
train <- randsplit$set1
testvalid <- randsplit$set2

# Stratified split
stratsplit <- stratified_split(df, group = "replicate", id = "eid", prop = 0.7)
train <- stratsplit$set1
testvalid <- stratsplit$set2
```

#### Listing 10: Code to do a 70/30 random and stratified split, with the rsample package.

```{r}
# Load package
library(rsample)

# For reproducibility
set.seed(112)

# Simple random split
randsplit <- initial_split(df, prop = 0.7)
train <- training(randsplit)
testvalid <- testing(randsplit)

# Stratified split
stratsplit <- initial_split(df, prop = 0.7, strata = "replicate")
train <- training(stratsplit)
testvalid <- testing(stratsplit)
```

#### Figure 2

```{r}
# Where the manuscript keeps its figures. Most readers will not have this
# folder, and then the plots just go to the screen as usual. Each export wraps
# the same call rather than repeating it, so the file and the figure you see on
# screen cannot drift apart.
figdir <- "./inst/figures"
to_file <- dir.exists(figdir)

# One column of the paper's two-column layout, and the full text width, both
# measured from the typeset manuscript. Drawing at the size a figure is printed
# at means nothing is scaled afterwards, which is what keeps the labels legible:
# a figure drawn at the full width and then squeezed into a column loses half
# its type size. The two decision-boundary plots take the wider pair, and they
# are the two the manuscript already sets as `figure*` -- at column width the
# studies collide into one band and the boundary smears.
col_w <- 3.44 * 1.25
col_h <- 2.45 * 1.25
full_w <- 6.94 * 0.8
full_h <- 4.94 * 0.8

# Select only two predictors
# Note: Reduce run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <- train[, union(subfeat, outnme)]
# Hyperparameter setup
treehypar <- rpart::rpart.control(
  # Maximum depth of any node tree
  maxdepth = 1,
  # Minimum n° observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion
  split = "gini"
) # or ID3: split = "information"
# Create a decision tree
h <- rpart::rpart(
  replicate ~ .,
  data = subdf,
  method = "class",
  control = treehypar
)
# Visualize
sty <- tree_colors(h)

if (to_file) {
  pdf(
    file.path(figdir, "cstump.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

rpart.plot::rpart.plot(
  h,
  extra = 102,
  digits = 4,
  box.col = sty$box,
  col = sty$text
)

if (to_file) {
  dev.off()
}
```

::: {.callout-important}
## Important note

We suggest running Listings 7 & 8 again, before proceeding with the other
modules. This should guarantee the same training, validation, and test set.
:::

# Module 2: Classification Stumps

#### Listing 11: Code to retrieve the frequency tables for the outcome variable under the two power conditions.

```{r}
# Condition for the split
splitcond <- train$power.o < 0.9305

# Studies with power.o < 0.9305
table(train$replicate[splitcond])

# Studies with power.o >= 0.9305
table(train$replicate[!splitcond])
```

#### Listing 12: Code to set the hyperparameters of a classification tree and fit the model to the training data.

```{r}
# Load package
library(rpart)

# Hyperparameter setup
treehypar <- rpart.control(
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  split = "gini"
)

# Fit classification stump
cstump <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  control = treehypar
)
```

#### Listing 13: Code to visualize a fitted classification stump.

```{r}
# Load package
library(rpart.plot)

# Visualize classification stump
sty <- tree_colors(cstump)
rpart.plot::rpart.plot(
  cstump,
  extra = 102,
  digits = 4,
  box.col = sty$box,
  col = sty$text
)
```

#### Figure 3

```{r}
# `dboundary`, not `cstump`: in the manuscript `cstump` is the tree diagram
# above, and this is the boundary that tree draws. Both are `figure*` floats
# there, which is why they get the wider geometry.
if (to_file) {
  pdf(
    file.path(figdir, "dboundary.pdf"),
    width = full_w,
    height = full_h,
    pointsize = 9
  )
}

# Visualize the stump's decision boundary. `shade = "margin"` colours by the
# predicted probability rather than by the class alone, so the two regions read
# as the moderate .26 and .76 they actually are -- and it is the same encoding
# Figure 9 uses, which is what makes the two figures comparable.
plot_adabound(
  h,
  data = subdf,
  shade = "margin",
  xlab = "Power (original study)",
  ylab = "Sample size (original study)"
)

if (to_file) {
  dev.off()
}
```

#### Figure 4

```{r}
if (to_file) {
  pdf(
    file.path(figdir, "gini.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

op <- par(
  mar = c(2.9, 2.9, 1.9, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

# a single curve, so one color rather than a scale -- the same anchor viridis
# uses for the first/darkest entry elsewhere in the tutorial
col <- viridisLite::viridis(1, end = 0.85)

# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01) # Values for fraction of class k
y <- 1 - (x^2) - (1 - x)^2 # Gini Impurity calculation
# Create the plot
plot(
  x,
  y,
  type = "l",
  lwd = 2, # Specify line plot
  col = col,
  xlab = "Proportion of Successes",
  ylab = "Impurity"
)
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1))

par(op)
if (to_file) {
  dev.off()
}
```

#### Listing 14: Code to calculate means of neighboring units and find the position of the relevant power value.

```{r}
# Sorted vector
power_srt <- sort(train$power.o)

# Adjacent means
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2

# Find position of value '0.9305'
which(round(adjmeans, 4) == 0.9305)

# Additional

adj_pat <- which(round(adjmeans, 4) == 0.9305)
const <- adjmeans[adj_pat]
N_total <- nrow(train) # 107
criterion_left <- train$power.o < const
N_left <- length(train$replicate[criterion_left]) # 66 total
left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1) # 17
left_failure <- N_left - left_success # 49 failures
# Gini impuritiy (left)
gini_left <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
gini_left # 0.382
criterion_right <- train$power.o >= const
N_right <- length(train$replicate[criterion_right]) # 41
right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1) # 31
right_failure <- N_right - right_success # 10
# Gini impuritiy (right)
gini_right <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
gini_right # 0.3688281
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) *
  gini_left +
  (N_right / N_total) * gini_right
round(gini_impurity, digits = 5) # 0.37724
# Sorted values
power_srt <- sort(train$power.o)
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2
gini_impurity <- vector(length = length(adjmeans))
for (i in seq_along(adjmeans)) {
  constant <- adjmeans[i]
  N_total <- length(y_train)
  # Left leaf
  criterion_left <- train$power.o < constant
  N_left <- length(train$replicate[criterion_left])
  left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- train$power.o >= constant
  N_right <- length(train$replicate[criterion_right])
  right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
  gini_d2 <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
  # Impurity
  gini_impurity[i] <- (N_left / N_total) *
    gini_d1 +
    (N_right / N_total) * gini_d2
}
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjmeans[which.min(gini_impurity)]
```

#### Listing 15: Code to make the assumption of equal weights explicit.

```{r}
# Load package
library(rpart.plot)

# Number of rows
m <- nrow(train)

# Equal weight to each observation
D <- rep(1, m) / m

# Fit decision stump
cstump <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D,
  control = treehypar
)

# Visualize
sty <- tree_colors(cstump)
rpart.plot(cstump, digits = 4, box.col = sty$box, col = sty$text)
```

#### Listing 16: Code to show the sensitivity to observation weights by using normalized random weights from a uniform distribution.

```{r}
# For reproducibility
set.seed(112)

# Draws from standard uniform
D_raw <- runif(107)

# Normalize (sum to 1)
D <- D_raw / sum(D_raw)

# Create decision stump
cstump_ast <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D,
  control = treehypar
)

# Visualize
sty <- tree_colors(cstump_ast)

if (to_file) {
  pdf(
    file.path(figdir, "cstumpwght.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

rpart.plot(
  cstump_ast,
  extra = 100,
  digits = 4,
  box.col = sty$box,
  col = sty$text
)

if (to_file) {
  dev.off()
}
```

# Module 3: AdaBoost

#### Listing 17: Code to understand the givens in the AdaBoost algorithm.

```{r}
# Predictors and outcome (names)
prednms
outnme

# Predictor matrix
x_train <- train[, prednms]

# Outcome vector
y_train <- train[, outnme]

# Indices
i <- 1
m <- nrow(train)

# i-th row or element
x_train[i, ]
y_train[i]

# i-th to m-th row (i.e., all)
cbind(x_train, y_train)[i:m, ]
```

#### Listing 18: Code to generate initialization weights in the AdaBoost algorithm.

```{r}
# Initialization weights
D1 <- rep(1, m) / m
```

#### Listing 19: Code to fit a classification stump with the initialization weights.

```{r}
# Hyperparameter setup
treehypar <- rpart.control(maxdepth = 1, split = "gini")

# Fit classification stump
h1 <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D1,
  control = treehypar
)
```

#### Listing 20: Code to calculate the classification stump's error.

```{r}
# Retrodictions (training data)
yretro <- predict(h1, newdata = train[, voinms], type = "class")

# Retrodictive error
epsilon <- sum(D1 * (yretro != y_train))
```

#### Listing 21: Code to calculate the classification stump's performance.

```{r}
# Retrodivitve performance
perf <- sum(D1 * (yretro == y_train))

# Retrodictive error
epsilon <- 1 - perf
```

#### Listing 22: Code to calculate the classification stump's weight assuming a learning rate eta of 1.

```{r}
# Learning rate
eta <- 1

# Model weight
alpha1 <- 1 / 2 * log(perf / epsilon) * eta
```

#### Figure 6

```{r}
if (to_file) {
  pdf(
    file.path(figdir, "importance.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

# Tighter margins than the default: at this size the defaults would leave
# almost no room to plot in. `par(op)` at the end puts them back, because the
# whole document runs in one session and the settings would otherwise leak
# into every later figure.
op <- par(
  mar = c(2.9, 2.9, 1.9, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

# three ordered learning rates, so the palette runs with them. `end = 0.85`
# stops short of viridis's yellow, which is invisible as a line on white
pal <- viridisLite::viridis(3, end = 0.85)

fat <- function(x, eta) 1 / 2 * log(x / (1 - x)) * eta
curve(
  fat(x, 1),
  from = 0.5,
  to = 1,
  lwd = 2,
  col = pal[1],
  ylab = "Importance",
  xlab = "Performance"
)
grid()
curve(
  fat(x, 0.5),
  from = 0.5,
  to = 1,
  lty = 2,
  lwd = 2,
  col = pal[2],
  add = TRUE
)
curve(
  fat(x, 0.1),
  from = 0.5,
  to = 1,
  lty = 4,
  lwd = 2,
  col = pal[3],
  add = TRUE
)
# The legend sits in the margin above the panel, where `main` would go, so it
# never covers a curve. `xpd = NA` is what lets it draw outside the plotting
# region; the default top margin already reserves the room. `text.width = NA`
# gives each entry its own width -- without it `horiz` pads every column out to
# the widest label, which strings short entries across the whole panel.
usr <- par("usr")
legend(
  x = (usr[1] + usr[2]) / 2,
  y = usr[4] + 0.055 * (usr[4] - usr[3]),
  xjust = 0.5,
  yjust = 0,
  horiz = TRUE,
  xpd = NA,
  bty = "n",
  text.width = NA,
  x.intersp = 0.7,
  seg.len = 1.2,
  cex = 0.9,
  legend = c(
    expression(eta * " = 1"),
    expression(eta * " = 0.5"),
    expression(eta * " = 0.1")
  ),
  col = pal,
  lty = c(1, 2, 4),
  lwd = 2
)

par(op)
if (to_file) {
  dev.off()
}
```

#### Listing 23: Code to calculate the update factor for the initialization weights.

```{r}
# Indicator trick
chi1 <- (yretro == y_train) * 1 + (yretro != y_train) * -1

# Unnormalized weights
D2_raw <- D1 * exp(-alpha1 * chi1)

# Normalization constant
Z1 <- sum(D2_raw)

# Normalized weights
D2 <- D2_raw / Z1
```

#### Figure 7

```{r}
if (to_file) {
  pdf(
    file.path(figdir, "updatefactor.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

op <- par(
  mar = c(2.9, 2.9, 1.9, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

# same three-step palette as Figure 6: the model weights are ordered too
pal <- viridisLite::viridis(3, end = 0.85)

feat <- function(at, x) exp(-at * x)
curve(
  feat(0.719, x),
  from = -1,
  to = 1,
  ylim = c(0.3, 2.2),
  lty = 3,
  col = pal[1],
  xlab = "Classification",
  ylab = "Weight update factor",
  xaxt = "n"
)
grid()
points(
  c(1, -1),
  c(feat(0.719, 1), feat(0.7, -1)),
  pch = 20,
  cex = 2,
  col = pal[1]
)
curve(
  feat(0.619, x),
  from = -1,
  to = 1,
  ylim = c(0, 2),
  col = pal[2],
  lty = 3,
  add = TRUE
)
points(
  c(1, -1),
  c(feat(0.619, 1), feat(0.6, -1)),
  pch = 17,
  cex = 2,
  col = pal[2]
)
curve(
  feat(0.519, x),
  from = -1,
  to = 1,
  ylim = c(0, 2),
  col = pal[3],
  lty = 3,
  add = TRUE
)
points(
  c(1, -1),
  c(feat(0.519, 1), feat(0.51, -1)),
  pch = 15,
  cex = 2,
  col = pal[3]
)
abline(h = 1, lty = 2, lwd = 1.5, col = "gray30")
usr <- par("usr")
legend(
  x = (usr[1] + usr[2]) / 2,
  y = usr[4] + 0.055 * (usr[4] - usr[3]),
  xjust = 0.5,
  yjust = 0,
  horiz = TRUE,
  xpd = NA,
  bty = "n",
  text.width = NA,
  x.intersp = 0.7,
  seg.len = 1.2,
  cex = 0.9,
  legend = c(
    expression(a[t] * " = 0.719"),
    expression(a[t] * " = 0.619"),
    expression(a[t] * " = 0.519")
  ),
  col = pal,
  lty = c(1, 1, 1),
  pch = c(20, 17, 15),
  lwd = 2
)
axis(1, at = c(-1, 1))

par(op)
if (to_file) {
  dev.off()
}
```

#### Figure 8

::: {.callout-note}
## What the yellow means here

The yellow band marks the studies the stump got **wrong**, which is not what
yellow means in Figures 3 and 9. There it is the bright end of the scale over
the ensemble's margin, so it reads as a predicted replication *success*. Here it
is a background highlight rather than a data point's colour, and it marks the
studies whose weight is about to rise.
:::

```{r}
if (to_file) {
  pdf(
    file.path(figdir, "weightone.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

op <- par(
  mar = c(2.9, 2.9, 1.9, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

n_25 <- 25 # TODO: Do we need this?
scaling <- 2

# D1 and D2 keep a colour each. `end = 0.5` leaves D2 teal rather than
# yellow-green, which would go muddy against the yellow band below
pal <- viridisLite::viridis(2, end = 0.5)
band <- adjustcolor(viridisLite::viridis(1, begin = 1), 0.35)

# area, not diameter: cex scales the width of the point, so the eye reads a
# doubled weight as four times the ink unless we take the square root
bubble <- function(w) sqrt(w / D1[1]) * scaling

wrong <- which(y_train[1:n_25] != yretro[1:n_25])
ylim <- c(0, D1[1] + 0.012)

plot(
  c(0.5, n_25 + 0.5),
  ylim,
  type = "n",
  xaxt = "n",
  xlab = "Data point",
  ylab = "Weight"
)
# the band goes behind everything: misclassification belongs to the data point,
# not to either of its two weights
rect(wrong - 0.42, ylim[1], wrong + 0.42, ylim[2], col = band, border = NA)
box()
# every fifth point: 25 labels do not fit the width of a column
axis(1, at = c(1, seq(5, n_25, 5)))
for (i in 1:n_25) {
  lines(c(i, i), c(D1[i], D2[i]), lty = 3, col = "grey45")
}

points(
  1:n_25,
  D1[1:n_25],
  pch = 21,
  bg = pal[1],
  col = "white",
  cex = bubble(D1[1:n_25]),
  lwd = 1.3
)
points(
  1:n_25,
  D2[1:n_25],
  pch = 21,
  bg = pal[2],
  col = "white",
  cex = bubble(D2[1:n_25]),
  lwd = 1.3
)
usr <- par("usr")
legend(
  x = (usr[1] + usr[2]) / 2,
  y = usr[4] + 0.055 * (usr[4] - usr[3]),
  xjust = 0.5,
  yjust = 0,
  horiz = TRUE,
  xpd = NA,
  bty = "n",
  text.width = NA,
  x.intersp = 0.7,
  seg.len = 1.2,
  cex = 0.9,
  legend = c(expression(D[1]), expression(D[2]), "misclassified"),
  pt.bg = c(pal, band),
  col = c("white", "white", "grey60"),
  pch = c(21, 21, 22),
  pt.cex = 1.3
)

par(op)
if (to_file) {
  dev.off()
}
```

#### Listing 24: Code to loop through the algorithmic steps. Note that the code can take several seconds to complete.

```{r}
# Hyperparameter setup
treehypar <- rpart.control(maxdepth = 1)

# Number of trees
T <- 1000

# Learning rate
eta <- 1

# Lists for models and weights
fit <- vector("list", T)

# Initial observation weights
D <- rep(1, m) / m

# Loop through steps
for (t in seq_len(T)) {
  h <- rpart(
    replicate ~ .,
    data = train[, voinms],
    method = "class",
    weights = D,
    control = treehypar
  )
  yretro <- predict(object = h, newdata = train[, prednms], type = "class")
  e <- sum(D * (yretro != y_train))
  a <- 0.5 * log((1 - e) / e) * eta
  chi <- (yretro == y_train) * 1 + (yretro != y_train) * -1
  D_raw <- (D * exp(-a * chi))
  D <- D_raw / sum(D_raw)
  fit[[t]] <- list("h" = h, "a" = a)
}
names(fit) <- paste0("t", seq_len(T))

# Show first model weight
cat("Model weight 1:", fit$t1$a)
```

#### Listing 25: Code to generate ensemble retrodictions.

```{r}
# Boosted classification stumps
h <- lapply(fit, "[[", "h")

# Model weights
a <- sapply(fit, "[[", "a")

# Retrodictions from stumps
yretro_stumps <- sapply(h, predict, newdata = train[, prednms], type = "vector")

# Transformed values (-1,1)
yretro_stumps <- 2 * yretro_stumps - 3

# Retrodictions from full model
yretro_ada <- sign(a %*% t(yretro_stumps))
```

#### Listing 26: Code to replace matrix multiplication with a loop.

```{r}
#for (m in seq_len(m)) {
#yretro_ada[m] <- sign(sum(a * yretro_stumps[m, ]))
#}
```

#### Figure 9

```{r}
# Tree hyperparameter setting
treehypar <- rpart::rpart.control(
  # Slightly different hyperparameters to improve visibility of boundary
  maxdepth = 2,
  minsplit = 5,
  minbucket = round(5 / 3),
  cp = 1e-2,
  split = "gini"
)

# Fit AdaBoost on the two features
fit <- trainAda(
  replicate ~ .,
  data = train[, c("replicate", "power.o", "n.o")],
  T = 5e2,
  eta = 1,
  treehypar = treehypar,
  verbose = FALSE
)

# Both decision-boundary plots span the full text width: at column width the
# studies collide into a single band and the boundary stops reading as a line
if (to_file) {
  pdf(
    file.path(figdir, "boundaryada.pdf"),
    width = full_w,
    height = full_h,
    pointsize = 9
  )
}

# Visualize the ensemble's decision boundary
plot_adabound(
  fit,
  data = train,
  shade = "margin",
  xlab = "Power (original study)",
  ylab = "Sample size (original study)"
)

if (to_file) {
  dev.off()
}
```

#### Listing 27: Code to evaluate the retrodictive performance of the ensemble.

```{r}
# Confusion matrix
table(y_train, yretro_ada) # Perfect retrodictions
```

#### Listing 28: Code to generate ensemble predictions.

```{r}
# Outcome variable (test set)
y_test <- test[, outnme]

# Predictions (test set)
ypred_stumps <- sapply(h, predict, newdata = test[, prednms], type = "vector")
ypred_stumps <- 2 * ypred_stumps - 3
ypred_ada <- sign(a %*% t(ypred_stumps))
```

#### Listing 29: Code to evaluate the predictive performance of the ensemble.

```{r}
# Confusion matrix
table(y_test, ypred_ada) # Inperfect predictions
```

# Module 4: Leave-Project-Out Cross-Validation and Hyperparameter Optimization

## Two Wrapper Functions

#### Listing 30: Code introducing the wrapper functions. Note that we reduced the number of trees to speed up runtime in the upcoming sections.

```{r}
# Boosted stumps and weights
fit <- trainAda(
  replicate ~ .,
  data = train[, voinms],
  # Boosting hyperparameter
  T = 100,
  eta = 1,
  # Tree hyperparameter
  treehypar = treehypar
)

# Predictions on test set (class labels, for the confusion matrix)
ypred <- testAda(fit, data = test[, prednms], type = "class")

# Predictions on test set (margins, for the AUC)
ymargin <- testAda(fit, data = test[, prednms], type = "margin")
```

::: {.callout-important}
## `type = "class"` gives balanced accuracy, not AUC

A two-valued score collapses the ROC curve to a single interior point, whose
area is exactly (sensitivity + specificity) / 2. Use `type = "margin"` for the
AUC; keep `type = "class"` for the confusion matrix.
:::

## Evaluating Model Performance

#### Listing 31: Code showing different measures to evaluate the predictive performance of the ensemble.

```{r}
# Confusion matrix
confusion <- table(ypred, y_test)

# Accuracy
sum(diag(confusion)) / sum(confusion)

# Area under the curve (scored on the margin)
auroc(y_test, ymargin)

# Base rates
df$replicate_01 <- as.integer(df$replicate) - 1
baserates <- aggregate(replicate_01 ~ pid, df, mean)

# Average base rate
mean(baserates$replicate_01)
```

::: {.callout-important}
## Switched to `PRROC` package
Cite and fix everywhere - especially in the tutorial.
:::

::: {.callout-note}
## Where the AUC comes from

`auroc()` is a thin wrapper around `PRROC::roc.curve()` (Grau, Grosse &
Keilwagen), which does the actual work; `auprc()` wraps `pr.curve()` from the
same package. The wrapper exists so the outcome can be passed in whatever form
you have it -- a factor, 0/1, -1/1 or logical. `roc.curve()` itself expects a
0/1 vector and returns a wrong number without complaint if handed a 1/2 one, so
the conversion is worth doing once, in one place.

:::

## Leave-Project-Out Cross-Validation

#### Listing 32: Code to do a leave-project-out cross-validation. Note that this ran for approximately 2 minutes on our laptops.

```{r}
# Partition data by project ID
df_split <- split(df, df$pid)

# List for AUC results
auc <- list()

runtime <- system.time({
  # LPO-CV
  for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
    message("\nThe ", paste0(rp), " model.\n")
    # Training and test set
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]
    # Fit model to data
    fit <- trainAda(
      replicate ~ .,
      data = train[, voinms],
      T = 1000,
      eta = 1,
      treehypar = treehypar
    )
    # Predictions (left-out project), on the margin so the AUC sees the ranking
    ypred <- testAda(fit, data = test[, prednms], type = "margin")
    # Performance evaluation
    auc[[rp]] <- auroc(test$replicate, ypred)
  }
})

# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")

# Results
# TODO: From mean to plot?
c(unlist(auc), "mean" = mean(unlist(auc)))
```

## Hyperparameter Optimization

#### Listing 33: Code to define a hypergrid.

```{r}
# Define hypergrid
hgrid <- expand.grid(
  # Tree-specific:
  # Tree depth
  "depth" = 1:3,
  # Boosting specific:
  # Number of trees
  "T" = c(10, 100, 1000),
  # Learning rate
  "eta" = c(1, 0.1),
  # Container for fold performance
  "eerp" = NA,
  "ml1" = NA,
  "ml3" = NA,
  "rpp" = NA,
  "ssrp" = NA,
  "auc.mean" = NA,
  "runtime" = NA
)
```

#### Listing 34: Code to do hyperparameter optimization using a leave-project-out cross-validation strategy. Note that this ran for approximately 3 minutes on our laptops.

```{r}
# Number of rows
hgrid_rows <- nrow(hgrid)

# Partition data by project ID
df_split <- split(df, df$pid)

# Hyperparameter optimization
for (i in seq_len(hgrid_rows)) {
  # Iteration time
  runtime <- system.time({
    # Leave-project-out CV
    for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
      # Progress report
      cat(
        "\rProgress: ",
        round(i / hgrid_rows * 100),
        "%",
        " | T: ",
        hgrid$T[i],
        " | eta: ",
        hgrid$eta[i],
        " | depth: ",
        hgrid$depth[i],
        " | model: ",
        rp
      )
      # Training and test set
      test <- df_split[[rp]]
      train <- df[!(df$pid %in% rp), ]
      # Fit model to data
      fit <- trainAda(
        replicate ~ .,
        data = train[, voinms],
        T = hgrid$T[i],
        eta = hgrid$eta[i],
        treehypar = rpart.control(maxdepth = hgrid$depth[i]),
        verbose = FALSE
      )
      # Predictions (on the margin, so the AUC sees the ranking)
      ypred <- testAda(
        fit,
        data = test[, prednms],
        type = "margin",
        verbose = FALSE
      )
      # Performance evaluation
      hgrid[[rp]][i] <- auroc(test$replicate, ypred)
    }
    # Average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i, 4:8]))
  })
  hgrid$runtime[i] <- runtime[["elapsed"]]
} #

# Hypergrid sorted by average AUC
hgrid[order(-hgrid$auc.mean), ]
```

## Selecting versus Evaluating

Listing 34 scored every setting in the hypergrid on all five projects, and we
took the row with the best average. That number is not an estimate of how the
chosen model will do on a new project: the same five projects picked the winner
and then graded it. To see how much that matters, the selection has to happen
where the project being scored cannot influence it.

Nested leave-project-out cross-validation does exactly that. The outer loop holds
out one project. The inner loop repeats the whole grid search on the four that
remain, choosing hyperparameters without ever seeing the held-out project. The
selected setting is then refit on those four projects and scored on the fifth.
Repeat five times.

The result is easy to misread, so it is worth being precise about what it
estimates. **Nested LPO-CV does not give you a performance estimate for a
particular hyperparameter setting.** It estimates the *procedure* -- "run the
grid search, then fit the winner." Every model built inside the nesting is
discarded; none of them is the model you ship. You still fit one setting on all
five projects, as in Listing 35. Nesting tells you what to expect from that
model, not what it scores.

Two things follow. First, the five inner searches may each select a different
setting, and the setting that won in Listing 34 may be chosen by none of them.
That is expected rather than alarming: with five projects there is no stable
"best" setting to find, and seeing that instability directly is more informative
than the average that hides it. Second, the nested estimate is *conservative*
here, because each nested model saw four projects while the model you ship sees
five.

If nesting does not hand you a model, what does? The three listings have three
different jobs, and it is worth being explicit about which is which:

| job | listing |
|:----|:--------|
| Choose the hyperparameters, using all five projects | Listing 34 |
| Estimate what that choosing procedure is worth | Listing 35 |
| Fit the model you will actually use | Listing 36 |

So Listing 34 still picks the setting and Listing 36 still fits it on the full
dataset -- nesting changes neither. What it changes is the number you quote
alongside that model: not the winning average from Listing 34, which the same
five projects produced by selecting it, but the nested estimate, which no project
had a hand in choosing.

The settings the inner searches picked are worth looking at, but they are a
diagnostic rather than a shortlist. They have the same standing as the
coefficients fitted inside each cross-validation fold back in Module 1: nobody
averages those, and nobody asks which fold's coefficients to keep. They are
by-products of a procedure being measured, not candidates for shipping.

::: {.callout-tip}
## Set the depth with `maxdepth`, not with `treehypar`

`rpart.control(maxdepth = 3)` returns *every* `rpart` default, not just the depth
you named -- including `cp = 0.01`, `xval = 10` and `maxsurrogate = 5`. Passed
through `treehypar`, those overwrite the settings `trainAda()` chose for
boosting. `trainAda()` keeps `cp` at `rpart`'s own 0.01 precisely so that the two
forms fit the *same* trees: we checked, and the trees, the model weights and the
margins are identical. What still differs is `xval = 10`, ten internal
cross-validations per weak learner that nothing reads, costing about thirty per
cent of the runtime.

So this is a tidiness point, not a correctness one. Use the `maxdepth` argument
for depth, and reserve `treehypar` for controls you genuinely mean to change --
`minsplit` and `minbucket` are the ones that actually change how the trees grow.
:::

#### Listing 35: Code to do a nested leave-project-out cross-validation. Note that this ran for approximately 10 minutes on our laptops.

```{r}
# Projects, in the order they take their turn
projects <- c("eerp", "ml1", "ml3", "rpp", "ssrp")

# Partition data by project ID
df_split <- split(df, df$pid)

# One honest score and one selected setting per project
nested_auc <- list()
selected <- list()

runtime <- system.time({
  # OUTER loop: evaluates
  for (rp in projects) {
    # Progress report
    message("\nOuter fold: ", rp, "\n")

    # Training and test set. The test project leaves the data entirely, so
    # nothing below can see it.
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]

    # INNER loop: selects. This is the same grid search as Listing 34, run on
    # the training projects only -- so we can hand it to lpocv().
    inner <- lpocv(
      replicate ~ power.o + effect_size.o + n.o + p_value.o,
      data = train,
      group = "pid",
      grid = hgrid[, c("depth", "T", "eta")]
    )

    # Rank the settings by their mean over the inner folds. Here the mean is a
    # ranking device, not a performance estimate.
    inner_auc <- summary(inner, metric = "auroc")
    ranked <- aggregate(estimate ~ setting, data = inner_auc, FUN = mean)

    # The setting the four training projects chose, on their own
    best <- hgrid[ranked$setting[which.max(ranked$estimate)], ]
    selected[[rp]] <- best[, c("depth", "T", "eta")]

    # Refit at that setting and score the untouched project
    fit <- trainAda(
      replicate ~ .,
      data = train[, voinms],
      T = best$T,
      eta = best$eta,
      maxdepth = best$depth,
      verbose = FALSE
    )
    ypred <- testAda(
      fit,
      data = test[, prednms],
      type = "margin",
      verbose = FALSE
    )
    nested_auc[[rp]] <- auroc(test$replicate, ypred)
  }
})

# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")

# The honest expectation, per project
unlist(nested_auc)

# What each outer fold selected, and how often they disagree
do.call(rbind, selected)
```

Only the outer loop is new. The inner one is the leave-project-out search we
already ran in Listing 34, so we hand it to `lpocv()` rather than write it out
again -- the single change is that it now sees `train` instead of `df`.

The whole procedure is also one call, which is what we use from here on:

```{r}
nested <- lpocv(
  replicate ~ power.o + effect_size.o + n.o + p_value.o,
  data = df,
  group = "pid",
  grid = hgrid[, c("depth", "T", "eta")],
  nested = TRUE,
  criterion = "auroc"
)
nested
```

## Finalizing an Ensemble

#### Listing 36: Code to fit the final model to the data.

```{r}
# Fit optimized model
fit_final <- trainAda(
  replicate ~ .,
  data = df[, voinms],
  T = 10,
  eta = 1,
  treehypar = rpart.control(maxdepth = 1)
)
```
