---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(adatutoR)
```


# Module 1: AdaBoost


## Dataset

Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Load the data
data("altmext723")

# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")

# Outcome (classification task) 
target <- "replicated"

# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")

# Note: No missing values 
df <- altmext723[,voi]
```

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

```{r}
# Reproducibility
set.seed(112)

# Split proportion
sprop <- 0.7

# Number of effects 
n_effect <- nrow(df)

# Number of training samples
n_train <- round(n_effect * sprop)

# Training index
train_idx <- sample(seq_len(n_effect), n_train)

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

However, there are imbalances in the data set. 

```{r}
table(df$replicated)
```

Therefore we should do a stratified split. This means we need to partition the
data into groups based on a stratum variable - here: replicated, and then draw
a random sample from each stratum.

```{r}
# Reproducibility
set.seed(112)

# Partition the data based on the outcome variable
df_split <- split(df, df$replicated)

# Number of training samples (stratified)
n_train <- lapply(df_split, function(df) round(nrow(df) * sprop))

# Training index (stratified)
strat_idx <- mapply(function(df, n) sample(rownames(df), n), df_split, n_train)

# Training index 
train_idx <- as.numeric(unlist(strat_idx))

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

Now the training and testing set have approximately the same proportion of
replication and failure as the original outcome variable. We could also use the
`rsample` package for that.

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```

## Predictions

First we fit the model to the data (training) and then test its predictive
capabilities on unseen data (testing). Since we will use the data to find
the best hyperparameters this is better thought of as validation set.

### Training 

Look under the hood of the function 

```{r}
train_adaboost
```

Tree hyperparameter settings

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 
```

### Illustration of learning the weights across iterations (Daniel)

```{r}
a <- sapply(ensemble, "[[", "learner_weight")
D <- t(sapply(ensemble, "[[", "observation_weight"))
plot(a, type = "l", ylim = c(0, .04), ylab = "Weights", xlab = "Trees", col = "red")

for(i in 1:ncol(D)) 
  lines(1:nrow(D), D[,i] + runif(1, -.00014, +.00014), col = adjustcolor(1, alpha.f = 0.1))

# DANIEL: why do the weights of the trees have these regular peaks for the sequential updates?
```
 
Daniel: Das Updating der Gewichtung der Beobachtungen pro Tree könnte man auch
illustrieren durch eine Art Heat Map (x = Trees, y = Observations, degree of
gray = weighting). Direkt untendrunter könnte man mit einer ähnlichen
graphischen Matrix die correct predictions als black/white Heatmap illustrieren
oder so

### Prediction

```{r}
test_adaboost
```

Make predictions with the trained model and add them to the testing set.

```{r}
y_class <- test_adaboost(X_test, ensemble)$class
test$y_pred <- factor( (y_class + 3) / 2, labels = c("no", "yes") )
```

## Evaluation

```{r}
ROC <- AUC::roc(test$y_pred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```


# Module 2: Cross Validation

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 1e1,  eta = 1e-1, hyptree)
    # Make predictions on validation set 
    ypred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(ypred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

This will run for approximately 4 minutes.

## Performance estimates

```{r}
auc
```

# Module 3: Hyperparameter Optimization

```{r}
# TODO: Increase n_trees 1e2:1e5 and reduce eta 1e-2 : 1e-5
hgrid <- expand.grid(
  # Tree-specific 
  maxdepth = 1:3, # Interaction depth
  # Ensemble specific
  n_tree = c(1e1, 1e2), # Number of trees
  eta = c(1, 0.1), # Learning rate
  # Container for fold performance
  ee = NA, ml1 = NA, ml3 = NA, rpp = NA,  ssrp = NA, 
  auc.mean = NA, runtime = NA) # Average performance
# Number of combinations 
hgrid_rows <- nrow(hgrid)
```

Hyperparameter optimization using Leave-Project-Out Cross-validation.

```{r}
# Partition the data by "Project ID"
df_split <- split(df[,voi], df$pid) 
# Setup a progress bar
pb <- txtProgressBar(0, hgrid_rows, initial = 0, style = 3) 
# For each entry in the hypergrid, do...
for(i in seq_len(hgrid_rows)) {
  # Progress bar
  setTxtProgressBar(pb, i) 
  # Track the iteration time (time-accuracy trade-off)
  runtime <- system.time({
    # Leave-One-Project-Out Cross-Validation
    for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
      # Define the testing set
      test <- df_split[[project]] # TODO: Inconsistent (partial matching vs explicit)
      # Define the training set
      train <- df[!(df$pid %in% project), ]
      # Fit the model to the data to get the ensemble
      ensemble <- train_adaboost(train[,features], train[,target],
                                 n_trees = hgrid$n_tree[i],  
                                 eta = hgrid$eta[i], 
                                 tree_hyperpar = rpart::rpart.control(
                                   maxdepth = hgrid$maxdepth[i]))
      # Make predictions on validation set 
      ypred <- test_adaboost(test[,features], ensemble)$class
      # Store the performance evaluation (level: effects)
      hgrid[[project]][i] <- AUC::auc( AUC::roc(ypred, test$replicated)) }
    # Calculate the average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i,4:8]))
  }) 
  # TODO: Inconsistent (partial matching vs explicit)
  hgrid$runtime[i] <- runtime[["elapsed"]]
}
# Close connection 
close(pb)
```

Performance evaluation sorted by highest AUC.

```{r}
# Note: EE is highly anomalous!
hgrid[order(-hgrid$auc.mean), ] 
```

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1.

## Optimized Ensemble 

If we do not stack the results, we should at least fit the model with the
hyperparameters that had the best average performance (here AUC) across all
replication projects. However, we don't get a true sense of the model's
out-of-sample capabilities because we don't have another project it hasn't seen
yet.

```{r}
# Define the testing set
test <- df_split[[project]] # TODO: Maybe inconsistent
# Define the training set
train <- df[!(df$pid %in% project), ]
# Fit the model to the data to get the ensemble
ensemble <- train_adaboost(train[,features], train[,target], 
                           n_trees = 1e4,  eta = 1e-3, 
                           tree_hyperpar = rpart::rpart.control(
                                 maxdepth = 3))
# Make predictions on validation set 
ypred <- test_adaboost(test[,features], ensemble)$class
```
