---
title: "Code Along Script For The Tutorial"
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Code Along Script For The Tutorial}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Prerequisites

#### Listing 1

Code to install the adatutor package from GitHub, load and attach it.

```{r listing-1}
# Load package
library(remotes)

# Install package from GitHub
# install_github("sbissantz/adatutor")

# Load package
library(adatutor)
```

# Empirical Example

#### Listing 2

Code to load the dataset.

```{r listing-2}
# Load data
data(altmejd)
```

#### Listing 3

Code to prepare the data.

```{r listing-3}
# Reviewer metrics (names)
pred_nms <- c("power.o", "effect_size.o", "n.o", "p_value.o")

# Outcome variable (name)
out_nme <- "replicate"

# Variables of interest
voi_nms <- c(pred_nms, out_nme)

# Shorten 'altmejd'
df <- altmejd
```

# Module 1: Data Splitting

#### Listing 4

Code to do a 70/30 simple random split.

```{r listing-4}
# For reproducibility
set.seed(112)

# Split proportion
prop_train <- 0.7

# Number of effects
n_fx <- nrow(df)

# Number of training samples
n_train <- round(n_fx * prop_train)

# Row positions
fx_seq <- seq_len(n_fx)

# Sample from sequence
fx_smpl <- sample(fx_seq, n_train)

# Inclusion indicator
train_ind <- fx_seq %in% fx_smpl

# Training set
train <- df[train_ind, ]

# Test set (as leftover set)
test <- df[!train_ind, ]
```

::: {.callout-important}
The code has changed. It is not identical to what is in the tutorial.
:::

#### Listing 5

Code to do a 70/15/15 simple random split in one pass.

```{r listing-5}
# For reproducibility
set.seed(112)

# Split proportions, of the whole data
prop <- c(train = 0.70, valid = 0.15, test = 0.15)

# Set sizes, counted from the whole data
n <- round(n_fx * prop)

# Training sample and inclusion indicator
train_smpl <- sample(fx_seq, n[["train"]])
train_ind <- fx_seq %in% train_smpl

# Validation sample and inclusion indicator
valid_smpl <- sample(fx_seq[!train_ind], n[["valid"]])
valid_ind <- fx_seq %in% valid_smpl

# Training set
train <- df[train_ind, ]

# Validation set
valid <- df[valid_ind, ]

# Test set (as rest set)
test <- df[!train_ind & !valid_ind, ]
```

::: {.callout-important}
The code has changed. It is not identical to what is in the tutorial. Need to rewrite the section.
:::

**Side note**: Every count comes from `n_fx`, so `0.15` means fifteen percent of 152 effects. Splitting in two independent steps (as we did earlier) would make the second proportion relative to what remains, and fifteen percent of the whole would then have to be written `0.15 / (1 - 0.70) = 0.5`.

#### Listing 6

Code to check for imbalances in the outcome variable of the original dataset
and the held-out set.

```{r listing-6}
# Frequency tables
tbl_orig <- table(df$replicate)
tbl_out <- table(df$replicate[!train_ind])

# Relative frequencies
proportions(tbl_orig)
proportions(tbl_out)
```

::: {.callout-important}
The code has changed. It is not identical to what is in the tutorial.
:::

#### Listing 7

Code to do a 70/30 stratified split.

```{r listing-7}
# For reproducibility
set.seed(112)

# Split proportion
prop_train <- 0.7

# Partition data based on outcome
set_split <- split(df, df$replicate)

# Number of training samples
n_train <- lapply(set_split, function(set) round(nrow(set) * prop_train))

# Training sample, drawn inside each stratum
train_smpl <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_train
)

# Inclusion indicator
train_ind <- df$eid %in% unlist(train_smpl)

# Combined sets
strat2_diy <- list(
  "train" = df[train_ind, ],
  "test" = df[!train_ind, ]
)
```

::: {.callout-important}
The code has changed. It is not identical to what is in the tutorial
:::

#### Listing 8

Code to do a 70/15/15 stratified split in one pass.

**Note**: Switched from `props`to `prop`

```{r listing-8}
# For reproducibility
set.seed(112)

# Split proportions, of the whole data
prop <- c(train = 0.70, valid = 0.15, test = 0.15)

# Partition data based on outcome
set_split <- split(df, df$replicate)

# Set sizes per stratum, both counted from the whole data
n_train <- lapply(set_split, function(set) round(nrow(set) * prop[["train"]]))
n_valid <- lapply(set_split, function(set) round(nrow(set) * prop[["valid"]]))

# Training sample, drawn inside each stratum
train_smpl <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_train
)
train_eid <- unlist(train_smpl)
train_ind <- df$eid %in% train_eid

# Validation sample, drawn from each stratum's effects that are still available
valid_smpl <- mapply(
  function(set, n) sample(setdiff(set[, "eid"], train_eid), n),
  set_split,
  n_valid
)
valid_ind <- df$eid %in% unlist(valid_smpl)

# Combined sets
strat_diy <- list(
  "train" = df[train_ind, ],
  "valid" = df[valid_ind, ],
  "test" = df[!train_ind & !valid_ind, ]
)
```

::: {.callout-important}
The code has changed. It is not identical to what is in the tutorial. Need to rewrite the section.
:::

#### Listing 9

Code to do a 70/30 random and stratified split, with the pre-implemented
function of the adatutor package.

```{r listing-9}
# Simple random split (two way)
set.seed(112)
rand_split <- partition(df, prop = 0.7)

# Stratified split (three way)
set.seed(112)
strat_split <- partition(df, prop = c(0.7, 0.15, 0.15), strata = "replicate")

# Verification
mapply(identical, strat_diy, strat_split)
```

We recommend the adatutor's partition() function for the explorations and exercises in the tutorial. It validates inputs and is overall more robust than the DIY code above. For instance,

`altmejd` the listings and the function agree, but not everywhere:
`sample(x, k)` reads a length-one `x` as `1:x`, so a stratum or a remainder
holding a single row silently draws a value that is not in the data -- with a
four-member stratum, on more than half of all seeds. `partition()` samples
positions with `sample.int()` instead and cannot do this, which is the reason to
reach for a maintained implementation -- this one or `rsample`'s -- once you are
past the tutorial's own data.

#### Figure 1 [NEW]

Visualization of the leave-one-group-out cross-validation (LOGO-CV) process, with the replication projects as the groups. Reading the graph from left to right, the data are first partitioned into the Many Labs 1 (ML 1) project, the Replication Project Psychology (RPP), the Many Labs 3 (ML 3) project, the Experimental Economics Replication Project (EERP), and the Social Science Replication Project (SSRP). In Iteration 1, the ML 1 project is then used as the test set. The other data are used for model training. After fitting the model to the training data, its performance is evaluated on the hold-out or test data from the Many Labs 1 project. This is repeated for the other replication projects (Iterations 2--5), ultimately yielding a distribution of five performance estimates. It might be tempting to summarize the distribution with a point estimates, but keep in mind that the distribution always provides the richer picture of the validation process.

```{r figure-1}
tutplot_logoscheme(df$pid, levels = c("ml1", "rpp", "ml3", "eerp", "ssrp"))
```

#### Listing 10

Code to do a 70/30 random and stratified split, with the rsample package.

```{r listing-10}
# For reproducibility
set.seed(112)

# Simple random split
rand_split_pkg <- rsample::initial_split(df, prop = 0.7)
train_pkg <- rsample::training(rand_split_pkg)
test_pkg <- rsample::testing(rand_split_pkg)

# Stratified split
strat_split_pkg <- rsample::initial_split(df, prop = 0.7, strata = "replicate")
train_pkg <- rsample::training(strat_split_pkg)
test_pkg <- rsample::testing(strat_split_pkg)
```

For real work you will outgrow a single split -- cross-validation, bootstraps,
nested resampling -- and at that point the tool to reach for is the `rsample`
package, which the wider tidymodels ecosystem is built around. Its
`initial_split()` does what `partition()` does and takes the same `strata`
argument, so the idea carries over unchanged; what it adds is the two dozen
other resampling schemes you will want next.

TODO: Teach "two functions become one." A simple random split is a stratified split with one stratum — let the code demonstrate it rather than asserting it.
Retire `rsample` or teach how to use it?

Note: tidymodels consumes rsample

#### Listing [NEW]

Code to restore the stratified split from the package.

```{r restore-split}
# Load the data
data(altmejd_splits)

# The split every later module is written against
train <- altmejd_splits$train
valid <- altmejd_splits$valid
test <- altmejd_splits$test
```

::: {.callout-important title="Restore the split before going on"}

Module 1 showed several ways to split. Everything from Module 2 onward is built on the stratified split of Listings 7 and 8. The package ships it, so load it back and the stumps, weights and figures
come out as described. Setting a seed would not be enough on its own: R changed
`sample()` in version 3.6.0 and moved every seeded split in the wild with it,
which is why the sets themselves travel with the package rather than the recipe
for them.
:::

# Module 2: Classification Stumps

#### Listing 11

Code to retrieve the frequency tables for the outcome variable under the two
power conditions.

```{r listing-11}
# Condition for the split
split_cond <- train$power.o < 0.9305

# Studies with power.o < 0.9305
table(train$replicate[split_cond])

# Studies with power.o >= 0.9305
table(train$replicate[!split_cond])
```

#### Figure 2

Visualization of a classification tree of depth one or, equivalently, a classification stump. The stump was fitted to the training data using two reviewer metrics: the power of the original study and the number of observations. The middle rectangle represents the root node, and the two lower rectangles are the leaf nodes. The power value used to partition the root node is 0.9305. The left leaf node contains about 62% of the training data and the right leaf node contains about 38%. The label in each node ("success", "failure") represents the stump's prediction for each study effect in that node. For example, in the left leaf node, the stump predicts that studies with a power of less than 0.9305 will fail replication. According to the training data, 49 of the 66 study effects with a power of less than 0.9305 actually failed replication.

```{r figure-2}
# Fit classification stump
# Note: Two metrics only
cstump_int <- rpart::rpart(
  replicate ~ power.o + n.o,
  data = train[, voi_nms],
  maxdepth = 1,
  # keep to recycle below
  model = TRUE
)

# Visualize
tutplot_cstump(cstump_int, extra = 102)
```

#### Listing 12

Code to set the hyperparameters of a classification tree and fit the model to
the training data.

```{r listing-12}
# Fit classification stump
cstump <- rpart::rpart(
  replicate ~ power.o + n.o,
  data = train[, voi_nms],
  # Classification tree
  method = "class",
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  parms = list(split = "gini")
)
```

#### Listing 13

Code to visualize a fitted classification stump. `cstump` is the model behind Figure 2: the same two reviewer metrics, splitting on `power.o` at 0.9305 into leaves of 66 and 41. It is `cstump_int` with the arguments spelled out, so this listing reproduces that figure. The four-metric stump -- which splits on `p_value.o` at 0.0006 into leaves of 72 and 35 -- is `cstump_fit`, fitted in Listing 24.

```{r listing-13}
# Color-blind pallette
sty <- viridis_tree(cstump)

# Visualize classification stump
cstump |>
  rpart.plot::rpart.plot(
    extra = 100,
    digits = 4,
    box.col = sty$box,
    col = sty$text
  )
```

#### Figure 3

Visualization of the decision boundary of a classification stump. The predictor variable on the x-axis is the power of the original study effect. The predictor variable on the y-axis is the number of observations. So the y-axis does not display the outcome variable, but another study characteristic. However, since classification stumps are classification trees of depth one, only one study characteristic (i.e., power in this case) is used as a predictor to separate successfully replicated study effects from failed ones. The decision boundary is set at a power value of 0.9305 (vertical line). The model predicts that any point in the gray region is a successful replication, while anything in the blue region is considered a failure. Gray triangles represent replications of study effects that actually succeeded (i.e., according to the replication metric), while blue circles represent replications that actually failed. Thus, gray triangles in the blue region are misclassifications, and vice versa.

```{r figure-3}
# Visualize
# Note: the stump from Figure 2, now as the boundary it draws
tutplot_boundary(
  cstump_int,
  data = train,
  xlab = "Power (Original Study)",
  ylab = "Sample Size (Original Study)"
)
```

#### Figure 4

Visualization of the Gini impurity for a two-class classification problem. The x-axis is the proportion or probability of Class 1 (e.g., the proportion of successes in the empirical example). The y-axis is the measure of Gini impurity. The Gini impurity is highest when the classes are equally represented. This occurs when the proportion of Class 1 is 0.5, meaning there are as many instances of Class 1 (i.e., successes) as of Class 2 (i.e., failures). The more homogeneous a class becomes, the lower the Gini impurity. The Gini impurity is zero if there are only instances of one type (i.e., only successes or failures).

```{r figure-4}
# Visualize
tutplot_gini(x_seq = seq(0.01, 1, by = 0.01), lwd = 2)
```

#### Listing 14

Code to calculate means of neighboring units and find the position of the
relevant power value.

```{r listing-14}
# Sorted vector
power_srt <- sort(train$power.o)

# Adjacent means
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2

# Find position of value '0.9305'
which(round(adjmeans, 4) == 0.9305)

# Additional

adj_pat <- which(round(adjmeans, 4) == 0.9305)
const <- adjmeans[adj_pat]
N_total <- nrow(train) # 107
criterion_left <- train$power.o < const
N_left <- length(train$replicate[criterion_left]) # 66 total
left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1) # 17
left_failure <- N_left - left_success # 49 failures

# Gini impuritiy (left)
gini_left <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
gini_left # 0.382
criterion_right <- train$power.o >= const
N_right <- length(train$replicate[criterion_right]) # 41
right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1) # 31
right_failure <- N_right - right_success # 10

# Gini impuritiy (right)
gini_right <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
gini_right # 0.3688281
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) *
  gini_left +
  (N_right / N_total) * gini_right
round(gini_impurity, digits = 5) # 0.37724
# Sorted values
power_srt <- sort(train$power.o)
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2
gini_impurity <- vector(length = length(adjmeans))
for (i in seq_along(adjmeans)) {
  constant <- adjmeans[i]
  N_total <- nrow(train)
  # Left leaf
  criterion_left <- train$power.o < constant
  N_left <- length(train$replicate[criterion_left])
  left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- train$power.o >= constant
  N_right <- length(train$replicate[criterion_right])
  right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
  gini_d2 <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
  # Impurity
  gini_impurity[i] <- (N_left / N_total) *
    gini_d1 +
    (N_right / N_total) * gini_d2
}
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjmeans[which.min(gini_impurity)]
```

#### Listing 15

Code to make the assumption of equal weights explicit.

```{r listing-15}
# Number of rows
m <- nrow(train)

# Equal weight to each observation
D <- rep(1, m) / m

# Fit decision stump
cstump_eq <- rpart::rpart(
  replicate ~ power.o + n.o,
  data = train[, voi_nms],
  method = "class",
  maxdepth = 1,
  # Equal weights
  weights = D
)

# Visualize
sty <- viridis_tree(cstump)
cstump_eq |>
  rpart.plot::rpart.plot(
    extra = 100,
    digits = 4,
    box.col = sty$box,
    col = sty$text
  )
```


Visualization of a classification stump fitted with normalized equal weights. Note that we simplified the visual output shown here by using the extra argument of the `rpart.plot()` function with a value set to 100.

```{r}
# Visualize
tutplot_cstump(cstump, extra = 100)
```

#### Listing 16

Code to show the sensitivity to observation weights by using normalized random
weights from a uniform distribution.

Visualization of a classification stump fitted with normalized random weights. Note that we simplified the visual output shown here by using the `extra` argument of the `rpart.plot()` function with a value set to `100`.

:::{ .callout-note}
The tutorila says: "one of the p-values", but now the variable doesn't change at all; the threshold does. (power.o < 0.6358)
:::

```{r listing-16}
# For reproducibility
set.seed(110)

# Draws from standard uniform
D_raw <- runif(m)

# Normalize (sum to 1)
D <- D_raw / sum(D_raw)

# Create decision stump
# Note: the same two metrics as Listing 15, so only the weights differ
cstump_ast <- rpart::rpart(
  replicate ~ power.o + n.o,
  data = train[, voi_nms],
  method = "class",
  maxdepth = 1,
  weights = D
)

# Visualize
sty <- viridis_tree(cstump_ast)

cstump_ast |>
  rpart.plot::rpart.plot(
    extra = 100,
    digits = 4,
    box.col = sty$box,
    col = sty$text
  )
```

# Module 3: AdaBoost

#### Listing 17

Code to understand the givens in the AdaBoost algorithm.

```{r listing-17}
# Predictors and outcome (names)
pred_nms
out_nme

# Predictor matrix
x_train <- train[, pred_nms]

# Outcome vector
y_train <- train[, out_nme]

# Indices
i <- 1
m <- nrow(train)

# i-th row or element
x_train[i, ]
y_train[i]

# i-th to m-th row (i.e., all)
cbind(x_train, y_train)[i:m, ]
```

#### Listing 18

Code to generate initialization weights in the AdaBoost algorithm.

```{r listing-18}
# Initialization weights
D1 <- rep(1, m) / m
```

#### Listing 19

Code to fit a classification stump with the initialization weights.

```{r listing-19}
# Fit classification stump
h1 <- rpart::rpart(
  replicate ~ .,
  data = train[, voi_nms],
  method = "class",
  maxdepth = 1,
  weights = D1
)
```

#### Listing 20

Code to calculate the classification stump's error.

```{r listing-20}
# Retrodictions (training data)
y_retro <- predict(h1, newdata = train[, voi_nms], type = "class")

# Retrodictive error
epsilon <- sum(D1 * (y_retro != y_train))
```

#### Listing 21

Code to calculate the classification stump's performance.

```{r listing-21}
# Retrodicitve performance
perf <- sum(D1 * (y_retro == y_train))

# Retrodictive error
epsilon <- 1 - perf
```

#### Listing 22

Code to calculate the classification stump's weight assuming a learning rate
eta of 1.

```{r listing-22}
# Learning rate
eta <- 1

# Model weight
alpha1 <- 1 / 2 * log(perf / epsilon) * eta
```

#### Figure 6

Visualization of how the importance of a classification stump ($\alpha_t$) varies with its performance ($1-\epsilon_t$). Overall, the better the performance of a stump, the higher its weight. Put another way, the lower a stump's error ($\epsilon_t$), the higher its importance ($a_t$). The boosting hyperparameter ($\eta$) allows to further attenuate the weight of a classification stump. This could later help to improve the predictive performance of AdaBoost (see Module 4).

```{r figure-6}
# Learning rates
eta_int <- c(eta, 0.5, 0.1)

# Visualize
tutplot_importance(eta = eta_int, mark_perf = perf, mark_eta = eta)
```

#### Listing 23

Code to calculate the update factor for the initialization weights.

```{r listing-23}
# Indicator trick
chi1 <- (y_retro == y_train) * 1 + (y_retro != y_train) * -1

# Unnormalized weights
D2_raw <- D1 * exp(-alpha1 * chi1)

# Normalization constant
Z1 <- sum(D2_raw)

# Normalized weights
D2 <- D2_raw / Z1
```

#### Figure 7

Visualization of how the update factor ($e^{-\alpha\chi}$) varies depending on the classification result and the importance of the model ($\alpha_t$). If the classifier classifies an observation correctly, $\chi$ takes the value 1 and the weight or importance of the observation is reduced in the next iteration. In the case of a misclassification, $\chi$ takes the value -1 and the weight of the observation increases. The observation will be more important in the next iteration.

**TODO** I use $a$ as model weight not alpha! Either a or alpha needs fixing. Switch to alpha?

```{r figure-7}
# Model weights: yours in the middle, a heavier and a lighter one either
# side. Taken relative to alpha1, so they follow it if you change eta
alpha_int <- round(alpha1 * c(1.35, 1, 0.6), 3)
# Visualize
# Note: One curve per model weight
tutplot_updatefactor(
  # Model weights
  alpha = alpha_int,
  # Rise figure
  pch = c(20, 17, 15),
  lty = 3,
  lwd = 1,
  ylim = c(0.3, 2.2)
)
```

#### Figure 8

Visualization of the result of the weight updating process after Iteration 1 for the first 25 data points in the `altmejd` dataset. The gray initial weights $D_1$ increase or decrease depending on whether the first classification stump classified them correctly (green dot in the middle) or not (red dot in the middle). If the stump misclassifies an observation, its weight moves up and becomes larger, i.e., its importance increases. Correctly classified data points experience a weight decrease. They become less important in the next iteration. Visually, they move down and become smaller.

```{r figure-8}
# Visualize
# Note: The band marks the points the first stump missed
tutplot_weightone(d1 = D1, d2 = D2, chi = chi1)
```

Passing your own `D1`, `D2` and `chi1` is what ties this figure to the listings
above: change the learner or the weights and the picture follows. `chi1` is not
optional here, because the weights alone cannot say which points were missed --
a round grows the misclassified ones only while the stump beats chance. The call
returns those weights and the missed points invisibly, so assigning it gives you
the numbers as well as the plot.

#### Listing 24

Code to loop through the algorithmic steps.

::: {.callout-note}
The code can take several seconds to complete.
:::

```{r listing-24}
# Number of trees
T <- 1000

# Learning rate
eta <- 1

# Lists for models and weights
ada_diy <- vector("list", T)

# Initial observation weights
D <- rep(1, m) / m

# Loop through steps
for (t in seq_len(T)) {
  h <- rpart::rpart(
    replicate ~ .,
    data = train[, voi_nms],
    method = "class",
    maxdepth = 1,
    # Turn off, surrogates stand in for missing values, but we have none.
    # Leaving it on would put extra variables into each stump's
    # `variable.importance` and pull the scores below apart.
    maxsurrogate = 0,
    weights = D
  )
  y_retro <- predict(object = h, newdata = train[, pred_nms], type = "class")
  e <- sum(D * (y_retro != y_train))
  a <- 0.5 * log((1 - e) / e) * eta
  chi <- (y_retro == y_train) * 1 + (y_retro != y_train) * -1
  D_raw <- (D * exp(-a * chi))
  D <- D_raw / sum(D_raw)
  ada_diy[[t]] <- list("h" = h, "a" = a)
}
names(ada_diy) <- paste0("t", seq_len(T))

# Show first model weight
cat("Model weight 1:", ada_diy$t1$a)
```

#### The Package Solution [NEW]

Code to fit the same ensemble with the adatutor package.

```{r listing-24-pkg}
# Use stump as weak learner
cstump_fit <- rpart::rpart(
  replicate ~ .,
  data = train[, voi_nms],
  method = "class",
  # Tree hyperparameter
  maxdepth = 1,
  # Keep the training data
  model = TRUE,
  parms = list(split = "gini")
)

# Take the stump and boost it
ada_fit <- cstump_fit |> adaboost(n_iter = T, eta = eta)

# Verification
identical(sapply(ada_diy, "[[", "a"), sapply(ada_fit, "[[", "a"))
```

Note: Adatutor is a teaching package, hence forces you to be explicit and fails otherwise. `model = TRUE` in the `rpart()` function keeps the training data on the stump (outcome and predictors). Boosting only makes sense for the same (training) dataset used to fit the tree, so `adaboost()` has no slots to tinker with it. It is always the "specify once, resuse many" principle.

As both are weight by weight identical, we will use the package solution in the following to save some typing and support readers on their exploraiton of the algorithm with some additional sugar and safe guards we implemented in the package function.

Box:
Furthermore, the signature l. 802 also makes explicit what is often taken for granted: the weak learner is a (classification) stump. In fact the algorithm is written to accept any weak learner, with a performance slighly better than guessing (we see this in Eq. xxx). So the weak learner is iself a choice (a free hyperparameter if you will) and in fact you can also boost other models as well, such as deep trees or single-layer perceptrons (if you want to get wild). That said, `adaboost()` only accepts trees as this tutorial introduced them in Module 2. Indeed, stumps are a convenient choice, especially because of their simplicity and thus explainability which we will discuss next.

#### Listing [NEW]?

**TODO**: Prompt to the tidymodels package? and link the resources and say we designed the package so it mimics the structure for easier transition?

## NEW: Feature Importance
Note: We should describe the package function before module 4 and maybe move the feature importance before hyperparameter optimization. (before we touch performance)

TODO: Describe better.
So far, we restricted AdaBoost to work with decision stumps. From Module 2 we know, that stumps are trees of depth one that split (at most - recall: cp = 0.01) once on a variable. Recall also, that each stump carries a voting weight (`a`) based on its accuracy. The (relative) feature imporatance score we introduce below is to group stumps by the variable they split on, extract the (adaboost) weight, and sum their given weights. Why does it work? Well, AdaBoost sequentially targets the hardest-to-predict observations by re-weighting data points, so a variable that is frequently chosen, especially by high-weight stumps, is one that consistently helped correct the ensemble's "biggest mistakes". The total
accumulated weight is thus a direct measure of a feature's predictive contribution or value.

That is the whole idea with nothing borrowed: count up the voting weight of the
stumps that chose each variable.

Start with initializing a named vector of zeros (to keep the loop readable) growing it in Line -- `imp_diy[v] <- imp_diy[v] + ada_diy[[t]]$a` starting from an empty vector. If it does not split the addition step is skipped, so it silently produces nothing.

For stumps we can find the split directly: `frame` is the tree's node table.
It's first row is our root, so `frame$var[1]` is the variable the stump splits on 
If it does not split `rpart` writes `<leaf>`.

```{r importance-frame}
n_preds <- length(pred_nms)
# initialization: vector of zeros
voz <- numeric(n_preds)
imp_diy <- setNames(voz, pred_nms)
for (t in seq_along(ada_diy)) {
  # variable at the root (stumps)
  v <- as.character(ada_diy[[t]]$h$frame$var[1])
  # if it does not split, skip it
  if (v == "<leaf>") {
    next
  }
  imp_diy[v] <- imp_diy[v] + ada_diy[[t]]$a
}
# relative importance: normalize and scale
imp_diy <- imp_diy / sum(imp_diy) * 100
# highest importance first
sort(imp_diy, decreasing = TRUE)
```

Alternative loop with `rpart` internals.
Note: Keep only loop not the full code.

```{r importance-byhand}
# initialization: its own vector, so the loop above is not added to twice
imp_alt <- setNames(voz, pred_nms)
# walk ensemble, tree by tree
for (t in seq_along(ada_diy)) {
  # Shorcut: extract rpart's included measure
  vi <- ada_diy[[t]][["h"]][["variable.importance"]]
  # if it does not split, skip it
  if (is.null(vi)) {
    next
  }
  # stumps split (at most once), i.e. whole model weight goes to that variable
  imp_alt[names(vi)] <- imp_alt[names(vi)] + ada_diy[[t]]$a
}
# relative importance: normalize and scale
imp_alt <- imp_alt / sum(imp_alt) * 100

# Verification: same scores, different route to them
identical(imp_alt, imp_diy)
```

Note to self: Because in rpart cp is 0.01, by default, the trees do not need to split at all.
In fact, 81 of the 1000 stumps decided not to, which is why `next` exists.
Note to self: cp and pruning maybe as box?

Outlook? If we grow deeper trees they can split more than once, and the weight is shared between those variables. Specifically, the weight is shared in proportion to how much each split improved the overall fit. The difference comes down to a one-line difference:

version with rpart helper function, general case
```{r importance-general}
# Fit the weak learner: a depth-4 tree, not a stump
ctree <- rpart::rpart(
  replicate ~ .,
  data = train[, voi_nms],
  # Tree hyperparameter
  maxdepth = 4,
  model = TRUE
)

# Boost it
# Note: the boosting hyperparameters are all that is left here
ada_deep <- adaboost(ctree, n_iter = 10, eta = 1)

n_preds <- length(pred_nms)
voz <- numeric(n_preds)
imp_deep <- setNames(voz, pred_nms)
for (t in seq_along(ada_deep)) {
  vi <- ada_deep[[t]][["h"]][["variable.importance"]]
  if (is.null(vi)) {
    next
  }
  # Split the weight across the variables this tree used, in proportion to
  # each split's improvement
  share <- ada_deep[[t]]$a * vi / sum(vi)
  imp_deep[names(share)] <- imp_deep[names(share)] + share
}
imp_deep <- 100 * imp_deep / sum(imp_deep)
sort(imp_deep, decreasing = TRUE)
```

If we only have a stump, then `vi` holds a single number, so `vi / sum(vi)` is always 1. So we can the general version also for stumps.

In adaboost's gauge we mainly use the `rpart` helper function `variable.importance`. 
That is what `gauge()` does, so the two agree. 

```{r}
#| label: importance
important <- gauge(ada_fit)
head(important, 10)

# `gauge()` reads `variable.importance` too, so it and `imp_diy` agree. It
# returns its rows sorted, so index the hand-built vector by its names first.
imp <- setNames(important$importance, important$variable)
identical(imp, imp_diy[names(imp)])
```

We also provide a quick and easy way to visualize the results.

```{r}
#| label: importance-plot
#| fig-alt: "Horizontal bar chart of AdaBoost feature importance"
important |> plot(top_n = 15)
```

**Important note** The provided scores are relative (to eachother), so we believe it is more accurate to say, that they describe how the ensemble divided its attention rather than making a satement
on the net wort of a variable. They are also not "signed", so a variable may rand  

attention rather than what any one variable is worth on its own. They carry no
sign: a variable can rank highly because high values predict replication success
or because they predict failure, and this does not distinguish the two.
Correlated predictors split their share instead of each showing the full effect.
And the ranking belongs to *this* fit on *this* training data -- it moves with
the hyperparameters, since deeper trees spread weight over more variables by
construction.

For more see Molnar & Pargent.

TODO: Can we bootstrap the importance scores to get a sense of their stability? Or havent we done that already with the LOGO-CV?

# NEW: Module Retrodictions and Pretrodictions

#### Listing 25

TODO: Here we'll start the new module on prediction and retrodiction.

Code to generate ensemble retrodictions.

**TODO**: Switch from a to alpha!

```{r listing-25}
# Boosted classification stumps
h_fit <- lapply(ada_fit, "[[", "h")

# Model weights
alpha_fit <- sapply(ada_fit, "[[", "a")

# Retrodictions from stumps
y_retro_cstp <- sapply(
  h_fit,
  predict,
  newdata = train[, pred_nms],
  type = "vector"
)

# Transform values (-1,1)
y_retro_1m1 <- 2 * y_retro_cstp - 3

# Weighted vote across all stumps
retro_mgn_diy <- as.vector(alpha_fit %*% t(y_retro_1m1))

# Retrodictions from full model: the vote's sign
y_retro_diy <- sign(retro_mgn_diy)
```

#### Listing 26

Code to replace matrix multiplication with a loop.

```{r listing-26}
#for (m in seq_len(m)) {
#y_retro_diy[m] <- sign(sum(a * y_retro_cstp[m, ]))
#}
```

#### The Package Solution [NEW]

Code to generate the same retrodictions with the adatutor package.

```{r listing-25-pkg}
# Retrodictions from the full model
y_retro_cls <- predict(ada_fit, newdata = train[, pred_nms], type = "class")

# Verification
identical(y_retro_diy, y_retro_cls)
```

As both are identical we will use the package solution in the following to save typing and support readers on their exploraiton of the algorithm with some additional sugar and safe guards we implemented in the package function. For instance, read the message the call printed. `predict()` noticed that the data it was asked to score is the data the model was trained on and notes that it is returning retrodictions, not predictions.

**[NEW]**
One more thing worth knowing before the next listing. `type = "class"` applies
`sign()`, exactly as Listing 25 does, which throws away everything but the
direction of the vote. The quantity *before* the sign is available too:

```{r listing-25-margin}
# The weighted vote from Listing 25 has a name: the margin. The package
# returns it directly, before sign() discards its magnitude
# Note: a retrodiction, so `y_retro_` -- Listing 28 scores the test set
y_retro_mgn <- predict(ada_fit, newdata = train[, pred_nms], type = "margin")

# Verification
identical(retro_mgn_diy, y_retro_mgn)
```

Keep that in mind for Module 4: a hard label can only ever be right or wrong,
while the margin carries the ranking that measures like area under the receiving operating curve need.

#### Figure 9

Visualization of the decision boundary of 1000 adaptively boosted shallow classification trees. The number of trees matches the Listings 24 and 25, but the trees were fitted to the training data using only two reviewer metrics: the power of the original study and the number of observations. The x-axis is the power of the original study effect. The y-axis is the number of observations. Note that the y-axis is not the outcome variable, but another study characteristic. The model predicts that any point in the gray region is a successful replication, while anything in the blue region is considered a failure. Gray triangles represent replications of study effects that actually succeeded (i.e., according to the replication metric), while blue circles represent replications that actually failed. Thus, gray triangles in the blue region are misclassifications, and vice versa.

```{r figure-9}
ada_int <- cstump_int |> adaboost(n_iter = 1e3, eta = 1, verbose = FALSE)

# Visualize
# Note: the same call as Figure 3, an ensemble in place of the stump
tutplot_boundary(
  ada_int,
  data = train,
  xlab = "Power (Original Study)",
  ylab = "Sample Size (Original Study)"
)
```

#### Listing 27

Code to evaluate the retrodictive performance of the ensemble.

```{r listing-27}
# Confusion matrix
table(y_train, y_retro_cls) # Perfect retrodictions
```

#### Listing 28

Code to generate ensemble predictions.

```{r listing-28}
# Outcome variable (test set)
y_test <- test[, out_nme]

# Stump predictions (test set)
y_pred_cstp <- sapply(
  h_fit,
  predict,
  newdata = test[, pred_nms],
  type = "vector"
)
# Transform values (-1,1)
y_pred_cstp <- 2 * y_pred_cstp - 3

# Weighted vote across all stumps
pred_mgn_diy <- as.vector(alpha_fit %*% t(y_pred_cstp))

# Predictions from full model: the vote's sign
y_pred_diy <- sign(pred_mgn_diy)
```

**[NEW]**

```{r listing-28-pkg}
# The package's predictions, both types. Listing 31 scores these
y_pred_cls <- predict(ada_fit, newdata = test[, pred_nms], type = "class")
y_pred_mgn <- predict(ada_fit, newdata = test[, pred_nms], type = "margin")

# Verification
identical(y_pred_diy, y_pred_cls)
identical(pred_mgn_diy, y_pred_mgn)
```

#### Listing 29

Code to evaluate the predictive performance of the ensemble.

```{r listing-29}
# Confusion matrix
table(y_test, y_pred_cls) # Inperfect predictions
```

# Module 4: Leave-One-Group-Out Cross-Validation and Hyperparameter Optimization

## Two Wrapper Functions [REMOVE]

TODO: Wrapper function should be introduced earlier after the DIY version?

#### Listing 30

Code introducing the wrapper functions.

::: {.callout-note}
We reduced the number of trees to speed up runtime in the upcoming sections.
:::

```{r listing-30}
# Boosted stumps and weights
# Fit the weak learner
# Note: the tree hyperparameters live here, on the learner itself
# h <- rpart::rpart(
#  replicate ~ .,
#  data = train[, voi_nms],
#  maxdepth = 1,
#  model = TRUE
#)
#
## Boost it
#fit <- adaboost(h, n_iter = 100, eta = 1)
#
## Predictions on test set (class labels, for the confusion matrix)
#y_pred <- predict(fit, newdata = test[, pred_nms], type = "class")
#
## Predictions on test set (margins, for the AUC)
#ymargin <- predict(fit, newdata = test[, pred_nms], type = "margin")
```

::: {.callout-important}
## `type = "class"` gives balanced accuracy, not AUC

A two-valued score collapses the ROC curve to a single interior point, whose
area is exactly (sensitivity + specificity) / 2. Use `type = "margin"` for the
AUC; keep `type = "class"` for the confusion matrix.
:::

## Evaluating Model Performance

#### Listing 31

Code showing different measures to evaluate the predictive performance of the
ensemble.

```{r listing-31}
# Confusion matrix
conf_mat <- table(y_pred_cls, y_test)

# Accuracy
sum(diag(conf_mat)) / sum(conf_mat)

# Balanced Accuracy
auroc(y_test, y_pred_cls)

# Area under the curve (scored on the margin)
auroc(y_test, y_pred_mgn)

# Other metrics
assess(y_test, score = y_pred_mgn)

# Base rates
df$replicate_ind <- as.integer(df$replicate) - 1
baserate <- aggregate(replicate_ind ~ pid, df, mean)

# Average base rate
mean(baserate$replicate_ind)
```

**Note on margins**: A hard label can only ever be right or wrong. The margin says how confidently (unnormalized), and every measure that reads a ranking needs that. Keep it in mind for Module 4, where the cross-validation scores on the margin throughout.

::: {.callout-important}
## Switched to `PRROC` package
Cite and fix everywhere - especially in the tutorial.
:::

::: {.callout-note}
`auroc()` is a thin wrapper around `PRROC::roc.curve()` (Grau, Grosse &
Keilwagen). Specifically, `auprc()` wraps `pr.curve()`. The wrapper exists so the outcome can be passed in whatever form you have it, for instance as a factor, 0/1, -1/1 or logical. `roc.curve()` itself expects a 0/1 vector and returns a "wrong" number without complaining if you feed it a 1/2.

The precision-recall curve plots `ppv` (precision) against `sens` (recall) over every threshold, just as the ROC curve plots `sens` against 1 - `spec`. See `?assess` for the other names each measure goes by.
:::

## Leave-One-Group-Out Cross-Validation

#### Listing 32

Code to do a leave-one-group-out cross-validation.

::: {.callout-note}
This ran for approximately 2 minutes on our laptops.
:::

```{r listing-32}
# Partition data by project ID
df_split <- split(df, df$pid)

# List for AUC results
auc <- list()

runtime <- system.time({
  # LOGO-CV
  for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
    message("\nProject: ", paste0(rp), "\n")
    # Training and test set
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]
    # Fit model to data
    h <- rpart::rpart(
      replicate ~ .,
      data = train[, voi_nms],
      maxdepth = 1,
      model = TRUE
    )
    fit <- adaboost(h, n_iter = 1000, eta = 1)
    # Predictions (left-out project), on the margin so the AUC sees the ranking
    y_pred <- predict(fit, newdata = test[, pred_nms], type = "margin")
    # Performance evaluation
    auc[[rp]] <- auroc(test$replicate, y_pred)
  }
})

# Waiting time
message("\n Run time: ", runtime["elapsed"], " s\n")

# Results
c(unlist(auc))
```

**TODO**: Remove the mean?

#### Figure [New] 

```{r figure-xx}
# LOGO-CV: refit the same recipe with each project held out
logo_int <- h |>
  adaboost(n_iter = 1000, eta = 1, verbose = FALSE) |>
  logo_cv(data = df, group = "pid")

# Visualize (manuscript figure)
set.seed(112)
tutplot_logocv(logo_int, n_resample = 500)
```

`logo_cv()` uses the fit's settings, not its data: it refits the same recipe
once per project, holding that project out. The plot shows each project's
estimate and its spread, not a mean. The projects differ in size and share
most of their training data, so an average would hide both.

## Hyperparameter Optimization

#### Listing 33

Code to define a hypergrid.

```{r listing-33}
# Define hypergrid
hgrid <- expand.grid(
  # Tree-specific:
  # Tree depth
  "depth" = 1:3,
  # Boosting specific:
  # Number of trees
  "T" = c(10, 100, 1000),
  # Learning rate
  "eta" = c(1, 0.1),
  # Container for fold performance
  "eerp" = NA,
  "ml1" = NA,
  "ml3" = NA,
  "rpp" = NA,
  "ssrp" = NA,
  "auc.mean" = NA,
  "runtime" = NA
)
```

#### Listing 34

Code to do hyperparameter optimization using a leave-one-group-out
cross-validation strategy.

**TODO** Random grid search!

::: {.callout-note}
This ran for approximately 3 minutes on our laptops.
:::

```{r listing-34}
# Number of rows
hgrid_rows <- nrow(hgrid)

# Partition data by project ID
df_split <- split(df, df$pid)

# Hyperparameter optimization
for (i in seq_len(hgrid_rows)) {
  # Iteration time
  runtime <- system.time({
    # Leave-one-group-out CV
    for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
      # Progress report
      cat(
        "\rProgress: ",
        round(i / hgrid_rows * 100),
        "%",
        " | T: ",
        hgrid$T[i],
        " | eta: ",
        hgrid$eta[i],
        " | depth: ",
        hgrid$depth[i],
        " | model: ",
        rp
      )
      # Training and test set
      test <- df_split[[rp]]
      train <- df[!(df$pid %in% rp), ]
      # Fit model to data
      h <- rpart::rpart(
        replicate ~ .,
        data = train[, voi_nms],
        maxdepth = hgrid$depth[i],
        model = TRUE
      )
      fit <- adaboost(
        h,
        n_iter = hgrid$T[i],
        eta = hgrid$eta[i],
        verbose = FALSE
      )
      # Predictions (on the margin, so the AUC sees the ranking)
      y_pred <- predict(
        fit,
        newdata = test[, pred_nms],
        type = "margin",
        verbose = FALSE
      )
      # Performance evaluation
      hgrid[[rp]][i] <- auroc(test$replicate, y_pred)
    }
    # Average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i, 4:8]))
  })
  hgrid$runtime[i] <- runtime[["elapsed"]]
}

# Hypergrid sorted by average AUC
hgrid[order(-hgrid$auc.mean), ]
```

# NEW

Listing 34 searches 18 settings and keep the result in columns named after the
projects. The package ships the full search as `hypergrid`: 
960 settings, each scored on all five projects under 15 different measures.

`logo_cv()` returns that as an array rather than a table, because I think in 
setting x project x metric. So you can easily slice it without a reshape.

```{r hypergrid-array}
data(hypergrid)

# setting x project x metric
dim(hypergrid$estimates)
names(dimnames(hypergrid$estimates))

# `grid` says what each setting is. `cp`, the base learners' complexity
# parameter, is a column here like any other hyperparameter -- which is why 480
# combinations become 960 settings.
head(hypergrid$grid, 3)
hypergrid$projects
```

One measure across every setting and project: 960 x 5 matrix, and the grid
telly us which row is which:

```{r hypergrid-slice}
au <- hypergrid$estimates[,, "auroc"]
g <- hypergrid$grid

# One setting, sorted across projects
i <- which(g$depth == 1 & g$T == 1000 & g$eta == 0.15 & g$cp == 0.01)
sort(au[i, ], decreasing = TRUE)

# The best each project can be made to look, over the depth-1 stumps
d1 <- g$depth == 1 & g$cp == 0.01
apply(au[d1, ], 2, max)

# Rank the settings by their average over the projects
m <- rowMeans(au)
cbind(g, mean = round(m, 3))[order(-m), ][1:5, ]
```

An `NA` here means "undefined" rather than "failed": `ppv` has no value when a model predicts no positives at all, which a ten-tree ensemble does on the smaller projects. 

::: {.callout-important}
## The mean over projects ranks settings does not estimate performance

The last slice averages five folds that hold 10, 13, 18, 21 and 90 studies and
share most of their training data, so it hides both the imbalance and the
dependence. TODO
:::

## Selecting versus Evaluating

Listing 34 scores every setting in the hypergrid on all available (i.e., five) projects, and we took the row with the best average. I think that this number is not an estimate of how the chosen model will do on a new project, because the same five projects picked the winner and then graded it. To see how much that matters, we should do the selection where the project we want to score cannot influence it.

This is the job of nested leave-one-group-out cross-validation. The outer loop holds
out one project entirely and the inner loop repeats the whole grid search on the four that remain, choosing hyperparameters without ever seeing the held-out project. The
chosen setting is then refit on those four projects and scored on the fifth.
Repeat five times.

When I first met Nested LOGO-CV, I had a hard time to wrap my head around the fact, that it does not give me a(n overall) performance estimate for a particular hyperparameter setting. 
Instead, it puts a number (an estimate) on the procedure: "run the grid search, pick, then fit the winner." Every model built inside the nesting is discarded; none of them is the model you ship. You still fit one setting on all five projects, as in Listing 35. Nesting tells you what to expect from that
model, not what it scores.

completely separating the tuning process from the evaluation process

if you have a very large training validation and test set you can do the tripple, but if you have very vew data (as we comparably do) we might consider moving to nested cross-validation.

Take our data and split it into folds (the 5 groups in our case). Then we reserve one group, say Group 1, as test set (let's call this the outer test set). 
Then we take the remaining 4 (Group 2, 3, 4, 5) and split them again into 4 pieces: 3 pieces (Group 2, 3, 4) for the training set and one for the (inner) test set (Group 5). Then we (randomly, we do full Cartesian) search a hyperparameter setting from our test set and train the model with the inner 3 pieces (Group 2, 3, 4) and score it on the inner test set (Group 5) - and then again with a different hyperparameter setting. Then, we select the winning model with the best hyperparameter combi on Group 5 and pass it to the outer loop and train the model using the bigger training set with 4 pieces (Group 2, 3, 4, 5) and then evaluate it on the outer test set (Group 1). And then you repeat the whole procedure again with the next group
So instead of one model performance on on test set we have 5 (5 Groups) measures. What people than usually do is to take the mean performance across projects.
So we use all the data as a test set once

In small test set you can have the problem of bad luck and up (especially with small test sets) with a handy partition which is too easy to classify and (thus not representative of what would usually happen, i.e. performane overestimate).

TODO: Random grid search. Define 20 as a number in the random search

Two things follow. First, the five inner searches may each select a different
setting, and the setting that won in Listing 34 may be chosen by none of them.
That is expected rather than alarming: with five projects there is no stable
"best" setting to find, and seeing that instability directly is more informative
than the average that hides it. Second, the nested estimate is *conservative*
here, because each nested model saw four projects while the model you ship sees
five.

If nesting does not hand you a model, what does? The three listings have three
different jobs, and it is worth being explicit about which is which:

| job | listing |
|:----|:--------|
| Choose the hyperparameters, using all five projects | Listing 34 |
| Estimate what that choosing procedure is worth | Listing 35 |
| Fit the model you will actually use | Listing 36 |

So Listing 34 still picks the setting and Listing 36 still fits it on the full
dataset -- nesting changes neither. What it changes is the number you quote
alongside that model: not the winning average from Listing 34, which the same
five projects produced by selecting it, but the nested estimate, which no project
had a hand in choosing.

The settings the inner searches picked are worth looking at, but they are a
diagnostic rather than a shortlist. They have the same standing as the
coefficients fitted inside each cross-validation fold back in Module 1: nobody
averages those, and nobody asks which fold's coefficients to keep. They are
by-products of a procedure being measured, not candidates for shipping.

::: {.callout-tip}
## The depth belongs to the tree, not to the boosting

`adaboost()` takes a fitted learner, so there is only one place to set depth:
the `rpart()` call that builds it. Everything in `h$control` is carried into
every round -- `maxdepth`, `cp`, `minsplit`, `minbucket` -- so the tree you
looked at is the tree that gets boosted.

Two controls are overridden, and only those two. `xval` and `maxsurrogate` are
forced to zero: ten internal cross-validations per weak learner that nothing
reads cost about thirty per cent of the runtime, and neither setting changes
the tree that comes out. Everything you chose, you keep.

This is why AdaBoost has no `maxdepth` argument. The algorithm is defined over
*any* weak learner whose weighted error stays below one half; a tree is one
choice among many, and its settings are yours rather than the algorithm's.
:::

#### Listing 35

Code to do a nested leave-one-group-out cross-validation.

::: {.callout-note}
This ran for approximately 10 minutes on our laptops.
:::

```{r listing-35}
# Projects, in the order they take their turn
projects <- c("eerp", "ml1", "ml3", "rpp", "ssrp")

# Partition data by project ID
df_split <- split(df, df$pid)

# One honest score and one selected setting per project
nested_auc <- list()
selected <- list()

runtime <- system.time({
  # OUTER loop: evaluates
  for (rp in projects) {
    # Progress report
    message("\nOuter fold: ", rp, "\n")

    # Training and test set. The test project leaves the data entirely, so
    # nothing below can see it.
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]

    # INNER loop: selects. This is the same grid search as Listing 34, run on
    # the training projects only -- so we can hand it to logo_cv().
    inner <- logo_cv(
      replicate ~ power.o + effect_size.o + n.o + p_value.o,
      data = train,
      group = "pid",
      grid = hgrid[, c("depth", "T", "eta")]
    )

    # Rank the settings by their mean over the inner folds. Here the mean is a
    # ranking device, not a performance estimate.
    inner_auc <- summary(inner, metric = "auroc")
    ranked <- aggregate(estimate ~ setting, data = inner_auc, FUN = mean)

    # The setting the four training projects chose, on their own
    best <- hgrid[ranked$setting[which.max(ranked$estimate)], ]
    selected[[rp]] <- best[, c("depth", "T", "eta")]

    # Refit at that setting and score the untouched project
    h <- rpart::rpart(
      replicate ~ .,
      data = train[, voi_nms],
      maxdepth = best$depth,
      model = TRUE
    )
    fit <- adaboost(h, n_iter = best$T, eta = best$eta, verbose = FALSE)
    y_pred <- predict(
      fit,
      newdata = test[, pred_nms],
      type = "margin",
      verbose = FALSE
    )
    nested_auc[[rp]] <- auroc(test$replicate, y_pred)
  }
})

# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")

# The honest expectation, per project
unlist(nested_auc)

# What each outer fold selected, and how often they disagree
do.call(rbind, selected)
```

Only the outer loop is new. The inner one is the leave-one-group-out search we
already ran in Listing 34, so we hand it to `logo_cv()` rather than write it out
again -- the single change is that it now sees `train` instead of `df`.

The whole procedure is also one call, which is what we use from here on:

```{r listing-35-oneline}
nested <- logo_cv(
  replicate ~ power.o + effect_size.o + n.o + p_value.o,
  data = df,
  group = "pid",
  grid = hgrid[, c("depth", "T", "eta")],
  nested = TRUE,
  criterion = "auroc"
)
nested
```

## Finalizing an Ensemble

#### Listing 36

Code to fit the final model to the data.

```{r listing-36}
# Fit optimized model
h_final <- rpart::rpart(
  replicate ~ .,
  data = df[, voi_nms],
  maxdepth = 1,
  model = TRUE
)
fit_final <- adaboost(h_final, n_iter = 10, eta = 1)
```

# Figure map

Which chunk above produces which figure in the paper. Established by matching
the actual splits against the captions, not by reading the file names --
several of those mislead, and one was actively wrong until it was fixed.

| label | file | float | produced by | geometry |
|:--|:--|:--|:--|:--|
| `fig:lpocv` | `ext/lpocv` | `figure*` | Figure 1, `tutplot_logoscheme(df$pid)` | full |
| `fig:cstump` | `ext/cstump` | `figure` | Figure 2, `rpart.plot(h)`, **two features** | column |
| `fig:dboundstump` | `ext/dboundstump` | `figure*` | Figure 3, `tutplot_boundary(cstump_int)` | full |
| `fig:gini` | `ext/gini` | `figure` | Figure 4, the Gini curve | column |
| `fig:cstumpwght` | `ext/cstumpwght` | `figure` | Listing 16, `cstump_ast`, `extra = 100` | column |
| `fig:importance` | `ext/importance` | `figure` | Figure 6, the eta curves | column |
| `fig:updatefactor` | `ext/updatefactor` | `figure` | Figure 7 | column |
| `fig:weightone` | `ext/weightone` | `figure` | Figure 8 | column |
| `fig:dboundada` | `ext/dboundada` | `figure*` | Figure 9, `tutplot_boundary(ada_int)` | full |
| `fig:logocv` | `ext/logocv` | `figure` | new, `tutplot_logocv(res)` | column |

Every file now comes from R: six single-column and three `figure*`. The LOGO-CV
diagram used to be TikZ drawn in the manuscript, which is why its LaTeX source
still sits in this vignette -- kept for reference, not executed.

## Four things found while establishing that map

1. **`cstump` is a tree, not a boundary.** This vignette used to export the
   boundary plot under that name, and had already overwritten the tree in the
   manuscript's `ext/`. Fixed: the boundary is `dboundstump.pdf` now, and
   the ensemble's is `dboundada.pdf`.
2. **Module 2's narrative is the two-feature model.** `fig:cstump`'s caption
   says "two reviewer metrics: the power of the original study and the number of
   observations", and Listing 11 and Listing 14's arithmetic both describe the
   `power.o < 0.9305` split with 66/41 leaves. Listing 12 fits all four features
   and splits on `p_value.o` with 72/35. Only Listing 12 disagrees with the
   story. Open.
3. **`cstumpwght` does not reproduce.** Its caption is Listing 16 exactly --
   normalized random weights, `extra = 100` -- but the shipped file shows a
   `p_value.o` split with 72/35 leaves while Listing 16 gives `power.o` with
   66/41. Open.
4. **The canonical `train` is Listing 7's, and the RNG order used to matter.**
   Every hardcoded number in Listing 14 reproduces under it -- 66, 49, 41, .382,
   .3688281, .37724, and the search finds .9305. This used to be fragile:
   Listing 9 drew a simple random split first, which advanced the generator, so
   *its* stratified `train` split at .9261 with 70/37 leaves and none of Listing
   14's numbers held. Listing 9 now re-seeds before the stratified call, and its
   `train` is `identical()` to `altmejd_splits$train` -- 107 rows, `power.o` at
   .9305, leaves 66/41. The callout after Figure 2 still earns its place, since
   Module 1 offers several splits and only this one is canonical.
