---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

- Why Boosting?

- Why AdaBoost?

- What is Boosting 

- What means adaptive? (see, e.g. Kunapuli, 4.2.1)

- Comparison to prediction markets: lay person (boosting) vs experts (random forest)


# Approach (hands dirty)

- Describe the approach taken
  - Why implementing it yourself? 
    - Understanding
    - The role of hyperparameters!
    - Hyperparameters and tuning! 

- Note Florian Pargent et al.'s paper (supplement not complement)
  - For basic terminology (e.g., bias-variance trade-off, classification, AUC)
  - For the right mindset (e.g., predictions in psychology)
  - and other things

# Software (maybe together with Approach)

- Describe the software used

```{r setup}
library(adatutoR)
```

- Describe the data set, i.e. data sets used

```{r}
# Load the data
data("altmext723")
```

- TODO: Document the data set in R so that it is accessible with `?altmext723`

# Module 1: Splitting 

# Module 2: A Primer on Classification Trees 

# Module 3: Adaptive Boosting 

# Module 4: Leave-Subject-Out Cross-Validation

# Module 5: Hyperparameter Optimization using LPO-CV

# Module 6: Stacking (?) 

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1. 

