---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Software (maybe together with Approach)

- Describe the software used

```{r setup}
library(adatutoR)
```

- Describe the data set, i.e. data sets used

```{r}
# Load the data
data("altmext")
```

- TODO: Document the data set in R so that it is accessible with `?altmext`

# Introduction 

Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "p_value.o", 
         "replicated")

# Name of outcome variable (classification task)
outcome <- "replicated"

# Names of predictors 
predictors <- c("power.o", "effect_size.o", "n.o", "p_value.o")

# Names of outcomes and predictors
outpred <- union(outcome, predictors)

# Data frame
df <- altmext[,voi]
```

# Module 1: Splitting 

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

### Simple random split 

Simple random split using the full data set

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.7
# Number of effects 
n_effect <- nrow(df)
# Number of training samples
n_train <- round(n_effect * prop)
# Training index
train_idx <- sample(seq_len(n_effect), n_train)
# Training and testing set 
train <- df[train_idx, ]
testvalid  <- df[-train_idx, ]
```
Simple wrapper function

```{r}
tilde <- random_split(df, prop = 0.7, seed = 112)
train_tilde <- tilde$set1
testvalid_tilde <- tilde$set2
```

Check wrapper with unique effect identifier 

```{r}
all.equal(train$eid, train_tilde$eid)
all.equal(testvalid$eid, testvalid_tilde$eid)
```

Simple random split using the test set 

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.5
# Number of effects 
n_effect <- nrow(testvalid)
# Number of training samples
n_valid <- round(n_effect * prop)
# Training index
valid_idx <- sample(seq_len(n_effect), n_valid)
# Training and testing set 
valid <- testvalid[valid_idx, ]
test  <- testvalid[-valid_idx, ] 
```

With wrapper function

```{r}
tilde <- random_split(testvalid, prop = 0.5, seed = 112)
valid_tilde  <- tilde$set1
test_tilde <- tilde$set2
```

Check wrapper with unique effect identifier 

```{r}
all.equal(valid$eid, valid_tilde$eid)
all.equal(test$eid, test_tilde$eid)
```

### Stratified sampling & imbalances

Show the imbalances in the outcome variable

```{r}
tbl_orig <- table(df$replicated)
proportions(tbl_orig)
```

Show the imbalances in the training and testing set

```{r}
tbl_train  <- table(train$replicated)
proportions(tbl_train)
tbl_test <- table(test$replicated)
proportions(tbl_test)
```
Stratified split using the full data set

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.7
# Partition the data based on the outcome variable
set_split <- split(df, df$replicated)
# Number of training samples (stratified)
n_train <- lapply(set_split, function(set) round(nrow(set) * prop))
# Training index (stratified)
strat_idx <- mapply(function(set, n) sample(set[,"eid"], n), 
                    set_split, n_train)
# Training index 
train_idx <- df$eid %in% unlist(strat_idx) 
# Training and testing/validation set 
train <- df[train_idx, ]
testvalid  <- df[!train_idx, ]
```

Check the results

```{r}
proportions(table(train$replicated))
proportions(table(testvalid$replicated))
```
Simple wrapper function

```{r}
tilde <- stratified_split(df, group = "replicated", id = "eid", prop = 0.7,  
                          seed = 112)
train_tilde <- tilde$set1
testvalid_tilde <- tilde$set2
```

Check wrapper on unqiue effect identifier

```{r}
all.equal(train$eid, train_tilde$eid)
all.equal(testvalid$eid, testvalid_tilde$eid)
```

Stratified split using the test set

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.5
# Partition the data based on the outcome variable
set_split <- split(testvalid, testvalid$replicated)
# Number of training samples (stratified)
n_valid <- lapply(set_split, function(set) round(nrow(set) * prop))
# Validation index (stratified)
strat_idx <- mapply(function(set, n) sample(set[,"eid"], n), 
                    set_split, n_valid)
valid_idx <- testvalid$eid %in% unlist(strat_idx) 
# Training and testing set 
valid <- testvalid[valid_idx, ]
test  <- testvalid[!valid_idx, ]
```

Check the results

```{r}
proportions(table(test$replicated))
proportions(table(valid$replicated))
```
Simple wrapper function

```{r}
tilde <- stratified_split(testvalid, group = "replicated", id = "eid", 
                          prop = 0.5, seed = 112)
valid_tilde <- tilde$set1
test_tilde <- tilde$set2
```

Check wrapper on unique effect identifier 

```{r}
all.equal(valid$eid, valid_tilde$eid)
all.equal(test$eid, test_tilde$eid)
```

Pre-implemented solution using the rsample package

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```

# Module 2: A Primer on Classification Trees 

## Classification problems

## Classification trees

Figure 2.1

```{r, dev = "png"}
# Extract only two predictors 
# Note: Reduces run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <-  train[,union(subfeat, outcome)]
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
# Create a decision tree 
h <- rpart::rpart(formula = replicated ~ ., data = subdf,  method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, extra = 102, digits = 5, box.palette="BuGy")
```

#### Classification stumps

Figure 2.2

```{r}
# Outcome variable
y_train <- train[, outcome]
# Number of classes
k <- length(unique(subdf$replicated))
# Range(s) as list
rng <- lapply(subdf[,subfeat], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Stump")
points(grid, col = grid_colors, pch = ".")
points(subdf[,subfeat], col = symbol_colors, pch = as.integer(y_train) + 15)
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)
```

Code Chunk 1

```{r Code Chunk 1}
# Define the condition for the split
splitcond <- train$power.o < 0.92866
# Studies with df$power.o < 0.92866 
table(train$replicated[splitcond])
# Studies with df$power.o >= 0.92866 
table(train$replicated[!splitcond])
```

## Fit and visuaize a classification stump

Code Chunk 2.2

```{r}
# Hyperparameter setup for a classification stump
hyptree <- rpart::rpart.control(
  maxdepth = 1, # Maximum depth
  split = "gini") # Gini impurity (default)
# Fit a classification stump
cstump <- rpart::rpart(formula = replicated ~ ., 
                       data = train[,outpred],
                       method = "class", control = hyptree)
```

Code Chunk 2.3

```{r, eval=FALSE}
# Visualize a classification stump
rpart.plot::rpart.plot(cstump, extra = 102, digits = 5, box.palette="BuGy")
```

### Gini impurity

Figure 2.3

```{r}
# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01)  # Values for fraction of class k
y <- 1 - (x^2) - (1-x)^2      # Gini Impurity calculation

# Create the plot
plot(x, y, type = "l", lwd = 2, # Specify line plot
     xlab = "Proportion of Successes",
     ylab = "Impurity" )

mtext("Two Class Classification Problem")
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1))
```

#### Why `power.o > 0.92866`?

Code Chunk 2.4

```{r}
power_sorted <- sort(train$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
```

Additional material. Calculations in the text. Didiactic not performant.

```{r}
# Example above:
constant <- adjacent_means[69] # 0.9286641
N_total <- nrow(train) # N_total: 107

# Left leaf (power.o < 0.9286641: yes)
criterion_left <- train$power.o < constant
N_left <- length(train$replicated[criterion_left]) # 69 total
left_success <- sum(as.numeric(train$replicated[criterion_left]) - 1) # 19 successes
left_failure <- N_left - left_success # 50 failures

# Right leaf (power.o < 0.9286641: no)
criterion_right <- train$power.o >= constant
N_right <- length(train$replicated[criterion_right]) # 38 
right_success <- sum(as.numeric(train$replicated[criterion_right]) - 1) # 29 
right_failure <- N_right - right_success # failures
# Individual Gini impurities
gini_left <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
gini_right <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) * gini_left + (N_right / N_total) * gini_right
round(gini_impurity, digits = 5)
```

Control calculations. Quick and dirty.

```{r}
power_sorted <- sort(train$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
gini_impurity <- vector(length = length(adjacent_means))
  for(i in 1:length(adjacent_means) ) {
  constant <- adjacent_means[i]
  N_total <- length(y_train)
  # Left leaf
  criterion_left <- train$power.o < constant
  N_left <- length(train$replicated[criterion_left])
  left_success <- sum(as.numeric(train$replicated[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- train$power.o >= constant
  N_right <- length(train$replicated[criterion_right])
  right_success <- sum(as.numeric(train$replicated[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
  gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
  # Impurity
  gini_impurity[i] <- (N_left / N_total) * gini_d1 + 
    (N_right / N_total) * gini_d2
  }
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjacent_means[which.min(gini_impurity)]
```

#### Sensitivity to observation weights

Normalized initialization weights. Should not change anything to before.

```{r}
N_sample <- nrow(train)
# Equal weight to each observation
D <- rep(1, N_sample)/N_sample
# Create a decision stump
h <- rpart::rpart(formula = replicated ~ ., 
                  data = train[, outpred], 
                  method = "class", weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, digits = 5, box.palette="BuGy")
```

Normalized random observation weights for demonstration. Changes result.

```{r}
# Reproducibility
set.seed(112)
# Random draws from Uniform(0,1)
D_unorm <- runif(107)
# Normalize (sum: 1)
D <- D_unorm/sum(D_unorm) 
# Create a decision stump
cstump_ast <- rpart::rpart(formula = replicated ~ ., 
                  data = train[,outpred], 
                  method = "class", weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(cstump_ast, extra = 100, digits = 5, box.palette="BuGy")
```

# Module 3: Adaptive Boosting 

Setup.

```{r}
n_trees <- 100
# Extract the number of rows
n_samples <- nrow(train[,outpred])
# Pre-allocate the empty ensemble
H <- vector("list", n_trees)
```

## Adaptive boosting (1 iteration)

```{r}
# Outcome vectors 
y_train <- train[, outcome]
(y_test <- test[, outcome])
```

Initialization weights are normalized equal weights.

```{r}
m <- n_samples
one <- rep(1, n_samples)
D_t0 <- one/m
```

Train a single weak learner ($h_t$) with the initialization weights and the
given tree hyperparameter setup.

```{r}
h <- rpart::rpart(formula = replicated ~ ., data = train[,outpred] ,
method = "class", weights = D_t0, control = hyptree)
```

Make predictions with the weak learner (h_t).

```{r}
y_class <- predict(object = h, newdata = train[, outpred], type = "class")
```

Normalization constant.

```{r}
Z_t0 <- sum(D_t0)
```

Weighted accuracy for the training error.

```{r}
wacc <- sum(D * (y_train == y_class)) / Z_t0
```

Compute the training error (e_t) of the weak learner (h_t). This is simply one
minus the weighted accuracy, given weights D

```{r}
e <- 1 - wacc
```

Compute the the weight (a_t) of the weak learner (h_t).
Reduce the learner weight by a factor eta (learning rate).

Note: First AdaBoost hyperparameter.

```{r}
eta <- 1
```

Importance -- or weight -- of a classification tree. 

```{r}
a_t0 <- 1/2 * log( (1 - e)  / e) * eta
```

The better the model performs (that is, the lower its error) the more weight it
gets in the overall ensemble - independent of the learning rate $\eta$.

```{r}
curve((1/2*log(x/(1-x)) * 1), from = 0.5, to = 1, lwd = 2, 
      ylab = expression("Importance "*(a[t])), 
      xlab = expression("Performance "*(1-epsilon[t])), col = "ivory4")
curve((1/2*log(x/(1-x)) * 0.5), from = 0.5, to = 1, lwd = 2, col = "black", 
      add = TRUE)
curve((1/2*log(x/(1-x)) * 0.1), from = 0.5, to = 1, lwd = 2, col = "steelblue", 
      add = TRUE)
grid()
legend("topleft", legend = c(expression(eta*" = 0.1"), 
                              expression(eta*" = 0.5"), 
                              expression(eta*" = 1")),
       col = c("steelblue", "black", "ivory4"),
       lty = c(1, 1, 1), lwd = 2)
```

Indicator function that gives a one for correct classifications (prediction is
equal to observed value in the training set) and a minus one for a
misclassification (prediction is unequal to observed value in the training set).

```{r}
chi <- (y_train == y_class) * 1 + (y_train != y_class) * -1
```

Update the sample weights

- Increase for misclassified examples (y == y_pred)
- Decrease for correctly classified examples: (y != y_pred)

```{r}
D_t1unorm <- D_t0 * exp(-a_t0 * chi)
```

Define the normalization constant.

```{r}
Z <- sum(D_t1unorm)
```

Normalize the weights (maintain the probabilistic interpretation).

```{r}
D_t1 <- D_t1unorm/Z
```

Weight updating process. Answers, why correctly classified training data points
are less reduced than the falsely classified are increased.

```{r}
curve(exp(0.5*-x), from = -1, to = 1, ylim = c(0,3), lwd = 2, 
      xlab = "Classification", 
      ylab = expression("Weight update factor "*(exp(D[t]))),  xaxt = "n") 
curve(exp(1*-x), from = -1, to = 1, ylim = c(0,2), col = "ivory4", lwd = 2, 
      add = TRUE) 
curve(exp(0.1*-x), from = -1, to = 1, ylim = c(0,2), col = "steelblue", 
      lwd = 2, add = TRUE) 
abline(h = 1, lty=2)
legend("topright", legend = c(expression(a[t]*" = 0.1"), 
                              expression(a[t]*" = 0.5"), 
                              expression(a[t]*" = 1")),
       col = c("steelblue", "black", "ivory4"),
       lty = c(1, 1, 1), lwd = 2)
axis(1,at = c(-1,1))
grid()
```
$a = f(e, \eta)$ The more weight $a$ we assign to an individual learner the
stronger the effect on the observation weights (increase/decrease).
Misclassifications are always more penalized (increase in observation weight)
than correct classifications (decrease in observation weights).

Visualize the result of the weight updating process.

```{r}
n_25 <- 25
scaling <- 175
plot(c(1, n_25), c(0, D_t0[1] + 0.015) , type = "n", xaxt = "n",
     xlab = "Data point", ylab = "Weight")
axis(1, at = 1:n_25)
for(i in 1:n_25) {
  lines(c(i,i),c(D_t0[i], D_t1[i]), lty = 2)
}
points(D_t0[1:n_25], pch = 20, col = "ivory4", cex = D_t0[1] * scaling)
points(D_t1[1:n_25], pch = 20, col = "steelblue", cex = D_t1[1:n_25] * scaling) 
legend("topright", legend = c(expression(D[t[0]]), expression(D[t[1]])), 
       col = c("ivory4", "steelblue"),
       pch = c(20, 20))
```


Output.

```{r}
# Learner
h
# Learner weight
a_t0
# Observation weight
D_t1
```

Repeat.

## Adaptive boosting

Do the loop and explain the second hyperparameter (n_tree)

TODO: Work example.

```{r}
n_tree <- 1e2
```

For all $t$ trees or more precisely stumps.

```{r}
for (tree in 1:n_trees) {
  h <- rpart::rpart(formula = replicated ~ ., data = train[,outpred] ,
  method = "class", weights = D, control = hyptree)
  y_pred <- predict(object = h, newdata = test[, outpred], type = "class")
  wacc <- sum(D * (y == y_pred))/sum(D)
  e <- 1 - wacc
  a <- 0.5 * log( (1 - e)  / e) * eta
  chi <- (y == y_pred) * 1 + (y != y_pred) * -1
  D <- D * exp(-a * chi)
  D <- D/sum(D)
  # Ensemble 
  H[[tree]] <- list("learner" = h, "learner_weight" = a,
                    "observation_weight" = D)
  # sign in testing
}

# sign in testing
```

```{r}
# Feature matrices
X_train <- train[, predictors] 
X_test <- test[, predictors] 
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 
```


### Prediction

```{r}
test_adaboost
```

Make predictions with the trained model and add them to the testing set.

```{r}
y_class <- test_adaboost(X_test, ensemble)$class
test$y_pred <- factor( (y_class + 3) / 2, labels = c("no", "yes") )
```

## Evaluation

```{r}
ROC <- AUC::roc(test$y_pred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```




# Module 4: Leave-Subject-Out Cross-Validation

# Module 5: Hyperparameter Optimization using LPO-CV

# Module 6: Stacking (?) 

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1. 

