---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Software (maybe together with Approach)

- Describe the software used

```{r setup}
library(adatutoR)
```

- Describe the data set, i.e. data sets used

```{r}
# Load the data
data("altmext")
```

- TODO: Document the data set in R so that it is accessible with `?altmext`

# Introduction 

Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "p_value.o", 
         "replicated")

# Name of outcome variable (classification task)
outcome <- "replicated"

# Names of predictors 
predictors <- c("power.o", "effect_size.o", "n.o", "p_value.o")

# Names of outcomes and predictors
outpred <- union(outcome, predictors)

# Data frame
df <- altmext[,voi]
```

# Module 1: Splitting 

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

### Simple random split 

Simple random split using the full data set

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.7
# Number of effects 
n_effect <- nrow(df)
# Number of training samples
n_train <- round(n_effect * prop)
# Training index
train_idx <- sample(seq_len(n_effect), n_train)
# Training and testing set 
train <- df[train_idx, ]
testvalid  <- df[-train_idx, ]
```
Simple wrapper function

```{r}
tilde <- random_split(df, prop = 0.7, seed = 112)
train_tilde <- tilde$set1
testvalid_tilde <- tilde$set2
```

Check wrapper with unique effect identifier 

```{r}
all.equal(train$eid, train_tilde$eid)
all.equal(testvalid$eid, testvalid_tilde$eid)
```

Simple random split using the test set 

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.5
# Number of effects 
n_effect <- nrow(testvalid)
# Number of training samples
n_valid <- round(n_effect * prop)
# Training index
valid_idx <- sample(seq_len(n_effect), n_valid)
# Training and testing set 
valid <- testvalid[valid_idx, ]
test  <- testvalid[-valid_idx, ] 
```

With wrapper function

```{r}
tilde <- random_split(testvalid, prop = 0.5, seed = 112)
valid_tilde  <- tilde$set1
test_tilde <- tilde$set2
```

Check wrapper with unique effect identifier 

```{r}
all.equal(valid$eid, valid_tilde$eid)
all.equal(test$eid, test_tilde$eid)
```

### Stratified sampling & imbalances

Show the imbalances in the outcome variable

```{r}
tbl_orig <- table(df$replicated)
proportions(tbl_orig)
```

Show the imbalances in the training and testing set

```{r}
tbl_train  <- table(train$replicated)
proportions(tbl_train)
tbl_test <- table(test$replicated)
proportions(tbl_test)
```
Stratified split using the full data set

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.7
# Partition the data based on the outcome variable
set_split <- split(df, df$replicated)
# Number of training samples (stratified)
n_train <- lapply(set_split, function(set) round(nrow(set) * prop))
# Training index (stratified)
strat_idx <- mapply(function(set, n) sample(set[,"eid"], n), 
                    set_split, n_train)
# Training index 
train_idx <- df$eid %in% unlist(strat_idx) 
# Training and testing/validation set 
train <- df[train_idx, ]
testvalid  <- df[!train_idx, ]
```

Check the results

```{r}
proportions(table(train$replicated))
proportions(table(testvalid$replicated))
```
Simple wrapper function

```{r}
tilde <- stratified_split(df, group = "replicated", id = "eid", prop = 0.7,  
                          seed = 112)
train_tilde <- tilde$set1
testvalid_tilde <- tilde$set2
```

Check wrapper on unqiue effect identifier

```{r}
all.equal(train$eid, train_tilde$eid)
all.equal(testvalid$eid, testvalid_tilde$eid)
```

Stratified split using the test set

```{r}
# Reproducibility
set.seed(112)
# Split proportion
prop <- 0.5
# Partition the data based on the outcome variable
set_split <- split(testvalid, testvalid$replicated)
# Number of training samples (stratified)
n_valid <- lapply(set_split, function(set) round(nrow(set) * prop))
# Validation index (stratified)
strat_idx <- mapply(function(set, n) sample(set[,"eid"], n), 
                    set_split, n_valid)
valid_idx <- testvalid$eid %in% unlist(strat_idx) 
# Training and testing set 
valid <- testvalid[valid_idx, ]
test  <- testvalid[!valid_idx, ]
```

Check the results

```{r}
proportions(table(test$replicated))
proportions(table(valid$replicated))
```
Simple wrapper function

```{r}
tilde <- stratified_split(testvalid, group = "replicated", id = "eid", 
                          prop = 0.5, seed = 112)
valid_tilde <- tilde$set1
test_tilde <- tilde$set2
```

Check wrapper on unique effect identifier 

```{r}
all.equal(valid$eid, valid_tilde$eid)
all.equal(test$eid, test_tilde$eid)
```

Pre-implemented solution using the `rsample` package

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```

# Module 2: A Primer on Classification Trees 

## Classification problems

## Classification trees

Figure 2.1

```{r, dev = "png"}
# Extract only two predictors 
# Note: Reduces run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <-  train[,union(subfeat, outcome)]
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
# Create a decision tree 
h <- rpart::rpart(formula = replicated ~ ., data = subdf,  method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, extra = 102, digits = 5, box.palette="BuGy")
```

#### Classification stumps

Figure 2.2

```{r}
# Outcome variable
y_train <- train[, outcome]
# Number of classes
k <- length(unique(subdf$replicated))
# Range(s) as list
rng <- lapply(subdf[,subfeat], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o")
points(grid, col = grid_colors, pch = ".")
points(subdf[,subfeat], col = symbol_colors, pch = as.integer(y_train) + 15)
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)
```

Code Chunk 1

```{r Code Chunk 1}
# Define the condition for the split
splitcond <- train$power.o < 0.92866
# Studies with df$power.o < 0.92866 
table(train$replicated[splitcond])
# Studies with df$power.o >= 0.92866 
table(train$replicated[!splitcond])
```

## Fit and visuaize a classification stump

Code Chunk 2.2

```{r}
# Hyperparameter setup for a classification stump
hyptree <- rpart::rpart.control(
  maxdepth = 1, # Maximum depth
  split = "gini") # Gini impurity (default)
# Fit a classification stump
cstump <- rpart::rpart(formula = replicated ~ ., 
                       data = train[,outpred],
                       method = "class", control = hyptree)
```

Code Chunk 2.3

```{r, eval=FALSE}
# Visualize a classification stump
rpart.plot::rpart.plot(cstump, extra = 102, digits = 5, box.palette="BuGy")
```

### Gini impurity

Figure 2.3

```{r}
# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01)  # Values for fraction of class k
y <- 1 - (x^2) - (1-x)^2      # Gini Impurity calculation

# Create the plot
plot(x, y, type = "l", lwd = 2, # Specify line plot
     xlab = "Proportion of Successes",
     ylab = "Impurity" )

mtext("Two Class Classification Problem")
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1))
```

#### Why `power.o > 0.92866`?

Code Chunk 2.4

```{r}
power_sorted <- sort(train$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
```

Additional material. Calculations in the text. Didiactic not performant.

```{r}
# Example above:
constant <- adjacent_means[69] # 0.9286641
N_total <- nrow(train) # N_total: 107

# Left leaf (power.o < 0.9286641: yes)
criterion_left <- train$power.o < constant
N_left <- length(train$replicated[criterion_left]) # 69 total
left_success <- sum(as.numeric(train$replicated[criterion_left]) - 1) # 19 successes
left_failure <- N_left - left_success # 50 failures

# Right leaf (power.o < 0.9286641: no)
criterion_right <- train$power.o >= constant
N_right <- length(train$replicated[criterion_right]) # 38 
right_success <- sum(as.numeric(train$replicated[criterion_right]) - 1) # 29 
right_failure <- N_right - right_success # failures
# Individual Gini impurities
gini_left <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
gini_right <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) * gini_left + (N_right / N_total) * gini_right
round(gini_impurity, digits = 5)
```

Control calculations. Quick and dirty.

```{r}
power_sorted <- sort(train$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
gini_impurity <- vector(length = length(adjacent_means))
  for(i in 1:length(adjacent_means) ) {
  constant <- adjacent_means[i]
  N_total <- length(y_train)
  # Left leaf
  criterion_left <- train$power.o < constant
  N_left <- length(train$replicated[criterion_left])
  left_success <- sum(as.numeric(train$replicated[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- train$power.o >= constant
  N_right <- length(train$replicated[criterion_right])
  right_success <- sum(as.numeric(train$replicated[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
  gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
  # Impurity
  gini_impurity[i] <- (N_left / N_total) * gini_d1 + 
    (N_right / N_total) * gini_d2
  }
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjacent_means[which.min(gini_impurity)]
```

#### Sensitivity to observation weights

Normalized initialization weights. Should not change anything to before.

```{r}
m <- nrow(train)
# Equal weight to each observation
D <- rep(1, m)/m
# Create a decision stump
h <- rpart::rpart(formula = replicated ~ ., 
                  data = train[, outpred], 
                  method = "class", weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, digits = 5, box.palette="BuGy")
```

Normalized random observation weights for demonstration. Changes result.

```{r}
# Reproducibility
set.seed(112)
# Random draws from Uniform(0,1)
D_unorm <- runif(107)
# Normalize (sum: 1)
D <- D_unorm/sum(D_unorm) 
# Create a decision stump
cstump_ast <- rpart::rpart(formula = replicated ~ ., 
                  data = train[,outpred], 
                  method = "class", weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(cstump_ast, extra = 100, digits = 5, box.palette="BuGy")
```

# Module 3: Adaptive Boosting 

Predictor "matrix" and outcome "vector". Actually the predictor matrix is a
data frame. The outcome is a factor variable.

```{r}
predictors ; outcome
x_train <- train[, predictors]
y_train <- train[, outcome]
i <- 1 ; m <- nrow(train)
x_train[i,] ; y_train[i]
cbind(x_train, y_train)[i:m,]
```

## Iteration 1 

```{r}
# Outcome vectors 
y_train <- train[, outcome]
(y_test <- test[, outcome])
```

Initialization weights are normalized equal weights.

```{r}
D1 <- rep(1, m)/m
```

Train a single weak learner ($h_t$) with the initialization weights and the
given tree hyperparameter setup.

```{r}
h <- rpart::rpart(formula = replicated ~ ., data = train[,outpred] ,
method = "class", weights = D1, control = hyptree)
```

Make predictions with the weak learner (h_t). Define the normalization constant
nd calculate the TODO

```{r}
y_class <- predict(object = h, newdata = train[, outpred], type = "class")
epsilon <- sum(D1 * (y_train != y_class)) 
```

Compute the training error (e_t) of the weak learner (h_t). This is simply one
minus the weighted accuracy, given weights D

TODO: alternative for performance accuracy weighting.

```{r}
perf <- sum(D1 * (y_train == y_class)) 
epsilon <- 1 - perf 
```

Compute the the weight (a_t) of the weak learner (h_t).
Reduce the learner weight by a factor eta (learning rate).

```{r}
eta <- 1 
alpha1 <- 1/2 * log(perf / epsilon) * eta
```

Second weighting process: learner weights. The better the model performs (that
is, the lower its error) the more weight it gets in the overall ensemble -
independent of the learning rate $\eta$.

```{r}
fat <- function(x, eta) 1/2 * log(x/(1-x)) * eta
curve(fat(x, 1), from = 0.5, to = 1, lwd = 2, 
      ylab = "Importance", xlab = "Performance")
grid()
curve(fat(x,0.5), from = 0.5, to = 1, lwd = 2, col = "ivory4", add = TRUE)
curve(fat(x,0.1), from = 0.5, to = 1, lwd = 2, col = "steelblue", add = TRUE)
legend("topleft", legend = c(expression(eta*" = 1"), expression(eta*" = 0.5"),  
                             expression(eta*" = 0.1")),
       col = c("black", "ivory4", "steelblue"), lty = c(1, 1, 1), lwd = 2)
```

Indicator function that gives a one for correct classifications (prediction is
equal to observed value in the training set) and a minus one for a
misclassification (prediction is unequal to observed value in the training set).

```{r}
chi1 <- (y_train == y_class) * 1 + (y_train != y_class) * -1
```

Update the sample weights. Define the normalization constant and (re)normalize
to maintain the probabilistic interpretation.

```{r}
D2_unorm <- D1 * exp(-alpha1 * chi1)
Z1 <- sum(D2_unorm)
D2 <- D2_unorm / Z1
```

Weight updating process. Answers, why correctly classified training data points
are less reduced than the falsely classified are increased.

```{r}
feat <- function(at, x) exp(-at*x)
curve(feat(0.719, x), from = -1, to = 1, ylim = c(0,3), lty = 3,
      xlab = "Classification", ylab = "Weight update factor",  xaxt = "n") 
grid()
points(c(1,-1), c(feat(0.719, 1), feat(0.7, -1)), pch = 20, cex = 2)
curve(feat(0.619,x), from = -1, to = 1, ylim = c(0,2), col = "ivory4", lty = 3,
      add = TRUE) 
points(c(1,-1), c(feat(0.619, 1), feat(0.6, -1)), pch = 20, cex = 2, col = "ivory4")
curve(feat(0.519,x), from = -1, to = 1, ylim = c(0,2), col = "steelblue", lty = 3,
      add = TRUE) 
points(c(1,-1), c(feat(0.519, 1), feat(0.51, -1)), pch = 20, cex = 2, 
       col = "steelblue")
abline(h = 1, lty = 2, lwd = 1.5, col = "gray30")
legend("topright", legend = c(expression(a[t]*" = 0.719"), 
                              expression(a[t]*" = 0.619"), 
                              expression(a[t]*" = 0.519")),
       col = c("black", "ivory4", "steelblue"),
       lty = c(1, 1, 1), lwd = 2)
axis(1,at = c(-1,1))
```
$a = f(e, \eta)$ The more weight $a$ we assign to an individual learner the
stronger the effect on the observation weights (increase/decrease).
Misclassifications are always more penalized (increase in observation weight)
than correct classifications (decrease in observation weights).

Visualize the result of the weight updating process.

```{r}
n_25 <- 25
scaling <- 175
plot(c(1, n_25), c(0, D1[1] + 0.02) , type = "n", xaxt = "n",
     xlab = "Data point", ylab = "Weight")
axis(1, at = 1:n_25, labels = paste(1:25), cex = 0.5)
for(i in 1:n_25) {
  lines(c(i,i),c(D1[i], D2[i]), lty = 3)
}
points(D1[1:n_25], pch = 20, col = "ivory4", cex = D1[1] * scaling)
points(D2[1:n_25], pch = 20, col = "steelblue", cex = D2[1:n_25] * scaling) 
legend("topright", legend = c(expression(D[1]), expression(D[2])), 
       col = c("ivory4", "steelblue"),
       pch = c(20, 20))
```

## The loop 

Put everything in a loop.

```{r}
hyptree <- rpart::rpart.control(maxdepth = 1) 
T <- 1e4 # Number of trees 
eta <- 1 # Learning rate 
H <- vector("list", T)
D <- rep(1, m)/m
for (t in seq_len(T)) {
  h <- rpart::rpart(formula = replicated ~ ., data = train[, outpred] ,
  method = "class", weights = D, control = hyptree)
  y_retro <- predict(object = h, newdata = train[, outpred], type = "class")
  e <- sum(D * (y_train != y_retro)) 
  a <- 0.5 * log( (1 - e)  / e) * eta
  chi <- (y_train == y_retro) * 1 + (y_train != y_retro) * -1
  D_unorm <- (D * exp(-a * chi))
  D <- D_unorm/sum(D_unorm) 
  # H[[t]] <- list("h" = h, "a" = a, "D" = D, "chi" = chi, "y_retro" =
  # as.numeric(y_retro))
  H[[t]] <- list("h" = h, "a" = a)
}
names(H) <- paste0("t", seq_len(T))

cat("Model weight 1:", H$t3$a) 
```

Make retrodictions.

```{r}
h <- lapply(H, "[[", "h")
a <- sapply(H, "[[", "a")
y_retro_stumps <- sapply(h, predict, 
                         newdata = train[, outpred], 
                         type = "vector")
y_retro_stumps <- 2 * y_retro_stumps - 3 
y_retro_ada <- sign(a %*% t(y_retro_stumps))
table(y_train, y_retro_ada) # In-sample 
```

Alternative for non-algebraists.

```{r}
for (n in seq_len(m)) {
  y_retro_ada[n] <- sign(sum(a * y_retro_stumps[n, ]))
}
```

Prepare the setup for the decision boundary. Note: This might run for 5 minutes.

```{r, eval=FALSE}
# Outcome vector
y_obs <- train[,outcome]
# Make 0/1 factor into something ADA can handle (-1/1)
y_obs_ml <- 2 * as.numeric(train[,outcome]) - 3
# Number of classes
k <- length(unique(y_obs_ml))
# Range(s) as list
X_obs <- train[,c("power.o", "n.o")]
rng <- lapply(X_obs, range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f)
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Make grid
# Tree hyperparameter setting
tree_hyperpar <- rpart::rpart.control(
  # Slightly different hyperparameters to improve visibility of boundary
  maxdepth = 3, minsplit = 5, minbucket = round(5 / 3),
  cp = 1e-2, split = "gini")
# Fit ADAboost
fit <- train_adaboost2(formula = replicated ~ .,
                       data = train[,c("replicated", "power.o", "n.o")],
                       T = 5e2, eta = 1, tree_hyperpar = tree_hyperpar)
# Predict ADAboost
y_pred_ml <- test_adaboost2(fit, grid)
# Make prediction 0/1 instead of -1/1
y_pred <- as.factor((y_pred_ml + 1) / 2)
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_obs == "failure", col1, col2)
grid_colors <- ifelse(y_pred == 0, col1_transp, col2_transp)
```

Visualize.

```{r, eval = FALSE}
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
     xlab = "power.o", ylab = "n.o")
points(grid, col = grid_colors, pch = ".")
# points(grid, col = as.integer(y_pred) + 6L, pch = ".")
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# For the black line 
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
        col = "black", levels = (1:(k-1))+.5)
points(X_obs, col = symbol_colors, pch = as.integer(y_obs) + 15)
legend("topleft", c("success", "failure") , pch = c(17,16), 
       col = c(col2, col1), cex=0.7)
```

Make predictions.

```{r}
x_test <- test[, predictors]
y_pred_stumps <- sapply(h, predict, 
                        newdata = test[, outpred], 
                        type = "vector")
y_pred_stumps <- 2 * y_pred_stumps - 3 
y_pred_ada <- sign(a %*% t(y_pred_stumps))
table(y_test, y_pred_ada)
```

Make predictions. 

# Module 4: Leave-Subject-Out Cross-Validation

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

Introduce new functions with some additional sugar.

## Training function

```{r}
# debugonce(train_adaboost3)
H <- train_adaboost3(formula = replicated ~ .,  
                     data = train[,outpred],   
                     T = 1e4, eta = 1,  # Boosting hyperparameter
                     tree_hyperpar = hyptree) # Tree hyperparameter
```
## Testing function

```{r}
ytilde_num <- test_adaboost3(H = H, data = x_test)
test$ytilde <- factor( (ytilde_num + 3) / 2, labels = c("failure", "success") )
```

## Performance evaluation

```{r}
ROC <- AUC::roc(test$ytilde, test$replicated)
# Area under the curve
AUC::auc(ROC)
```

## LOOCV

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    message("\nThe ", paste0(project), " model.\n")
    # Define the testing set
    test <- df_split[[project]]
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    H <- train_adaboost3(formula = replicated ~ ., 
                         data = train[,outpred],
                         T = 1e4, eta = 1,  # Boosting hyperparameter 
                         tree_hyperpar = hyptree) # Tree hyperparameter
    # Make predictions on validation set 
    ytilde_num <- test_adaboost3(H = H, data = test[, predictors]) 
    ytilde <- factor( (ytilde_num + 3) / 2, labels = c("failure", "success") )
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(ytilde, test$replicated) )
  }})
# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")
c(unlist(auc), "mean" = mean(unlist(auc)))
```

This will run for approximately 4 minutes.

# Module 5: Hyperparameter Optimization using LPO-CV

```{r}
# TODO: Increase n_trees 1e2:1e5 and reduce eta 1e-2 : 1e-5
hgrid <- expand.grid(
  # Tree-specific 
  maxdepth = 1:3, # Interaction depth
  # Ensemble specific
  n_tree = c(1e1, 1e2), # Number of trees
  eta = c(1, 0.1), # Learning rate
  # Container for fold performance
  ee = NA, ml1 = NA, ml3 = NA, rpp = NA,  ssrp = NA, 
  auc.mean = NA, runtime = NA) # Average performance
# Number of combinations 
hgrid_rows <- nrow(hgrid)
```

Hyperparameter optimization using Leave-Project-Out Cross-validation.

```{r}
# Partition the data by "Project ID"
df_split <- split(df[,voi], df$pid) 
# Setup a progress bar
pb <- txtProgressBar(0, hgrid_rows, initial = 0, style = 3) 
# For each entry in the hypergrid, do...
for(i in seq_len(hgrid_rows)) {
  # Track the iteration time (time-accuracy trade-off)
  runtime <- system.time({
    # Leave-One-Project-Out Cross-Validation
    for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
      # Define the testing set
      test <- df_split[[project]] 
      # Define the training set
      train <- df[!(df$pid %in% project), ]
      # Fit the model to the data to get the ensemble
      H <- train_adaboost3(formula = replicated ~ ., 
                                  data = train[,outpred],
                                  T = hgrid$n_tree[i],
                                  eta = hgrid$eta[i],  
                                  tree_hyperpar = rpart::rpart.control(
                                   maxdepth = hgrid$maxdepth[i]))
      # Make predictions on validation set 
      ytilde_num <- test_adaboost3(H = H, data = test[, predictors]) 
      ytilde <- factor( (ytilde_num + 3) / 2, labels = c("failure", "success") )
      # Store the performance evaluation (level: effects)
      hgrid[[project]][i] <- AUC::auc( AUC::roc(ytilde, test$replicated)) }
    # Calculate the average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i,4:8]))
  }) 
  hgrid$runtime[i] <- runtime[["elapsed"]]
  message("\nOverall progress:")
  # Progress bar
  setTxtProgressBar(pb, i) ; cat("\n\n")
}
# Close connection 
close(pb)
```

Performance evaluation sorted by highest AUC.

```{r}
# Note: EE is highly anomalous!
hgrid[order(-hgrid$auc.mean), ] 
```

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1.

## Optimized Ensemble 

If we do not stack the results, we should at least fit the model with the
hyperparameters that had the best average performance (here AUC) across all
replication projects. However, we don't get a true sense of the model's
out-of-sample capabilities because we don't have another project it hasn't seen
yet.

```{r}
# Define the testing set
test <- df_split[[project]] # TODO: Maybe inconsistent
# Define the training set
train <- df[!(df$pid %in% project), ]
# Fit the model to the data to get the ensemble
ensemble <- train_adaboost(train[,features], train[,target], 
                           n_trees = 1e4,  eta = 1e-3, 
                           tree_hyperpar = rpart::rpart.control(
                                 maxdepth = 3))
# Make predictions on validation set 
y_pred <- test_adaboost(test[,features], ensemble)$class
```

# Module 6: Stacking (?) 

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1. 

