---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

- Why Boosting?

- Why AdaBoost?

- What is Boosting 

- What means adaptive? (see, e.g. Kunapuli, 4.2.1)

- Comparison to prediction markets: lay person (boosting) vs experts (random forest)


# Approach (hands dirty)

- Describe the approach taken
  - Why implementing it yourself? 
    - Understanding
    - The role of hyperparameters!
    - Hyperparameters and tuning! 

- Note Florian Pargent et al.'s paper (supplement not complement)
  - For basic terminology (e.g., bias-variance trade-off, classification, AUC)
  - For the right mindset (e.g., predictions in psychology)
  - and other things

# Software (maybe together with Approach)

- Describe the software used

```{r setup}
library(adatutoR)
```

- Describe the data set, i.e. data sets used


```{r}
# Load the data
data("altmext723")
```

- TODO: Document the data set in R so that it is accessible with `?altmext723`



# Module 1: Splitting 


Full data set from Adam Altmejd and colleagues plus the SSRP data.

```{r}
# Variables of interest
voi <- c("pid","eid", "power.o", "effect_size.o", "n.o", "log_p.o", "replicated")

# Outcome (classification task) 
target <- "replicated"

# Predictors 
features <- c("power.o", "effect_size.o", "n.o", "log_p.o")

# Note: No missing values 
df <- altmext723[,voi]
```

## Training and testing split 

Start by randomly splitting the data in a training and testing set (70/30).

### Simple random sampling 

```{r}
# Reproducibility
set.seed(112)

# Split proportion
sprop <- 0.7

# Number of effects 
n_effect <- nrow(df)

# Number of training samples
n_train <- round(n_effect * sprop)

# Training index
train_idx <- sample(seq_len(n_effect), n_train)

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

### Stratified sampling & imbalances

However, there are imbalances in the data set. 

- What is and why use stratified sampling - with the outcome as stratum? 
"Recall that stratified means the folds are created in such a way that the class
distribution is preserved across the folds. This also helps with imbalanced data
sets, as stratification ensures that data from all classes is represented." (p. 103)


```{r}
table(df$replicated)
```

Therefore we should do a stratified split. This means we need to partition the
data into groups based on a stratum variable - here: replicated, and then draw
a random sample from each stratum.

```{r}
# Reproducibility
set.seed(112)

# Partition the data based on the outcome variable
df_split <- split(df, df$replicated)

# Number of training samples (stratified)
n_train <- lapply(df_split, function(df) round(nrow(df) * sprop))

# Training index (stratified)
strat_idx <- mapply(function(df, n) sample(rownames(df), n), df_split, n_train)

# Training index 
train_idx <- as.numeric(unlist(strat_idx))

# Training and testing set 
train <- df[train_idx, ]
test  <- df[-train_idx, ] 

# Feature matrices
X_train <- train[,features] 
X_test <- test[,features] 

# Target vectors 
y_train <- train[,target]
y_test <- test[,target]
```

Now the training and testing set have approximately the same proportion of
replication and failure as the original outcome variable. We could also use the
`rsample` package for that.

```{r, eval=FALSE}
# strat_split <- rsample::initial_split(df, prop = 0.7, strata = "replicated")
# train <- rsample::training(strat_split)
# test <- rsample::testing(strat_split)
```

# Module 2: AdaBoost

First we fit the model to the data (training) and then test its predictive
capabilities on unseen data (testing). Since we will use the data to find
the best hyperparameters this is better thought of as validation set.

- Discuss a couple of iterations of boosting (Kunapali, p.90) 

- (Extra: Decision stumps)

## An iterative process...

Setup.

```{r}
n_trees <- 100
# Extract the number of rows
n_samples <- nrow(X_train)
# Initialize the weights
D <- rep(1, n_samples)/n_samples
# Pre-allocate the empty ensemble
H <- vector("list", n_trees)
df <- data.frame(X_train, y_train)
```

### Digression: Classification stumps and shallow trees 

- Why learn about classification trees with AdaBoost?

The original AdaBoost algorithm boosts classification stumps or shallow
classification trees (ref.). Therefore, it seems obligatory to delve into this
workhorse machine-learning model. We will use the `rpart` package (ref.) to
build understanding in a hands-on manner and `rpart.plot` to visualize the
results.

##### Classification problem

The data set contains measurements on the power of the original study, its
effect size, there is information about its number of observations available,
and so on (see `?altmext723`, for the complete list). Given these measurement in
the training data, we want to learn/extract a pattern between features and
outcomes that allows to accurately predict whether a replication attempt is
successful (class 1) or fails (class 2) (see e.g., Breimann et al., 1984, p.2
for this view). The final evaluation is then done in Part xxx. For now focus is
on the strategy of classification trees to address this problem.

### A short primer on classification tree

Binary *classification trees* address this classification problem by asking
repeated (yes/no) questions about some feature of the data, allowing them to
organize the data into non-overlapping groups - more precisely "*partitions*":
Is the study's power greater than or equal to 0.9? If it is, the study goes into
the left group or *branch*; if it is not, it goes into the right branch.
According to this (splitting) rule, there are two resulting branches; one
containing studies with `power.o >= 0.9` and another with `power.o < 0.9`. (See
Figure 1, left)

A second question targeting all high-powered studies in the left branch
(`power.o >= 0.9`) could lead to a finer partitioning at a lower level (depth
2): Is the number of observations in the original study also greater than 100
(`n.o > 100`)? If so, move the study to the lower left branch. If not, move the
study to the lower right branch; and so on. Classification trees continue the
grouping process until the data cannot be further divided or a termination
criterion is reached. For example, termination criteria can be a minimum number
of 100 studies in the final group, or a minimum number of 100 studies to
consider a split (see e.g. `?rpart.control` for available options in `rpart`).

```{r}
subdf <- df[, c("power.o", "n.o", "y_train")]
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"

# Create a decision tree 
h <- rpart::rpart(formula = y_train ~ ., data = subdf , method = "class", 
                  control = hyptree)

# Visualize
rpart.plot::rpart.plot(h, main = "Classification Stump", extra = 102, 
                       digits = 5, box.palette="BuGy")

# Number of classes
k <- length(unique(subdf$y_train))
# Range(s) as list
rng <- lapply(subdf[-length(subdf)], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Stump")
points(grid, col = grid_colors, pch = ".")
points(subdf[-length(subdf)], col = symbol_colors, pch = as.integer(y_train) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)

# Maybe as comparison the decision boundaries from a GLM?
mdl <- glm(y_train ~., data = subdf, family=binomial(link='logit'))
# Change class attribute
class(mdl) <- c("lr", class(mdl))
# Predicted probability of replication success
y_prob <- predict(mdl, grid, type = "response")
# If p > 0.5 "replicated", else "not replicated"
y_pred <- factor(as.integer(y_prob > 0.5))
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "no" means "not replicated", "yes" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == 0, col1_transp, col2_transp) 
# Plot the features
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
     xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Bernoulli Logit Generalized Linear Model")
points(grid, col = grid_colors, pch = ".")
points(subdf[-length(subdf)], col = symbol_colors, pch = as.integer(y_train) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
        col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
       col = c(col2, col1), cex=0.7)
```

Binary *classification stumps* shorten the repetitive process by limiting
themselves to a single (yes/no) question, such as "`power.o >= 0.9`? (see Figure
1, left). Consequently, they terminate at a maximum depth of one to partitioning
the data into two final groups. (See Figure 1, left side & center). Calling the 
starting point of the tree its root or the *root node* and the terminal point
its leaves or *leave nodes*, a stump can be further classified as a tree with
one root and two leave nodes (see, ref.).

The final decision about whether studies with a given feature, such as a high
power, are considered successful or not is made by the stump based on a majority
vote. There are more failed replication attempts for studies in the left branch
so the overall prediction for studies with this feature "failure" In the right
branch there are more successes, so studies that fall within the right branch
are considered successful.

```{r}
split_criterion <- df$power.o < 0.92866
# Studies with df$power.o < 0.92866 
table(df$y_train[split_criterion])
# Studies with df$power.o >= 0.92866 
table(df$y_train[!split_criterion])
```
To reproduce the graphic we use the following code which fits a classification
stump with `rpart` (ref.) and visualizes the result with `rpart.plot` (ref.).

```{r}
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  split = "gini") 
# Fit a classification stump to the data
cstump <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(cstump, main = "Classification Stump", extra = 102,
                       digits = 5)
```

We call the parameter(s) we pass to `rpart.control()` (tree) hyperparameters. In
general, *hyperparameter* are external parameter that we as the user set *before*
fitting the model to the data to set its complexity (e.g., `maxdepth=1` or
`minbucket =5`) or influence the learning process (e.g., `split = gini`).
Psychologists usually learn about hyperparameters in the context of
(non-Bayesian, exploratory) factor analysis, for example, where they have to
decide on a number of factors to retain before being able to evaluate a
particular solution of the model - the number of factor problem (Lorenzo-Seva).
If working with hyperparameters seems unfamiliar, we recommend doing the
Additional Exercise 1 to solidfy and extend understanding of the previous
concepts.

**Additional exercise 1**: Access the `rpart.control()` documentation with
`?rpart.control` and get an overview over all hyperparameters of the `rpart`
package. Increase the maximum depth of the stump to two and three to create
shallow classification trees. Then, set the minimum number of observations in 
the terminal node to 1. How will these setting influence the shape of 
the tree. Think about the solution first, before running the code. (Are the
results senseful give your general knowledge about meta-science)

### Gini impurity

One question that remains unanswered so far is why `power.o > 0.93`? The sort
answer is, splitting the data according to this feature leads to the lowest
misclassification risk - as judged by Gini impurity.

Say: Gini impurity determines how our model learns (see above).

Make more abstract in one sentence and then apply to example:
The Gini impurity considers how often we would misclassify a replication study
(e.g. say it would "successfully replicate" when it would not) if we were to
randomly guess its class label (success versus failure) based on the class
distribution in that group.
Example: So the Gini impurity measures how often of misclassifying a replication
study's outcome (success or failure) if we were to randomly assign a label
(success or failure) based on the distribution of successes and failures within
our group of replication studies split according to a specific power threshold
(such as power.o >= 0.83).

[...] and by choosing the feature or split with the lowest Gini impurity helps
us minimizing this misclassification risk.


```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
```

The split point is chosen by sorting the individual values of a given feature
and then calculating the Gini impurity for each mid-point between adjacent values.


```{r}
# Example above:
constant <- adjacent_means[69] # 0.9286641
N_total <- length(y_train) # N_total: 107

# Left leaf (power.o < 0.9286641: yes)
criterion_left <- df$power.o < constant
N_left <- length(df$y_train[criterion_left]) # 69 total
left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1) # 19 successes
left_failure <- N_left - left_success # 50 failures

# Right leaf (power.o < 0.9286641: no)
criterion_right <- df$power.o >= constant
N_right <- length(df$y_train[criterion_right]) # 38 total
right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) # 29 successes
right_failure <- N_right - right_success # failures
```


```{r}
# Individual Gini impurities
gini_left <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
gini_right <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
```

Gini left: The Gini impurity suggests that if we were to randomly pick a
replication from our sample where the original had a power of less than 0.93 and
say that it would fail to replicate, there is approximately a 39.91% chance that
we would be wrong.

Gini right: The Gini impurity suggests that if we were to randomly pick a
replication from our sample where the original had a power of greater than 0.93 and
say that it would successfully replicate, there is approximately a 36.15% chance
that we would be wrong.

We finally weight the results to get the overall weighted Gini impurity
condidering the different class distributions.

```{r}

# Overall weighted Gini impurity 
gini_impurity <- (N_left / N_total) * gini_left + (N_right / N_total) * gini_right
round(gini_impurity, digits = 5)
```


G = 0 indicates a perfectly pure node (all elements belong to the same class).
G = 0.5 indicates maximum impurity (elements are evenly distributed among all classes).

Estimating the Gini impurity

Gini impurity = 1 - (probability of the most frequent class)^2 - (probability of
the second-most frequent class)^2 - ... (and so on for all classes)

When you're building a decision tree, you calculate the Gini impurity for each
possible split (question) you could make on the data. The question that leads to
the lowest Gini impurity (the purest resulting groups) is the one you choose.

Gini Impurity in Action

Group 1 (e.g., power > 0.8): What's the Gini impurity of this group in terms of
replication success vs. failure labels? Are the studies in this group mostly
successful replications, mostly failures, or a mix?

Group 2 (e.g., power ≤ 0.8): Calculate the Gini impurity for this group as well.
        
Choosing the best Split: The algorithm selects the power threshold split that
results in the lowest overall Gini impurity in the new groups. This split
maximizes class purity, meaning it clearly separates studies more likely to
succeed in replication from those more likely to fail.






For example, lets say we have



1. step gini_d1 and gini_d2: You're correctly calculating the standard Gini
impurity for the left and right groups separately using the formula: 1 -
(probability of class 1)^2 - (probability of class 2)^2 - ...

2. step (N_left / N_total) and (N_right / N_total):  These represent the weights
assigned to each group's Gini impurity, based on the proportion of examples in
each group relative to the total.

3. step: gini_impurity:  The final line combines the weighted Gini impurities of both
groups to calculate the overall weighted Gini impurity for the potential split.

TODO: Calculations seem to be different from, e.g.: https://machinelearningmastery.com/classification-and-regression-trees-for-machine-learning/

Additional exercise: We leave it as an additional exercise to the reader to
verify that the result has the lowest Gini impurity among all splits (a) within
the particular features and (b) overall.

# Control calculations 






```{r}
subdf <- df[, c("power.o", "n.o", "y_train")]
# Hyperparameter setup 
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 3, 
  # Minimum number of observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
  # Create a decision tree 
h <- rpart::rpart(formula = y_train ~ ., data = subdf , method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Classification Tree", extra = 102, 
                       digits = 5, box.palette="BuGy")

# With minbucket = 4 i can reproduce the following values. 
# Check: Does this make sense? Maybe ignore the decision tree and just use
# decision stump. Or describe that the last split is nonsensical.

table(subdf$y_train[subdf$power.o < 0.92866 & subdf$n.o >= 24])
table(subdf$y_train[subdf$power.o < 0.92866 & subdf$n.o < 24])
table(subdf$y_train[subdf$power.o >= 0.92866 & subdf$power.o >= 1])
table(subdf$y_train[subdf$power.o >= 0.92866 & subdf$power.o < 1])
```

```{r}
# Number of classes
k <- length(unique(subdf$y_train))
# Range(s) as list
rng <- lapply(subdf[-length(subdf)], range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3 
# Colors
col1 <- "cornflowerblue" ; col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f) 
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
y_pred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(y_pred), nrow = resolution)
# "0" means "not replicated", "1" means "replicated"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicated", "1" means "replicated"
grid_colors <- ifelse(y_pred == "failure", col1_transp, col2_transp)
plot(NULL, xlim = range(x1_seq), ylim = range(x2_seq), 
   xlab = "power.o", ylab = "n.o", main = "Decision Boundary")
mtext("Classification Tree")
points(grid, col = grid_colors, pch = ".")
points(subdf[-length(subdf)], col = symbol_colors, pch = as.integer(y_train) + 15, )
contour(x1_seq, x2_seq, z_mat, add = TRUE, drawlabels = FALSE, lwd = 2, 
      col = "black", levels = (1:(k-1))+.5)
legend("topleft", c("success", "failure") , pch = c(17,16), 
     col = c(col2, col1), cex=0.7)

```

Decision stumps simplify this repetitive partitioning process by asking only a
single question about some feature of the data. For example, is the power of the
original study greater than . Classification stump is a classification tree of
depth 1.


One question that is left unanswered now, is how do we get to a particular
question that allows us to parition the data in the way we've seen?

- Splitting criterion 
- Focus on CART with Gini impurity

A selection of most important tree hyperparameters for our AdaBoost
implementation.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree 
  maxdepth = 1, # Boost stumps
  # Original CART with gini impurity as splitting criterion  
  split = "gini") # or ID3: split = "information"
```

Fit a decision stump to the training data and visualize the result.

```{r}
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump", extra = 102, digits = 5)
```

- What do we see?
  - Why `digits =5`?

- Why this particular split?
  - Gini impurity
  - Sensitivity to observation weights (Assumption: equal) 


```{r}
power_sorted <- sort(df$power.o)
adjacent_means <- (power_sorted[-1] + power_sorted[-length(power_sorted)]) / 2
gini_impurity <- vector(length = length(adjacent_means))
  for(i in 1:length(adjacent_means) ) {
  constant <- adjacent_means[i] 
  N_total <- length(y_train) 
  # Left leaf 
  criterion_left <- df$power.o < constant
  N_left <- length(df$y_train[criterion_left])
  left_success <- sum(as.numeric(df$y_train[criterion_left]) - 1)
  left_failure <- N_left - left_success 
  # Right leaf
  criterion_right <- df$power.o >= constant
  N_right <- length(df$y_train[criterion_right]) 
  right_success <- sum(as.numeric(df$y_train[criterion_right]) - 1) 
  right_failure <- N_right - right_success 
  # Gini index
  gini_d1 <- 1 - ( (left_failure / N_left)^2 + (left_success / N_left)^2 )
  gini_d2 <- 1 - ( (right_failure / N_right)^2 + (right_success / N_right)^2 )
  # Impurity
  gini_impurity[i] <- (N_left / N_total) * gini_d1 + (N_right / N_total) * gini_d2
  }
# Position of smallest impurity 
which.min(gini_impurity)
# Outcome 
adjacent_means[which.min(gini_impurity)]
```


#### Sensitivity to observation weights

Normalized initialization weights. Should not change anything to before.

```{r}
# Equal weight to each observation
D <- rep(1, n_samples)/n_samples
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

Normalized random observation weights for demonstration. Changes result.

```{r}
# Reproducibility
set.seed(113)
# Random draws from Uniform(0,1)
D_unorm <- runif(107) 
# Normalize (sum: 1)
D <- D_unorm/sum(D_unorm)
# Create a decision stump
h <- rpart::rpart(formula = y_train ~ ., data = df , method = "class", 
                  weight = D, control = hyptree)
# Visualize
rpart.plot::rpart.plot(h, main = "Decision Stump")
```

##### Additional exercises

`rpart` offers a number of hyperparameters that affect the outcome of the
results. We leave it to the reader as an additional exercise to explore how each
of them influences the result. For our purposes, we will focus only on the
maximum depth and the splitting criterion.

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree (root node: depth = 0)
  # Important: We usually boost stumps (depth = 1) or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini" # or ID3: split = "information"
  )
```

##### Additional exercise 1

We created a decision stump using the gini impurity as splitting criterion.
Adapt the code above to fit a shallow tree of depth two and three.

##### Additional exercise 2

Despite the hyperparameters we already mentioned there are more. See the
`?rpart::rpart.control` documentation and explore how changing different
arguments influence the results.


### Adaboost continued

Normalized weights.

```{r}
# Initialize the weights (normalized)
D <- rep(1, n_samples)/n_samples
```

Adaptive boosting.

```{r}
# for (tree in 1:n_trees) {
  # Train a single weak learner (h_t) with normalized(!) sample weights (D)
  h <- rpart::rpart(formula = y ~ ., data = df , method = "class",
                    weights = D, control = tree_hyperpar)
  # Make predictions with the weak learner (h_t)
  y_pred <- predict(object = h, newdata = X, type = "class")
  # y_pred <- 2*as.numeric(y_class) - 3
  wacc <- sum(D * (y == y_pred)) / sum(D)
  # Compute the training error (e_t) of the weak learner (h_t)
  # This is simply one minus the weighted accuracy, given weights D
  e <- 1 - wacc
  # Equivalent but less readable: e <- sum(D * (y != y_pred)) / sum(D)
  # Compute the the weight (a_t) of the weak learner (h_t)
  # Reduce the learner weight by a factor eta (learning rate)
  a <- 0.5 * log( (1 - e)  / e) * eta
  # Pre-Prunning or Early Stopping:
  # Do not forget to add the parameter (prunning threshold)
  # if (a < prunning_threshold) {
  #  message("Prunning threshold reached")
  #  break
  # }
  # Updates the weights
  # Increase for misclassified examples: (y == y_pred)
  # decrease for correctly classified examples: (y!=y_pred)
  m <- (y == y_pred) * 1 + (y != y_pred) * -1
  # Update the sample weights and limit the contribution of h_t by eta
  D <- D * exp(-a * m)
  # Normalize weights (probabilistic interpretation)
  D <- D/sum(D)
  # Save the weak learner and its weight
  H[[tree]] <- list("learner" = h, "learner_weight" = a,
                    "observation_weight" = D)
  # Store sample weights for visualization
}
  # Return the ensemble
# H
}
```

```{r}
# Plot
# rattle::fancyRpartPlot(h, caption = NULL)
# Sys.sleep(2)
```


## Hands dirty: Implementation

### Training 

Look under the hood of the function 

```{r}
train_adaboost
```

Tree hyperparameter settings

```{r}
hyptree <- rpart::rpart.control(
  # Maximum depth of any node tree
  # Root node countes as depth = 0
  # Note: Values greatet 30 will give nonsense results on 32-bit machines
  # Important: We usually boost stumps (depth = 1), or shallow trees
  maxdepth = 1, # Boost stumps
  # Minimum n° obs in a node for a split to be attempted 
  minsplit = 20, 
  # Minimum n° of obs in any terminal <leaf> node 
  minbucket = round(10 / 3),
  # Complexity parameter: Include in
  # Split must decrease the fit lack of fit by a factor of cp
  cp = 1e-2,
  # Number of cross-validations 
  xval = 10,
  # Original CART with gini impurity as splitting criterion  
  split = "gini"
  # split = "information"
  )
```

Training on the training set with specified hyperparameters

```{r}
ensemble <- train_adaboost(X = X_train, # Feature matrix 
                           y = y_train, # Target vector 
                           # Ensemble hyperparameters
                           n_trees = 1e4, # Number of learner
                           eta = 1e-4, # Learning rate
                           # Tree hyperparameters (see above)
                           tree_hyperpar = hyptree) 
```


### Prediction

```{r}
test_adaboost
```

Make predictions with the trained model and add them to the testing set.

```{r}
y_class <- test_adaboost(X_test, ensemble)$class
test$y_pred <- factor( (y_class + 3) / 2, labels = c("no", "yes") )
```

## Evaluation

```{r}
ROC <- AUC::roc(test$y_pred, test$replicated)
# Area under the curve
AUC::auc(ROC)
```


# Module 3: Leave-Project-Out Cross-Validation

TODO: Check if code in Module 2 is running

Above, we randomly split the data on the level of study effects, taking care
that no trials in the training set is also in the testing set. However, a more
plausible scenario to validate the model is the following: Maybe in the near
future we might get another large-scale replication project. Thus, we would
probably use the existing ones to train the model and then use them to make
predictions for the new one. If so, the cross-validation procedure should mimic
this approach. We will use "Leave-One-Project-Out Cross-Validation" and try to
maximize the average performance of the model on the level of study effects (not
the individual trials) when applied to a new project.

```{r}
# Partition the data by "Project ID"
df_split <- split(df, df$pid) 

# Pre-allocate the AUC list
auc <- list()

# Leave-Project-Out Cross-Validation
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 5,  eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

This will run for approximately 4 minutes.

TODO: This seems to run, but why do I always get the same performance estimates?
No matter if I use 1 tree or 1e4 trees...

```{r}
runtime <- system.time({
  for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
    # Define the testing set
    test <- df_split[[project]] # TODO: Maybe inconsistent
    # Define the training set
    train <- df[!(df$pid %in% project), ]
    # Fit the model to the data to get the ensemble
    ensemble <- train_adaboost(train[,features], train[,target], 
                                 n_trees = 500, eta = 1, hyptree)
    # Make predictions on validation set 
    y_pred <- test_adaboost(test[,features], ensemble)$class
    # Store the performance evaluation (level: effects)
    auc[[project]] <- AUC::auc( AUC::roc(y_pred, test$replicated) )
  }})
# Waiting time
runtime["elapsed"]
```

## Performance estimates

```{r}
auc
```

# Module 4: Hyperparameter Optimization using LPO-CV

TODO: Check if code in Module 3 is running

```{r}
# TODO: Increase n_trees 1e2:1e5 and reduce eta 1e-2 : 1e-5
hgrid <- expand.grid(
  # Tree-specific 
  maxdepth = 1:3, # Interaction depth
  # Ensemble specific
  n_tree = c(1e1, 1e2), # Number of trees
  eta = c(1, 0.1), # Learning rate
  # Container for fold performance
  ee = NA, ml1 = NA, ml3 = NA, rpp = NA,  ssrp = NA, 
  auc.mean = NA, runtime = NA) # Average performance
# Number of combinations 
hgrid_rows <- nrow(hgrid)
```

Hyperparameter optimization using Leave-Project-Out Cross-validation.

```{r}
# Partition the data by "Project ID"
df_split <- split(df[,voi], df$pid) 
# Setup a progress bar
pb <- txtProgressBar(0, hgrid_rows, initial = 0, style = 3) 
# For each entry in the hypergrid, do...
for(i in seq_len(hgrid_rows)) {
  # Progress bar
  setTxtProgressBar(pb, i) 
  # Track the iteration time (time-accuracy trade-off)
  runtime <- system.time({
    # Leave-One-Project-Out Cross-Validation
    for (project in c("ee", "ml1", "ml3", "rpp", "ssrp")) { 
      # Define the testing set
      test <- df_split[[project]] # TODO: Inconsistent (partial matching vs explicit)
      # Define the training set
      train <- df[!(df$pid %in% project), ]
      # Fit the model to the data to get the ensemble
      ensemble <- train_adaboost(train[,features], train[,target],
                                 n_trees = hgrid$n_tree[i],  
                                 eta = hgrid$eta[i], 
                                 tree_hyperpar = rpart::rpart.control(
                                   maxdepth = hgrid$maxdepth[i]))
      # Make predictions on validation set 
      y_pred <- test_adaboost(test[,features], ensemble)$class
      # Store the performance evaluation (level: effects)
      hgrid[[project]][i] <- AUC::auc( AUC::roc(y_pred, test$replicated)) }
    # Calculate the average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i,4:8]))
  }) 
  # TODO: Inconsistent (partial matching vs explicit)
  hgrid$runtime[i] <- runtime[["elapsed"]]
}
# Close connection 
close(pb)
```

Performance evaluation sorted by highest AUC.

```{r}
# Note: EE is highly anomalous!
hgrid[order(-hgrid$auc.mean), ] 
```

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1.

## Optimized Ensemble 

If we do not stack the results, we should at least fit the model with the
hyperparameters that had the best average performance (here AUC) across all
replication projects. However, we don't get a true sense of the model's
out-of-sample capabilities because we don't have another project it hasn't seen
yet.

```{r}
# Define the testing set
test <- df_split[[project]] # TODO: Maybe inconsistent
# Define the training set
train <- df[!(df$pid %in% project), ]
# Fit the model to the data to get the ensemble
ensemble <- train_adaboost(train[,features], train[,target], 
                           n_trees = 1e4,  eta = 1e-3, 
                           tree_hyperpar = rpart::rpart.control(
                                 maxdepth = 3))
# Make predictions on validation set 
y_pred <- test_adaboost(test[,features], ensemble)$class
```

# Module 5: Stacking (?) 

Stacking the best performing algorithms could improve the predictions quite
significantly here! For example, learners 2 and 3 perform quite well on the ee
project, while learners 1 and 8 perform quite well on ml1. 
