---
title: "Code Along Script For The Tutorial"
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Code Along Script For The Tutorial}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Prerequisites

#### Listing 1

Code to install the adatutor package form GitHub, load and attach it.

```{r listing-1}
# Load package
library(remotes)

# Install package from GitHub
# install_github("sbissantz/adatutor")

# Load package
library(adatutor)
```

# Empirical Example

#### Listing 2

Code to load the dataset.

```{r listing-2}
# Load data
data(altmejd)
```

#### Listing 3

Code to prepare the data.

```{r listing-3}
# Reviewer metrics (names)
prednms <- c("power.o", "effect_size.o", "n.o", "p_value.o")

# Outcome variable (name)
outnme <- "replicate"

# Variables of interest
voinms <- c(prednms, outnme)

# Shorten 'altmejd'
df <- altmejd
```

# Module 1: Data Splitting

#### Listing 4

Code to do a 70/30 simple random split.

```{r listing-4}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.7

# Number of effects
n_fx <- nrow(df)

# Number of training samples
n_train <- round(n_fx * prop)

# Training index
train_idx <- sample(seq_len(n_fx), n_train)

# Training set
train <- df[train_idx, ]

# Test set (incl. validation)
testvalid <- df[-train_idx, ]
```

#### Listing 5

Code to complete a 70/15/15 simple random split

```{r listing-5}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.5

# Number of effects
n_fx <- nrow(testvalid)

# Number of validation samples
n_valid <- round(n_fx * prop)

# Training index
valid_idx <- sample(seq_len(n_fx), n_valid)

# Validation set
valid <- testvalid[valid_idx, ]

# Test set
test <- testvalid[-valid_idx, ]
```

#### Listing 6

Code to check for imbalances in the outcome variable of the original dataset
and the test set.

```{r listing-6}
# Frequency tables
tbl_orig <- table(df$replicate)
tbl_testvalid <- table(testvalid$replicate)

# Relative frequencies
proportions(tbl_orig)
proportions(tbl_testvalid)
```

#### Listing 7

Code to do a 70/30 stratified split.

```{r listing-7}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.7

# Partition data based on outcome
set_split <- split(df, df$replicate)

# Number of validation samples
n_train <- lapply(set_split, function(set) round(nrow(set) * prop))

# Training index
strat_idx <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_train
)

# Training index
train_idx <- df$eid %in% unlist(strat_idx)

# Training set
train <- df[train_idx, ]

# Test set (incl. validation)
testvalid <- df[!train_idx, ]
```

#### Listing 8

Code to complete a 70/15/15 stratified split

```{r listing-8}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.5

# Partition data based on outcome
set_split <- split(testvalid, testvalid$replicate)

# Numbers of validation samples
n_valid <- lapply(set_split, function(set) round(nrow(set) * prop))

# Validation index
strat_idx <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_valid
)
valid_idx <- testvalid$eid %in% unlist(strat_idx)

# Validation set
valid <- testvalid[valid_idx, ]

# Test set
test <- testvalid[!valid_idx, ]
```

#### Listing 9

Code to do a 70/30 random and stratified split, with the pre-implemented
functions of the adatutor package.

```{r listing-9}
# For reproducibility
set.seed(112)

# Simple random split
randsplit <- random_split(df, prop = 0.7)
train <- randsplit$set1
testvalid <- randsplit$set2

# Stratified split
stratsplit <- stratified_split(df, group = "replicate", id = "eid", prop = 0.7)
train <- stratsplit$set1
testvalid <- stratsplit$set2
```

#### Listing 10

Code to do a 70/30 random and stratified split, with the rsample package.

```{r listing-10}
# Load package
library(rsample)

# For reproducibility
set.seed(112)

# Simple random split
randsplit <- initial_split(df, prop = 0.7)
train <- training(randsplit)
testvalid <- testing(randsplit)

# Stratified split
stratsplit <- initial_split(df, prop = 0.7, strata = "replicate")
train <- training(stratsplit)
testvalid <- testing(stratsplit)
```

::: {.callout-important}
## Restore the split before going on

Module 1 showed several ways to split, and each one overwrote `train`. Whichever
you ran last is the one still in memory, so from here the numbers would drift
away from the text.

Everything from Module 2 onward is built on the stratified split of Listings 7
and 8. The package ships it, so load it back and the stumps, weights and figures
come out as described. Setting a seed would not be enough on its own: R changed
`sample()` in version 3.6.0 and moved every seeded split in the wild with it,
which is why the sets themselves travel with the package rather than the recipe
for them.
:::

```{r restore-split}
# The split every later module is written against
train <- altmejd_splits$train
valid <- altmejd_splits$valid
test <- altmejd_splits$test
```

#### Figure 2

Visualization of a classification tree of depth one or, equivalently, a classification stump. The stump was fitted to the training data using two reviewer metrics: the power of the original study and the number of observations. The middle rectangle represents the root node, and the two lower rectangles are the leaf nodes. The power value used to partition the root node is 0.9305. The left leaf node contains about 62\% of the training data and the right leaf node contains about 38\%. The label in each node (\enquote{success}, \enquote{failure}) represents the stump's prediction for each study effect in that node. For example, in the left leaf node, the stump predicts that studies with a power of less than 0.9305 will fail replication. According to the training data, 49 of the 66 study effects with a power of less than 0.9305 actually failed replication.

```{r figure-2}
# Where the manuscript keeps its figures. Most readers will not have this
# folder, and then the plots just go to the screen as usual. Each export wraps
# the same call rather than repeating it, so the file and the figure you see on
# screen cannot drift apart.
figdir <- "./inst/figures"
to_file <- dir.exists(figdir)

# One column of the paper's two-column layout, and the full text width, both
# measured from the typeset manuscript. Drawing at the size a figure is printed
# at means nothing is scaled afterwards, which is what keeps the labels legible:
# a figure drawn at the full width and then squeezed into a column loses half
# its type size. The two decision-boundary plots take the wider pair, and they
# are the two the manuscript already sets as `figure*` -- at column width the
# studies collide into one band and the boundary smears.
col_w <- 3.44 * 1.25
col_h <- 2.45 * 1.25
full_w <- 6.94 * 0.8
full_h <- 4.94 * 0.8

# Select only two predictors
# Note: Reduce run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <- train[, union(subfeat, outnme)]
# Hyperparameter setup
treehypar <- rpart::rpart.control(
  # Maximum depth of any node tree
  maxdepth = 1,
  # Minimum n° observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion
  split = "gini"
) # or ID3: split = "information"
# Create a decision tree
h <- rpart::rpart(
  replicate ~ .,
  data = subdf,
  method = "class",
  control = treehypar
)
# Visualize
sty <- viridis_tree(h)

if (to_file) {
  pdf(
    file.path(figdir, "cstump.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

rpart.plot::rpart.plot(
  h,
  extra = 102,
  digits = 4,
  box.col = sty$box,
  col = sty$text
)

if (to_file) {
  dev.off()
}
```

# Module 2: Classification Stumps

#### Listing 11

Code to retrieve the frequency tables for the outcome variable under the two
power conditions.

```{r listing-11}
# Condition for the split
splitcond <- train$power.o < 0.9305

# Studies with power.o < 0.9305
table(train$replicate[splitcond])

# Studies with power.o >= 0.9305
table(train$replicate[!splitcond])
```

#### Listing 12

Code to set the hyperparameters of a classification tree and fit the model to
the training data.

```{r listing-12}
# Load package
library(rpart)

# Hyperparameter setup
treehypar <- rpart.control(
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  split = "gini"
)

# Fit classification stump
cstump <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  control = treehypar
)
```

#### Listing 13

Code to visualize a fitted classification stump.

```{r listing-13}
# Load package
library(rpart.plot)

# Visualize classification stump
sty <- viridis_tree(cstump)
rpart.plot::rpart.plot(
  cstump,
  extra = 102,
  digits = 4,
  box.col = sty$box,
  col = sty$text
)
```

#### Figure 3

Visualization of the decision boundary of a classification stump. The predictor variable on the x-axis is the power of the original study effect. The predictor variable on the y-axis is the number of observations. So the y-axis does not display the outcome variable, but another study characteristic. However, since classification stumps are classification trees of depth one, only one study characteristic (i.e., power in this case) is used as a predictor to separate successfully replicated study effects from failed ones. The decision boundary is set at a power value of 0.9305 (vertical line). The model predicts that any point in the gray region is a successful replication, while anything in the blue region is considered a failure. Gray triangles represent replications of study effects that actually succeeded (i.e., according to the replication metric), while blue circles represent replications that actually failed. Thus, gray triangles in the blue region are misclassifications, and vice versa.

```{r figure-3}
# `dboundary`, not `cstump`: in the manuscript `cstump` is the tree diagram
# above, and this is the boundary that tree draws. Both are `figure*` floats
# there, which is why they get the wider geometry.
if (to_file) {
  pdf(
    file.path(figdir, "dboundary.pdf"),
    width = full_w,
    height = full_h,
    pointsize = 9
  )
}

# Visualize the stump's decision boundary. `shade = "margin"` colours by the
# predicted probability rather than by the class alone, so the two regions read
# as the moderate .26 and .76 they actually are -- and it is the same encoding
# Figure 9 uses, which is what makes the two figures comparable.
plot_adabound(
  h,
  data = subdf,
  shade = "margin",
  xlab = "Power (original study)",
  ylab = "Sample size (original study)"
)

if (to_file) {
  dev.off()
}
```

#### Figure 4

Visualization of the Gini impurity for a two-class classification problem. The x-axis is the proportion or probability of Class 1 (e.g., the proportion of successes in the empirical example). The y-axis is the measure of Gini impurity. The Gini impurity is highest when the classes are equally represented. This occurs when the proportion of Class 1 is 0.5, meaning there are as many instances of Class 1 (i.e., successes) as of Class 2 (i.e., failures). The more homogeneous a class becomes, the lower the Gini impurity. The Gini impurity is zero if there are only instances of one type (i.e., only successes or failures).

```{r figure-4}
if (to_file) {
  pdf(
    file.path(figdir, "gini.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

op <- par(
  mar = c(2.9, 2.9, 1.9, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

# a single curve, so one color rather than a scale -- the same anchor viridis
# uses for the first/darkest entry elsewhere in the tutorial
col <- viridisLite::viridis(1, end = 0.85)

# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01) # Values for fraction of class k
y <- 1 - (x^2) - (1 - x)^2 # Gini Impurity calculation
# Create the plot
plot(
  x,
  y,
  type = "l",
  lwd = 2, # Specify line plot
  col = col,
  xlab = "Proportion of Successes",
  ylab = "Impurity"
)
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1))

par(op)
if (to_file) {
  dev.off()
}
```

#### Listing 14

Code to calculate means of neighboring units and find the position of the
relevant power value.

```{r listing-14}
# Sorted vector
power_srt <- sort(train$power.o)

# Adjacent means
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2

# Find position of value '0.9305'
which(round(adjmeans, 4) == 0.9305)

# Additional

adj_pat <- which(round(adjmeans, 4) == 0.9305)
const <- adjmeans[adj_pat]
N_total <- nrow(train) # 107
criterion_left <- train$power.o < const
N_left <- length(train$replicate[criterion_left]) # 66 total
left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1) # 17
left_failure <- N_left - left_success # 49 failures
# Gini impuritiy (left)
gini_left <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
gini_left # 0.382
criterion_right <- train$power.o >= const
N_right <- length(train$replicate[criterion_right]) # 41
right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1) # 31
right_failure <- N_right - right_success # 10
# Gini impuritiy (right)
gini_right <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
gini_right # 0.3688281
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) *
  gini_left +
  (N_right / N_total) * gini_right
round(gini_impurity, digits = 5) # 0.37724
# Sorted values
power_srt <- sort(train$power.o)
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2
gini_impurity <- vector(length = length(adjmeans))
for (i in seq_along(adjmeans)) {
  constant <- adjmeans[i]
  N_total <- nrow(train)
  # Left leaf
  criterion_left <- train$power.o < constant
  N_left <- length(train$replicate[criterion_left])
  left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- train$power.o >= constant
  N_right <- length(train$replicate[criterion_right])
  right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
  gini_d2 <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
  # Impurity
  gini_impurity[i] <- (N_left / N_total) *
    gini_d1 +
    (N_right / N_total) * gini_d2
}
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjmeans[which.min(gini_impurity)]
```

#### Listing 15

Code to make the assumption of equal weights explicit.

```{r listing-15}
# Load package
library(rpart.plot)

# Number of rows
m <- nrow(train)

# Equal weight to each observation
D <- rep(1, m) / m

# Fit decision stump
cstump <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D,
  control = treehypar
)

# Visualize
sty <- viridis_tree(cstump)
rpart.plot(cstump, digits = 4, box.col = sty$box, col = sty$text)
```

#### Listing 16

Code to show the sensitivity to observation weights by using normalized random
weights from a uniform distribution.

Visualization of a classification stump fitted with normalized random weights. Note that we simplified the visual output shown here by using the \texttt{extra} argument of the \texttt{rpart.plot()} function with a value set to \texttt{100}.

```{r listing-16}
# For reproducibility
set.seed(112)

# Draws from standard uniform
D_raw <- runif(107)

# Normalize (sum to 1)
D <- D_raw / sum(D_raw)

# Create decision stump
cstump_ast <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D,
  control = treehypar
)

# Visualize
sty <- viridis_tree(cstump_ast)

if (to_file) {
  pdf(
    file.path(figdir, "cstumpwght.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

rpart.plot(
  cstump_ast,
  extra = 100,
  digits = 4,
  box.col = sty$box,
  col = sty$text
)

if (to_file) {
  dev.off()
}
```

# Module 3: AdaBoost

#### Listing 17

Code to understand the givens in the AdaBoost algorithm.

```{r listing-17}
# Predictors and outcome (names)
prednms
outnme

# Predictor matrix
x_train <- train[, prednms]

# Outcome vector
y_train <- train[, outnme]

# Indices
i <- 1
m <- nrow(train)

# i-th row or element
x_train[i, ]
y_train[i]

# i-th to m-th row (i.e., all)
cbind(x_train, y_train)[i:m, ]
```

#### Listing 18

Code to generate initialization weights in the AdaBoost algorithm.

```{r listing-18}
# Initialization weights
D1 <- rep(1, m) / m
```

#### Listing 19

Code to fit a classification stump with the initialization weights.

```{r listing-19}
# Hyperparameter setup
treehypar <- rpart.control(maxdepth = 1, split = "gini")

# Fit classification stump
h1 <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D1,
  control = treehypar
)
```

#### Listing 20

Code to calculate the classification stump's error.

```{r listing-20}
# Retrodictions (training data)
yretro <- predict(h1, newdata = train[, voinms], type = "class")

# Retrodictive error
epsilon <- sum(D1 * (yretro != y_train))
```

#### Listing 21

Code to calculate the classification stump's performance.

```{r listing-21}
# Retrodivitve performance
perf <- sum(D1 * (yretro == y_train))

# Retrodictive error
epsilon <- 1 - perf
```

#### Listing 22

Code to calculate the classification stump's weight assuming a learning rate
eta of 1.

```{r listing-22}
# Learning rate
eta <- 1

# Model weight
alpha1 <- 1 / 2 * log(perf / epsilon) * eta
```

#### Figure 6

Visualization of how the importance of a classification stump ($\alpha_t$) varies with its performance ($1-\epsilon_t$). Overall, the better the performance of a stump, the higher its weight. Put another way, the lower a stump's error ($\epsilon_t$), the higher its importance ($a_t$). The boosting hyperparameter ($\eta$) allows to further attenuate the weight of a classification stump. This could later help to improve the predictive performance of AdaBoost (see Module~4).

```{r figure-6}
if (to_file) {
  pdf(
    file.path(figdir, "importance.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

# Tighter margins than the default: at this size the defaults would leave
# almost no room to plot in. `par(op)` at the end puts them back, because the
# whole document runs in one session and the settings would otherwise leak
# into every later figure.
op <- par(
  mar = c(2.9, 2.9, 3.0, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

# three ordered learning rates, so palette runs with them. `end = 0.85`
# stops short of viridis's yellow, which is invisible as a line on white
pal <- viridisLite::viridis(3, end = 0.85)

fat <- function(x, eta) 1 / 2 * log(x / (1 - x)) * eta
curve(
  fat(x, 1),
  from = 0.5,
  to = 1,
  lwd = 2,
  col = pal[1],
  ylab = "Importance",
  xlab = "Performance"
)
grid()
curve(
  fat(x, 0.5),
  from = 0.5,
  to = 1,
  lty = 2,
  lwd = 2,
  col = pal[2],
  add = TRUE
)
curve(
  fat(x, 0.1),
  from = 0.5,
  to = 1,
  lty = 4,
  lwd = 2,
  col = pal[3],
  add = TRUE
)
# Legend sits in the margin above the panel (where `main` is), so it
# never covers a curve. `xpd = NA` is what lets it draw outside the plotting
# region; the default top margin already reserves the room. `text.width = NA`
# gives each entry its own width -- without it `horiz` pads every column out to
# the widest label, which strings short entries across the whole panel.
usr <- par("usr")
legend(
  x = (usr[1] + usr[2]) / 2,
  y = usr[4] + 0.055 * (usr[4] - usr[3]),
  xjust = 0.5,
  yjust = 0,
  horiz = TRUE,
  xpd = NA,
  bty = "n",
  text.width = NA,
  x.intersp = 0.7,
  seg.len = 1.2,
  cex = 0.9,
  legend = c(
    expression(eta * " = 1"),
    expression(eta * " = 0.5"),
    expression(eta * " = 0.1")
  ),
  col = pal,
  lty = c(1, 2, 4),
  lwd = 2
)

par(op)
if (to_file) {
  dev.off()
}
```

#### Listing 23

Code to calculate the update factor for the initialization weights.

```{r listing-23}
# Indicator trick
chi1 <- (yretro == y_train) * 1 + (yretro != y_train) * -1

# Unnormalized weights
D2_raw <- D1 * exp(-alpha1 * chi1)

# Normalization constant
Z1 <- sum(D2_raw)

# Normalized weights
D2 <- D2_raw / Z1
```

#### Figure 7

Visualization of how the update factor ($e^{-\alpha\chi}$) varies depending on the classification result and the importance of the model ($\alpha_t$). If the classifier classifies an observation correctly, $\chi$ takes the value 1 and the weight or importance of the observation is reduced in the next iteration. In the case of a misclassification, $\chi$ takes the value -1 and the weight of the observation increases. The observation will be more important in the next iteration.

```{r figure-7}
if (to_file) {
  pdf(
    file.path(figdir, "updatefactor.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

op <- par(
  mar = c(2.9, 2.9, 3.0, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

# same three-step palette as Figure 6: model weights are ordered too
pal <- viridisLite::viridis(3, end = 0.85)

feat <- function(at, x) exp(-at * x)
curve(
  feat(0.719, x),
  from = -1,
  to = 1,
  ylim = c(0.3, 2.2),
  lty = 3,
  col = pal[1],
  xlab = "Classification",
  ylab = "Weight update factor",
  xaxt = "n"
)
grid()
points(
  c(1, -1),
  c(feat(0.719, 1), feat(0.7, -1)),
  pch = 20,
  cex = 2,
  col = pal[1]
)
curve(
  feat(0.619, x),
  from = -1,
  to = 1,
  ylim = c(0, 2),
  col = pal[2],
  lty = 3,
  add = TRUE
)
points(
  c(1, -1),
  c(feat(0.619, 1), feat(0.6, -1)),
  pch = 17,
  cex = 2,
  col = pal[2]
)
curve(
  feat(0.519, x),
  from = -1,
  to = 1,
  ylim = c(0, 2),
  col = pal[3],
  lty = 3,
  add = TRUE
)
points(
  c(1, -1),
  c(feat(0.519, 1), feat(0.51, -1)),
  pch = 15,
  cex = 2,
  col = pal[3]
)
abline(h = 1, lty = 2, lwd = 1.5, col = "gray30")
usr <- par("usr")
legend(
  x = (usr[1] + usr[2]) / 2,
  y = usr[4] + 0.055 * (usr[4] - usr[3]),
  xjust = 0.5,
  yjust = 0,
  horiz = TRUE,
  xpd = NA,
  bty = "n",
  text.width = NA,
  x.intersp = 0.7,
  seg.len = 1.2,
  cex = 0.9,
  legend = c(
    expression(a[t] * " = 0.719"),
    expression(a[t] * " = 0.619"),
    expression(a[t] * " = 0.519")
  ),
  col = pal,
  lty = c(1, 1, 1),
  pch = c(20, 17, 15),
  lwd = 2
)
axis(1, at = c(-1, 1))

par(op)
if (to_file) {
  dev.off()
}
```

#### Figure 8

Visualization of the result of the weight updating process after Iteration 1 for the first 25 data points in the \texttt{altmejd} dataset. The gray initial weights $D_1$ increase or decrease depending on whether the first classification stump classified them correctly (green dot in the middle) or not (red dot in the middle). If the stump misclassifies an observation, its weight moves up and becomes larger, i.e., its importance increases. Correctly classified data points experience a weight decrease. They become less important in the next iteration. Visually, they move down and become smaller.

```{r figure-8}
if (to_file) {
  pdf(
    file.path(figdir, "weightone.pdf"),
    width = col_w,
    height = col_h,
    pointsize = 9
  )
}

op <- par(
  mar = c(2.9, 2.9, 3.0, 0.6),
  mgp = c(1.7, 0.45, 0),
  tcl = -0.25,
  cex.axis = 0.9
)

n_25 <- 25 # TODO: Do we need this?
scaling <- 2

# D1 and D2 keep a colour each. `end = 0.5` leaves D2 teal rather than
# yellow-green, which would go muddy against the yellow band below
pal <- viridisLite::viridis(2, end = 0.5)
band <- adjustcolor(viridisLite::viridis(1, begin = 1), 0.35)

# area, not diameter: cex scales the width of the point, so the eye reads a
# doubled weight as four times the ink unless we take the square root
bubble <- function(w) sqrt(w / D1[1]) * scaling

wrong <- which(y_train[1:n_25] != yretro[1:n_25])
ylim <- c(0, D1[1] + 0.012)

plot(
  c(0.5, n_25 + 0.5),
  ylim,
  type = "n",
  xaxt = "n",
  xlab = "Data point",
  ylab = "Weight"
)
# the band goes behind everything: misclassification belongs to the data point,
# not to either of its two weights
rect(wrong - 0.42, ylim[1], wrong + 0.42, ylim[2], col = band, border = NA)
box()
# every fifth point: 25 labels do not fit the width of a column
axis(1, at = c(1, seq(5, n_25, 5)))
for (i in 1:n_25) {
  lines(c(i, i), c(D1[i], D2[i]), lty = 3, col = "grey45")
}

points(
  1:n_25,
  D1[1:n_25],
  pch = 21,
  bg = pal[1],
  col = "white",
  cex = bubble(D1[1:n_25]),
  lwd = 1.3
)
points(
  1:n_25,
  D2[1:n_25],
  pch = 21,
  bg = pal[2],
  col = "white",
  cex = bubble(D2[1:n_25]),
  lwd = 1.3
)
usr <- par("usr")
legend(
  x = (usr[1] + usr[2]) / 2,
  y = usr[4] + 0.055 * (usr[4] - usr[3]),
  xjust = 0.5,
  yjust = 0,
  horiz = TRUE,
  xpd = NA,
  bty = "n",
  text.width = NA,
  x.intersp = 0.7,
  seg.len = 1.2,
  cex = 0.9,
  legend = c(expression(D[1]), expression(D[2]), "misclassified"),
  pt.bg = c(pal, band),
  col = c("white", "white", "grey60"),
  pch = c(21, 21, 22),
  pt.cex = 1.3
)

par(op)
if (to_file) {
  dev.off()
}
```

#### Listing 24

Code to loop through the algorithmic steps.

::: {.callout-note}
The code can take several seconds to complete.
:::

```{r listing-24}
# Hyperparameter setup
treehypar <- rpart.control(maxdepth = 1)

# Number of trees
T <- 1000

# Learning rate
eta <- 1

# Lists for models and weights
fit <- vector("list", T)

# Initial observation weights
D <- rep(1, m) / m

# Loop through steps
for (t in seq_len(T)) {
  h <- rpart(
    replicate ~ .,
    data = train[, voinms],
    method = "class",
    weights = D,
    control = treehypar
  )
  yretro <- predict(object = h, newdata = train[, prednms], type = "class")
  e <- sum(D * (yretro != y_train))
  a <- 0.5 * log((1 - e) / e) * eta
  chi <- (yretro == y_train) * 1 + (yretro != y_train) * -1
  D_raw <- (D * exp(-a * chi))
  D <- D_raw / sum(D_raw)
  fit[[t]] <- list("h" = h, "a" = a)
}
names(fit) <- paste0("t", seq_len(T))

# Show first model weight
cat("Model weight 1:", fit$t1$a)
```

#### Listing 25

Code to generate ensemble retrodictions.

```{r listing-25}
# Boosted classification stumps
h <- lapply(fit, "[[", "h")

# Model weights
a <- sapply(fit, "[[", "a")

# Retrodictions from stumps
yretro_stumps <- sapply(h, predict, newdata = train[, prednms], type = "vector")

# Transformed values (-1,1)
yretro_stumps <- 2 * yretro_stumps - 3

# Retrodictions from full model
yretro_ada <- sign(a %*% t(yretro_stumps))
```

#### Listing 26

Code to replace matrix multiplication with a loop.

```{r listing-26}
#for (m in seq_len(m)) {
#yretro_ada[m] <- sign(sum(a * yretro_stumps[m, ]))
#}
```

#### Figure 9

Visualization of the decision boundary of 500 adaptively boosted shallow classification trees. The trees were fitted to the training data using two reviewer metrics: the power of the original study and the number of observations. The x-axis is the power of the original study effect. The y-axis is the number of observations. Note that the y-axis is not the outcome variable, but another study characteristic. The model predicts that any point in the gray region is a successful replication, while anything in the blue region is considered a failure. Gray triangles represent replications of study effects that actually succeeded (i.e., according to the replication metric), while blue circles represent replications that actually failed. Thus, gray triangles in the blue region are misclassifications, and vice versa.

```{r figure-9}
# Tree hyperparameter setting
treehypar <- rpart::rpart.control(
  # Slightly different hyperparameters to improve visibility of boundary
  maxdepth = 2,
  minsplit = 5,
  minbucket = round(5 / 3),
  cp = 1e-2,
  split = "gini"
)

# Fit AdaBoost on the two features
fit <- trainAda(
  replicate ~ .,
  data = train[, c("replicate", "power.o", "n.o")],
  T = 5e2,
  eta = 1,
  treehypar = treehypar,
  verbose = FALSE
)

# Both decision-boundary plots span the full text width: at column width the
# studies collide into a single band and the boundary stops reading as a line
if (to_file) {
  pdf(
    file.path(figdir, "boundaryada.pdf"),
    width = full_w,
    height = full_h,
    pointsize = 9
  )
}

# Visualize the ensemble's decision boundary
plot_adabound(
  fit,
  data = train,
  shade = "margin",
  xlab = "Power (original study)",
  ylab = "Sample size (original study)"
)

if (to_file) {
  dev.off()
}
```

#### Listing 27

Code to evaluate the retrodictive performance of the ensemble.

```{r listing-27}
# Confusion matrix
table(y_train, yretro_ada) # Perfect retrodictions
```

#### Listing 28

Code to generate ensemble predictions.

```{r listing-28}
# Outcome variable (test set)
y_test <- test[, outnme]

# Predictions (test set)
ypred_stumps <- sapply(h, predict, newdata = test[, prednms], type = "vector")
ypred_stumps <- 2 * ypred_stumps - 3
ypred_ada <- sign(a %*% t(ypred_stumps))
```

#### Listing 29

Code to evaluate the predictive performance of the ensemble.

```{r listing-29}
# Confusion matrix
table(y_test, ypred_ada) # Inperfect predictions
```

# Module 4: Leave-Project-Out Cross-Validation and Hyperparameter Optimization

## Two Wrapper Functions

#### Listing 30

Code introducing the wrapper functions.

::: {.callout-note}
We reduced the number of trees to speed up runtime in the upcoming sections.
:::

```{r listing-30}
# Boosted stumps and weights
fit <- trainAda(
  replicate ~ .,
  data = train[, voinms],
  # Boosting hyperparameter
  T = 100,
  eta = 1,
  # Tree hyperparameter
  maxdepth = 1
)

# Predictions on test set (class labels, for the confusion matrix)
ypred <- testAda(fit, data = test[, prednms], type = "class")

# Predictions on test set (margins, for the AUC)
ymargin <- testAda(fit, data = test[, prednms], type = "margin")
```

::: {.callout-important}
## `type = "class"` gives balanced accuracy, not AUC

A two-valued score collapses the ROC curve to a single interior point, whose
area is exactly (sensitivity + specificity) / 2. Use `type = "margin"` for the
AUC; keep `type = "class"` for the confusion matrix.
:::

## NEW: Feature Importance
Note: We should describe the package function before module 4 and maybe move the feature importance before hyperparameter optimization. (before we touch performance)

TODO: Describe better.
So far, we restricted AdaBoost to work with decision stumps. From Module 2 we know, that stumps are trees of depth one that split (at most - recall: cp = 0.01) once on a variable. Recall also, that each stump carries a voting weight (`a`) based on its accuracy. The (relative) feature imporatance score we introduce below is to group stumps by the variable they split on, extract the (adaboost) weight, and sum their given weights. Why does it work? Well, AdaBoost sequentially targets the hardest-to-predict observations by re-weighting data points, so a variable that is frequently chosen, especially by high-weight stumps, is one that consistently helped correct the ensemble's "biggest mistakes". The total
accumulated weight is thus a direct measure of a feature's predictive contribution or value.

That is the whole idea with nothing borrowed: count up the voting weight of the
stumps that chose each variable.

Start with initializing a named vector of zeros (to keep the loop readble) gorwing it in Line -- `imp[v] <- imp[v] + fit[[t]]$a` starting from an empty vector. If it does not split the addition step is skipped, so it silently produces nothing.

For stumps we can find the split directly: `frame` is the tree's node table.
It's first row is our root, so `frame$var[1]` is the variable the stump splits on 
If it does not split `rpart` writes `<leaf>`.

```{r importance-frame}
n_preds <- length(prednms)
# initialization: vector of zeros
voz <- numeric(n_preds)
imp <- setNames(voz, prednms)
for (t in seq_along(fit)) {
  # variable at the root (stumps)
  v <- as.character(fit[[t]]$h$frame$var[1])
  # if it does not split, skip it
  if (v == "<leaf>") {
    next
  }
  imp[v] <- imp[v] + fit[[t]]$a
}
# relative importance: normalize and scale
imp <- imp / sum(imp) * 100
# highest importance first
sort(imp, decreasing = TRUE)
```

version with rpart helper function

```{r importance-byhand}
# One slot per predictor, all starting at zero
n_preds <- length(prednms)
# initialization: vector of zeros
voz <- numeric(n_preds)
imp <- setNames(voz, prednms)
# walk ensemble, tree by tree
for (t in seq_along(fit)) {
  # Shorcut: extract rpart's included measure
  vi <- fit[[t]][["h"]][["variable.importance"]]
  # if it does not split, skip it
  if (is.null(vi)) {
    next
  }
  # stumps split (at most once), i.e. whole model weight goes to that variable
  imp[names(vi)] <- imp[names(vi)] + fit[[t]]$a
}
# Relative importance
imp <- 100 * imp / sum(imp)
# Most important first
sort(imp, decreasing = TRUE)
```

Note to self: Because in rpart cp is 0.01, by default, the trees do not need to split at all. 
In fact, twelve of the 100 stumps decided not to, which is why `next` exists. 

Note to self: cp and pruning maybe as box?

Outlook? If we grow deeper trees they can split more than once, and the weight is shared between those variables. Specifically, the weight is shared in proportion to how much each split improved the overall fit. The difference comes down to a one-line difference:

version with rpart helper function, general case
```{r importance-general}
fit_deep <- trainAda(
  replicate ~ .,
  data = train[, voinms],
  # Boosting hyperparameter
  T = 10,
  eta = 1,
  # Tree hyperparameter
  maxdepth = 4
)

n_preds <- length(prednms)
voz <- numeric(n_preds)
imp <- setNames(voz, prednms)
t <- 1
for (t in seq_along(fit_deep)) {
  vi <- fit_deep[[t]][["h"]][["variable.importance"]]
  if (is.null(vi)) {
    next
  }
  # Split the weight across the variables this tree used, in proportion to
  # each split's improvement
  share <- fit_deep[[t]]$a * vi / sum(vi)
  imp[names(share)] <- imp[names(share)] + share
}
imp <- 100 * imp / sum(imp)
sort(imp, decreasing = TRUE)
```

If we only have a stump, then `vi` holds a single number, so `vi / sum(vi)` is always 1. So we can the general version also for stumps.

In adaboost's gauge we mainly use the `rpart` helper function `variable.importance`. 
That is what `gauge()` does, so the two agree. 

```{r}
#| label: importance
important <- gauge(fit)
head(important, 10)

# The package function and the loop above compute the same thing
all.equal(
  setNames(important$importance, important$variable)
)
```

We also provide a quick and easy way to visualize the results.

```{r}
#| label: importance-plot
#| fig-alt: "Horizontal bar chart of AdaBoost feature importance"
fit |> gauge() |> plot()
plot(important, top_n = 15)
```


**Important note** The provided scores are relative (to eachother), so we believe it is more accurate to say, that they describe how the ensemble divided its attention rather than making a satement
on the net wort of a variable. They are also not "signed", so a variable may rand  

attention rather than what any one variable is worth on its own. They carry no
sign: a variable can rank highly because high values predict replication success
or because they predict failure, and this does not distinguish the two.
Correlated predictors split their share instead of each showing the full effect.
And the ranking belongs to *this* fit on *this* training data -- it moves with
the hyperparameters, since deeper trees spread weight over more variables by
construction.
:::

For more see Molnar & Pargent.

TODO: Can we bootstrap the importance scores to get a sense of their stability? Or havent we done that already with the LPO-CV?

## Evaluating Model Performance

#### Listing 31

Code showing different measures to evaluate the predictive performance of the
ensemble.

```{r listing-31}
# Confusion matrix
confusion <- table(ypred, y_test)

# Accuracy
sum(diag(confusion)) / sum(confusion)

# Area under the curve (scored on the margin)
auroc(y_test, ymargin)

# Base rates
df$replicate_01 <- as.integer(df$replicate) - 1
baserates <- aggregate(replicate_01 ~ pid, df, mean)

# Average base rate
mean(baserates$replicate_01)
```

::: {.callout-important}
## Switched to `PRROC` package
Cite and fix everywhere - especially in the tutorial.
:::

::: {.callout-note}
`auroc()` is a thin wrapper around `PRROC::roc.curve()` (Grau, Grosse &
Keilwagen). Specifically, `auprc()` wraps `pr.curve()`. The wrapper exists so the outcome can be passed in whatever form you have it, for instance as a factor, 0/1, -1/1 or logical. `roc.curve()` itself expects a 0/1 vector and returns a "wrong" number without complaining if you feed it a 1/2.
:::







## Leave-Project-Out Cross-Validation

Visualization of the leave-project-out cross-validation (LPO-CV) process. Reading the graph from left to right, the data are first partitioned into the Many Labs 1 (ML 1) project, the Replication Project Psychology (RPP), the Many Labs 3 (ML 3) project, the Experimental Economics Replication Project (EERP), and the Social Science Replication Project (SSRP). Then, in Iteration 1, the ML 1 project is used as the test set. The other data are used for model training. After fitting the model to the training data, its performance is evaluated on the hold-out data from the Many Labs 1 project. This is repeated for the other replication projects (Iterations 2--5), ultimately yielding the distribution of performance estimates across projects. A measure such as the mean can then be used to summarize the distribution with a point estimate. Note, however, that the distribution provides the richer picture of the validation process.

This diagram is drawn in the manuscript rather than produced from R, so its
LaTeX source is kept here. The block is not executed.

```tex
\begin{tikzpicture}[
    every node/.style={font=\normalsize},
]

% Styles
\tikzset{
    baseblock/.style={
        draw,
        minimum height=1.1cm,
        minimum width=1.9cm,
        text=black,
        font=\scriptsize
    },
    datanode/.style={
        baseblock,
        fill={backcolour},
        align=center,
        minimum height=5.9cm
    },
    testnode/.style={
        baseblock,
        fill={rgb,255:red,68; green,1; blue,84},
        text=white,
    },
    trainnode/.style={
        baseblock,
        fill={rgb,255:red,217; green,204; blue,220}
    },
    performnode/.style={
        baseblock,
        fill={rgb,255:red,33; green,144; blue,140},
        text=white,
        align=center,
    },
    iterationblock/.style={
        baseblock,
        fill={backcolour},
    },
    averageblock/.style={
        baseblock,
        fill={rgb,255:red,33; green,144; blue,140},
        text=white,
        align=center,
        minimum width=9.9cm,
        minimum height=1.1cm,
    }
}

% Left labels
\node[datanode] (data) at (-2, -2.4) {Data};
\node[iterationblock] (ml1)   at (0,  0) {ML1};
\node[iterationblock] (rpp)   at (0, -1.2) {RPP};
\node[iterationblock] (ml3)   at (0, -2.4) {ML3};
\node[iterationblock] (eerp)  at (0, -3.6) {EERP};
\node[iterationblock] (ssrp)  at (0, -4.8) {SSRP};

% Iteration headers
\node[iterationblock] (it1) at (2, 1.2) {Iteration 1};
\node[iterationblock] (it2) at (4, 1.2) {Iteration 2};
\node[iterationblock] (it3) at (6, 1.2) {Iteration 3};
\node[iterationblock] (it4) at (8, 1.2) {Iteration 4};
\node[iterationblock] (it5) at (10, 1.2) {Iteration 5};

% Iteration 1
\node[testnode]      (a) at (2, 0)   {Test};
\node[trainnode]     (b) at (2, -1.2) {Training};
\node[trainnode]     (c) at (2, -2.4) {Training};
\node[trainnode]     (d) at (2, -3.6) {Training};
\node[trainnode]     (e) at (2, -4.8) {Training};
\node[performnode]   (p1) at (2, -6) {\shortstack{Performance\\ML1}};

% Iteration 2
\node[trainnode]     (f) at (4, 0)   {Training};
\node[testnode]      (g) at (4, -1.2) {Test};
\node[trainnode]     (h) at (4, -2.4) {Training};
\node[trainnode]     (i) at (4, -3.6) {Training};
\node[trainnode]     (j) at (4, -4.8) {Training};
\node[performnode]   (p2) at (4, -6) {\shortstack{Performance\\RPP}};

% Iteration 3
\node[trainnode]     (k) at (6, 0)   {Training};
\node[trainnode]     (l) at (6, -1.2) {Training};
\node[testnode]      (m) at (6, -2.4) {Test};
\node[trainnode]     (n) at (6, -3.6) {Training};
\node[trainnode]     (o) at (6, -4.8) {Training};
\node[performnode]   (p3) at (6, -6) {\shortstack{Performance\\ML3}};

% Iteration 4
\node[trainnode]     (q) at (8, 0)   {Training};
\node[trainnode]     (r) at (8, -1.2) {Training};
\node[trainnode]     (s) at (8, -2.4) {Training};
\node[testnode]      (t) at (8, -3.6) {Test};
\node[trainnode]     (u) at (8, -4.8) {Training};
\node[performnode]   (p4) at (8, -6) {\shortstack{Performance\\EERP}};

% Iteration 5
\node[trainnode]     (v) at (10, 0)   {Training};
\node[trainnode]     (w) at (10, -1.2) {Training};
\node[trainnode]     (x) at (10, -2.4) {Training};
\node[trainnode]     (y) at (10, -3.6) {Training};
\node[testnode]      (z) at (10, -4.8) {Test};
\node[performnode]   (p5) at (10, -6) {\shortstack{Performance\\SSRP}};

% Average performance block
\node[averageblock]  (avg) at (6, -7.2) {Average performance};

\end{tikzpicture}
```

#### Listing 32

Code to do a leave-project-out cross-validation.

::: {.callout-note}
This ran for approximately 2 minutes on our laptops.
:::

```{r listing-32}
# Partition data by project ID
df_split <- split(df, df$pid)

# List for AUC results
auc <- list()

runtime <- system.time({
  # LPO-CV
  for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
    message("\nThe ", paste0(rp), " model.\n")
    # Training and test set
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]
    # Fit model to data
    fit <- trainAda(
      replicate ~ .,
      data = train[, voinms],
      T = 1000,
      eta = 1,
      treehypar = treehypar
    )
    # Predictions (left-out project), on the margin so the AUC sees the ranking
    ypred <- testAda(fit, data = test[, prednms], type = "margin")
    # Performance evaluation
    auc[[rp]] <- auroc(test$replicate, ypred)
  }
})

# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")

# Results
# TODO: From mean to plot?
c(unlist(auc), "mean" = mean(unlist(auc)))
```

## Hyperparameter Optimization

#### Listing 33

Code to define a hypergrid.

```{r listing-33}
# Define hypergrid
hgrid <- expand.grid(
  # Tree-specific:
  # Tree depth
  "depth" = 1:3,
  # Boosting specific:
  # Number of trees
  "T" = c(10, 100, 1000),
  # Learning rate
  "eta" = c(1, 0.1),
  # Container for fold performance
  "eerp" = NA,
  "ml1" = NA,
  "ml3" = NA,
  "rpp" = NA,
  "ssrp" = NA,
  "auc.mean" = NA,
  "runtime" = NA
)
```

#### Listing 34

Code to do hyperparameter optimization using a leave-project-out
cross-validation strategy.

::: {.callout-note}
This ran for approximately 3 minutes on our laptops.
:::

```{r listing-34}
# Number of rows
hgrid_rows <- nrow(hgrid)

# Partition data by project ID
df_split <- split(df, df$pid)

# Hyperparameter optimization
for (i in seq_len(hgrid_rows)) {
  # Iteration time
  runtime <- system.time({
    # Leave-project-out CV
    for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
      # Progress report
      cat(
        "\rProgress: ",
        round(i / hgrid_rows * 100),
        "%",
        " | T: ",
        hgrid$T[i],
        " | eta: ",
        hgrid$eta[i],
        " | depth: ",
        hgrid$depth[i],
        " | model: ",
        rp
      )
      # Training and test set
      test <- df_split[[rp]]
      train <- df[!(df$pid %in% rp), ]
      # Fit model to data
      fit <- trainAda(
        replicate ~ .,
        data = train[, voinms],
        T = hgrid$T[i],
        eta = hgrid$eta[i],
        treehypar = rpart.control(maxdepth = hgrid$depth[i]),
        verbose = FALSE
      )
      # Predictions (on the margin, so the AUC sees the ranking)
      ypred <- testAda(
        fit,
        data = test[, prednms],
        type = "margin",
        verbose = FALSE
      )
      # Performance evaluation
      hgrid[[rp]][i] <- auroc(test$replicate, ypred)
    }
    # Average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i, 4:8]))
  })
  hgrid$runtime[i] <- runtime[["elapsed"]]
}

# Hypergrid sorted by average AUC
hgrid[order(-hgrid$auc.mean), ]
```

# NEW

Listing 34 searches 18 settings and keep the result in columns named after the
projects. The package ships the full search as `hypergrid`: 
960 settings, each scored on all five projects under 15 different measures.

`lpocv()` returns that as an array rather than a table, because I think in 
setting x project x metric. So you can easily slice it without a reshape.

```{r hypergrid-array}
data(hypergrid)

# setting x project x metric
dim(hypergrid$estimates)
names(dimnames(hypergrid$estimates))

# `grid` says what each setting is. `cp`, the base learners' complexity
# parameter, is a column here like any other hyperparameter -- which is why 480
# combinations become 960 settings.
head(hypergrid$grid, 3)
hypergrid$projects
```

One measure across every setting and project: 960 x 5 matrix, and the grid
telly us which row is which:

```{r hypergrid-slice}
au <- hypergrid$estimates[,, "auroc"]
g <- hypergrid$grid

# One setting, sorted across projects
i <- which(g$depth == 1 & g$T == 1000 & g$eta == 0.15 & g$cp == 0.01)
sort(au[i, ], decreasing = TRUE)

# The best each project can be made to look, over the depth-1 stumps
d1 <- g$depth == 1 & g$cp == 0.01
apply(au[d1, ], 2, max)

# Rank the settings by their average over the projects
m <- rowMeans(au)
cbind(g, mean = round(m, 3))[order(-m), ][1:5, ]
```

An `NA` here means "undefined" rather than "failed": `ppv` has no value when a model predicts no positives at all, which a ten-tree ensemble does on the smaller projects. 

::: {.callout-important}
## The mean over projects ranks settings does not estimate performance

The last slice averages five folds that hold 10, 13, 18, 21 and 90 studies and
share most of their training data, so it hides both the imbalance and the
dependence. TODO
:::


In small test set you can have the problem of bad luck and up (especially with small test sets) with a handy partition which is too easy to classify and (thus not representative of what would usually happen, i.e. performane overestimate).


TODO: Random grid search. Define 20 as a number in the random search

Two things follow. First, the five inner searches may each select a different
setting, and the setting that won in Listing 34 may be chosen by none of them.
That is expected rather than alarming: with five projects there is no stable
"best" setting to find, and seeing that instability directly is more informative
than the average that hides it. Second, the nested estimate is *conservative*
here, because each nested model saw four projects while the model you ship sees
five.

If nesting does not hand you a model, what does? The three listings have three
different jobs, and it is worth being explicit about which is which:

| job | listing |
|:----|:--------|
| Choose the hyperparameters, using all five projects | Listing 34 |
| Estimate what that choosing procedure is worth | Listing 35 |
| Fit the model you will actually use | Listing 36 |

So Listing 34 still picks the setting and Listing 36 still fits it on the full
dataset -- nesting changes neither. What it changes is the number you quote
alongside that model: not the winning average from Listing 34, which the same
five projects produced by selecting it, but the nested estimate, which no project
had a hand in choosing.

The settings the inner searches picked are worth looking at, but they are a
diagnostic rather than a shortlist. They have the same standing as the
coefficients fitted inside each cross-validation fold back in Module 1: nobody
averages those, and nobody asks which fold's coefficients to keep. They are
by-products of a procedure being measured, not candidates for shipping.

::: {.callout-tip}
## Set the depth with `maxdepth`, not with `treehypar`

`rpart.control(maxdepth = 3)` returns *every* `rpart` default, not just the depth
you named -- including `cp = 0.01`, `xval = 10` and `maxsurrogate = 5`. Passed
through `treehypar`, those overwrite the settings `trainAda()` chose for
boosting. `trainAda()` keeps `cp` at `rpart`'s own 0.01 precisely so that the two
forms fit the *same* trees: we checked, and the trees, the model weights and the
margins are identical. What still differs is `xval = 10`, ten internal
cross-validations per weak learner that nothing reads, costing about thirty per
cent of the runtime.

So this is a tidiness point, not a correctness one. Use the `maxdepth` argument
for depth, and reserve `treehypar` for controls you genuinely mean to change --
`minsplit` and `minbucket` are the ones that actually change how the trees grow.
:::

#### Listing 35

Code to do a nested leave-project-out cross-validation.

::: {.callout-note}
This ran for approximately 10 minutes on our laptops.
:::

```{r listing-35}
# Projects, in the order they take their turn
projects <- c("eerp", "ml1", "ml3", "rpp", "ssrp")

# Partition data by project ID
df_split <- split(df, df$pid)

# One honest score and one selected setting per project
nested_auc <- list()
selected <- list()

runtime <- system.time({
  # OUTER loop: evaluates
  for (rp in projects) {
    # Progress report
    message("\nOuter fold: ", rp, "\n")

    # Training and test set. The test project leaves the data entirely, so
    # nothing below can see it.
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]

    # INNER loop: selects. This is the same grid search as Listing 34, run on
    # the training projects only -- so we can hand it to lpocv().
    inner <- lpocv(
      replicate ~ power.o + effect_size.o + n.o + p_value.o,
      data = train,
      group = "pid",
      grid = hgrid[, c("depth", "T", "eta")]
    )

    # Rank the settings by their mean over the inner folds. Here the mean is a
    # ranking device, not a performance estimate.
    inner_auc <- summary(inner, metric = "auroc")
    ranked <- aggregate(estimate ~ setting, data = inner_auc, FUN = mean)

    # The setting the four training projects chose, on their own
    best <- hgrid[ranked$setting[which.max(ranked$estimate)], ]
    selected[[rp]] <- best[, c("depth", "T", "eta")]

    # Refit at that setting and score the untouched project
    fit <- trainAda(
      replicate ~ .,
      data = train[, voinms],
      T = best$T,
      eta = best$eta,
      maxdepth = best$depth,
      verbose = FALSE
    )
    ypred <- testAda(
      fit,
      data = test[, prednms],
      type = "margin",
      verbose = FALSE
    )
    nested_auc[[rp]] <- auroc(test$replicate, ypred)
  }
})

# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")

# The honest expectation, per project
unlist(nested_auc)

# What each outer fold selected, and how often they disagree
do.call(rbind, selected)
```

Only the outer loop is new. The inner one is the leave-project-out search we
already ran in Listing 34, so we hand it to `lpocv()` rather than write it out
again -- the single change is that it now sees `train` instead of `df`.

The whole procedure is also one call, which is what we use from here on:

```{r listing-35-oneline}
nested <- lpocv(
  replicate ~ power.o + effect_size.o + n.o + p_value.o,
  data = df,
  group = "pid",
  grid = hgrid[, c("depth", "T", "eta")],
  nested = TRUE,
  criterion = "auroc"
)
nested
```

## Finalizing an Ensemble

#### Listing 36

Code to fit the final model to the data.

```{r listing-36}
# Fit optimized model
fit_final <- trainAda(
  replicate ~ .,
  data = df[, voinms],
  T = 10,
  eta = 1,
  treehypar = rpart.control(maxdepth = 1)
)
```
