---
title: "Code Along Script For The Tutorial"
execute:
  eval: false
vignette: >
  %\VignetteIndexEntry{Code Along Script For The Tutorial}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

# Prerequisites

#### Listing 1: Code to install the adatutor package form GitHub, load and attach it.

```{r}
# Load package
library(remotes)

# Install package from GitHub
install_github("sbissantz/adatutor")

# Load package
library(adatutor)
```

# Empirical Example

#### Listing 2: Code to load the dataset.

```{r}
# Load data
data(altmejd)
```

#### Listing 3: Code to prepare the data.

```{r}
# Reviewer metrics (names)
prednms <- c("power.o", "effect_size.o", "n.o", "p_value.o")

# Outcome variable (name)
outnme <- "replicate"

# Variables of interest
voinms <- c(prednms, outnme)

# Shorten 'altmejd'
df <- altmejd
```

# Module 1: Data Splitting

#### Listing 4: Code to do a 70/30 simple random split.

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.7

# Number of effects
n_fx <- nrow(df)

# Number of training samples
n_train <- round(n_fx * prop)

# Training index
train_idx <- sample(seq_len(n_fx), n_train)

# Training set
train <- df[train_idx, ]

# Test set (incl. validation)
testvalid <- df[-train_idx, ]
```

#### Listing 5: Code to complete a 70/15/15 simple random split

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.5

# Number of effects
n_fx <- nrow(testvalid)

# Number of validation samples
n_valid <- round(n_fx * prop)

# Training index
valid_idx <- sample(seq_len(n_fx), n_valid)

# Validation set
valid <- testvalid[valid_idx, ]

# Test set
test <- testvalid[-valid_idx, ]
```

#### Listing 6: Code to check for imbalances in the outcome variable of the original dataset and the test set.

```{r}
# Frequency tables
tbl_orig <- table(df$replicate)
tbl_testvalid <- table(testvalid$replicate)

# Relative frequencies
proportions(tbl_orig)
proportions(tbl_testvalid)
```

#### Listing 7: Code to do a 70/30 stratified split.

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.7

# Partition data based on outcome
set_split <- split(df, df$replicate)

# Number of validation samples
n_train <- lapply(set_split, function(set) round(nrow(set) * prop))

# Training index
strat_idx <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_train
)

# Training index
train_idx <- df$eid %in% unlist(strat_idx)

# Training set
train <- df[train_idx, ]

# Test set (incl. validation)
testvalid <- df[!train_idx, ]
```

#### Listing 8: Code to complete a 70/15/15 stratified split

```{r}
# For reproducibility
set.seed(112)

# Split proportion
prop <- 0.5

# Partition data based on outcome
set_split <- split(testvalid, testvalid$replicate)

# Numbers of validation samples
n_valid <- lapply(set_split, function(set) round(nrow(set) * prop))

# Validation index
strat_idx <- mapply(
  function(set, n) sample(set[, "eid"], n),
  set_split,
  n_valid
)
valid_idx <- testvalid$eid %in% unlist(strat_idx)

# Validation set
valid <- testvalid[valid_idx, ]

# Test set
test <- testvalid[!valid_idx, ]
```

#### Listing 9: Code to do a 70/30 random and stratified split, with the pre-implemented functions of the adatutor package.

```{r}
# For reproducibility
set.seed(112)

# Simple random split
randsplit <- random_split(df, prop = 0.7)
train <- randsplit$set1
testvalid <- randsplit$set2

# Stratified split
stratsplit <- stratified_split(df, group = "replicate", id = "eid", prop = 0.7)
train <- stratsplit$set1
testvalid <- stratsplit$set2
```

#### Listing 10: Code to do a 70/30 random and stratified split, with the rsample package.

```{r}
# Load package
library(rsample)

# For reproducibility
set.seed(112)

# Simple random split
randsplit <- initial_split(df, prop = 0.7)
train <- training(randsplit)
testvalid <- testing(randsplit)

# Stratified split
stratsplit <- initial_split(df, prop = 0.7, strata = "replicate")
train <- training(stratsplit)
testvalid <- testing(stratsplit)
```

#### Figure 2

```{r}
# Select only two predictors
# Note: Reduce run-time - decision boundary visualization
subfeat <- c("power.o", "n.o")
subdf <- train[, union(subfeat, outnme)]
# Hyperparameter setup
treehypar <- rpart::rpart.control(
  # Maximum depth of any node tree
  maxdepth = 1,
  # Minimum n° observations in terminal node
  minbucket = 4,
  # Original CART with gini impurity as splitting criterion
  split = "gini"
) # or ID3: split = "information"
# Create a decision tree
h <- rpart::rpart(
  replicate ~ .,
  data = subdf,
  method = "class",
  control = treehypar
)
# Visualize
rpart.plot::rpart.plot(h, extra = 102, digits = 4, box.palette = "BuGy")
```

::: {.callout-important}
## Important note

We suggest running Listings 7 & 8 again, before proceeding with the other
modules. This should guarantee the same training, validation, and test set.
:::

# Module 2: Classification Stumps

#### Listing 11: Code to retrieve the frequency tables for the outcome variable under the two power conditions.

```{r}
# Condition for the split
splitcond <- train$power.o < 0.9305

# Studies with power.o < 0.9305
table(train$replicate[splitcond])

# Studies with power.o >= 0.9305
table(train$replicate[!splitcond])
```

#### Listing 12: Code to set the hyperparameters of a classification tree and fit the model to the training data.

```{r}
# Load package
library(rpart)

# Hyperparameter setup
treehypar <- rpart.control(
  # Maximum depth
  maxdepth = 1,
  # Gini impurity (default)
  split = "gini"
)

# Fit classification stump
cstump <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  control = treehypar
)
```

#### Listing 13: Code to visualize a fitted classification stump.

```{r}
# Load package
library(rpart.plot)

# Visualize classification stump
rpart.plot::rpart.plot(cstump, extra = 102, digits = 4, box.palette = "BuGy")
```

#### Figure 3

```{r}
# Outcome variable
y_train <- train[, outnme]
# Number of classes
k <- length(unique(subdf$replicate))
# Range(s) as list
rng <- lapply(subdf[, subfeat], range, na.rm = TRUE)
# Grid resolution
resolution <- 1000
# Colors
col1 <- "cornflowerblue"
col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f)
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Predict grid
ypred <- predict(h, grid, type = "class")
# Contour matrix
z_mat <- matrix(as.integer(ypred), nrow = resolution)
# "0" means "not replicate", "1" means "replicate"
symbol_colors <- ifelse(y_train == "failure", col1, col2)
# "0" means "not replicate", "1" means "replicate"
grid_colors <- ifelse(ypred == "failure", col1_transp, col2_transp)
plot(
  NULL,
  xlim = range(x1_seq),
  ylim = range(x2_seq),
  xlab = "Power (original study)",
  ylab = "Sample size (original study)"
)
points(grid, col = grid_colors, pch = ".")
points(subdf[, subfeat], col = symbol_colors, pch = as.integer(y_train) + 15)
contour(
  x1_seq,
  x2_seq,
  z_mat,
  add = TRUE,
  drawlabels = FALSE,
  lwd = 2,
  col = "black",
  levels = (1:(k - 1)) + .5
)
legend(
  "topleft",
  c("success", "failure"),
  pch = c(17, 16),
  col = c(col2, col1),
  cex = 0.7
)
```

#### Figure 4

```{r}
# Calculate Gini Impurity
x <- seq(0.01, 1, by = 0.01) # Values for fraction of class k
y <- 1 - (x^2) - (1 - x)^2 # Gini Impurity calculation
# Create the plot
plot(
  x,
  y,
  type = "l",
  lwd = 2, # Specify line plot
  xlab = "Proportion of Successes",
  ylab = "Impurity"
)
mtext("Two Class Classification Problem")
# Customize axes
axis(side = 1, at = seq(0, 1, by = 0.1))
```

#### Listing 14: Code to calculate means of neighboring units and find the position of the relevant power value.

```{r}
# Sorted vector
power_srt <- sort(train$power.o)

# Adjacent means
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2

# Find position of value '0.9305'
which(round(adjmeans, 4) == 0.9305)

# Additional

adj_pat <- which(round(adjmeans, 4) == 0.9305)
const <- adjmeans[adj_pat]
N_total <- nrow(train) # 107
criterion_left <- train$power.o < const
N_left <- length(train$replicate[criterion_left]) # 66 total
left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1) # 17
left_failure <- N_left - left_success # 49 failures
# Gini impuritiy (left)
gini_left <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
gini_left # 0.382
criterion_right <- train$power.o >= const
N_right <- length(train$replicate[criterion_right]) # 41
right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1) # 31
right_failure <- N_right - right_success # 10
# Gini impuritiy (right)
gini_right <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
gini_right # 0.3688281
# Overall weighted Gini impurity
gini_impurity <- (N_left / N_total) *
  gini_left +
  (N_right / N_total) * gini_right
round(gini_impurity, digits = 5) # 0.37724
# Sorted values
power_srt <- sort(train$power.o)
adjmeans <- (power_srt[-1] + power_srt[-length(power_srt)]) / 2
gini_impurity <- vector(length = length(adjmeans))
for (i in seq_along(adjmeans)) {
  constant <- adjmeans[i]
  N_total <- length(y_train)
  # Left leaf
  criterion_left <- train$power.o < constant
  N_left <- length(train$replicate[criterion_left])
  left_success <- sum(as.numeric(train$replicate[criterion_left]) - 1)
  left_failure <- N_left - left_success
  # Right leaf
  criterion_right <- train$power.o >= constant
  N_right <- length(train$replicate[criterion_right])
  right_success <- sum(as.numeric(train$replicate[criterion_right]) - 1)
  right_failure <- N_right - right_success
  # Gini index
  gini_d1 <- 1 - ((left_failure / N_left)^2 + (left_success / N_left)^2)
  gini_d2 <- 1 - ((right_failure / N_right)^2 + (right_success / N_right)^2)
  # Impurity
  gini_impurity[i] <- (N_left / N_total) *
    gini_d1 +
    (N_right / N_total) * gini_d2
}
# Position of smallest impurity
which.min(gini_impurity)
# Outcome
adjmeans[which.min(gini_impurity)]
```

#### Listing 15: Code to make the assumption of equal weights explicit.

```{r}
# Load package
library(rpart.plot)

# Number of rows
m <- nrow(train)

# Equal weight to each observation
D <- rep(1, m) / m

# Fit decision stump
cstump <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weight = D,
  control = treehypar
)

# Visualize
rpart.plot(cstump, digits = 4, box.palette = "BuGy")
```

#### Listing 16: Code to show the sensitivity to observation weights by using normalized random weights from a uniform distribution.

```{r}
# For reproducibility
set.seed(112)

# Draws from standard uniform
D_raw <- runif(107)

# Normalize (sum to 1)
D <- D_raw / sum(D_raw)

# Create decision stump
cstump_ast <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weight = D,
  control = treehypar
)

# Visualize
rpart.plot(cstump_ast, extra = 100, digits = 4, box.palette = "BuGy")
```

# Module 3: AdaBoost

#### Listing 17: Code to understand the givens in the AdaBoost algorithm.

```{r}
# Predictors and outcome (names)
prednms
outnme

# Predictor matrix
x_train <- train[, prednms]

# Outcome vector
y_train <- train[, outnme]

# Indices
i <- 1
m <- nrow(train)

# i-th row or element
x_train[i, ]
y_train[i]

# i-th to m-th row (i.e., all)
cbind(x_train, y_train)[i:m, ]
```

#### Listing 18: Code to generate initialization weights in the AdaBoost algorithm.

```{r}
# Initialization weights
D1 <- rep(1, m) / m
```

#### Listing 19: Code to fit a classification stump with the initialization weights.

```{r}
# Hyperparameter setup
treehypar <- rpart.control(maxdepth = 1, split = "gini")

# Fit classification stump
h1 <- rpart(
  replicate ~ .,
  data = train[, voinms],
  method = "class",
  weights = D1,
  control = treehypar
)
```

#### Listing 20: Code to calculate the classification stump's error.

```{r}
# Retrodictions (training data)
yretro <- predict(h1, newdata = train[, voinms], type = "class")

# Retrodictive error
epsilon <- sum(D1 * (yretro != y_train))
```

#### Listing 21: Code to calculate the classification stump's performance.

```{r}
# Retrodivitve performance
perf <- sum(D1 * (yretro == y_train))

# Retrodictive error
epsilon <- 1 - perf
```

#### Listing 22: Code to calculate the classification stump's weight assuming a learning rate eta of 1.

```{r}
# Learning rate
eta <- 1

# Model weight
alpha1 <- 1 / 2 * log(perf / epsilon) * eta
```

#### Figure 6

```{r}
fat <- function(x, eta) 1 / 2 * log(x / (1 - x)) * eta
curve(
  fat(x, 1),
  from = 0.5,
  to = 1,
  lwd = 2,
  ylab = "Importance",
  xlab = "Performance"
)
grid()
curve(
  fat(x, 0.5),
  from = 0.5,
  to = 1,
  lty = 2,
  lwd = 2,
  col = "ivory4",
  add = TRUE
)
curve(
  fat(x, 0.1),
  from = 0.5,
  to = 1,
  lty = 4,
  lwd = 2,
  col = "steelblue",
  add = TRUE
)
legend(
  "topleft",
  legend = c(
    expression(eta * " = 1"),
    expression(eta * " = 0.5"),
    expression(eta * " = 0.1")
  ),
  col = c("black", "ivory4", "steelblue"),
  lty = c(1, 2, 4),
  lwd = 2
)
```

#### Listing 23: Code to calculate the update factor for the initialization weights.

```{r}
# Indicator trick
chi1 <- (yretro == y_train) * 1 + (yretro != y_train) * -1

# Unnormalized weights
D2_raw <- D1 * exp(-alpha1 * chi1)

# Normalization constant
Z1 <- sum(D2_raw)

# Normalized weights
D2 <- D2_raw / Z1
```

#### Figure 7

```{r}
feat <- function(at, x) exp(-at * x)
curve(
  feat(0.719, x),
  from = -1,
  to = 1,
  ylim = c(0.3, 2.2),
  lty = 3,
  xlab = "Classification",
  ylab = "Weight update factor",
  xaxt = "n"
)
grid()
points(c(1, -1), c(feat(0.719, 1), feat(0.7, -1)), pch = 20, cex = 2)
curve(
  feat(0.619, x),
  from = -1,
  to = 1,
  ylim = c(0, 2),
  col = "ivory4",
  lty = 3,
  add = TRUE
)
points(
  c(1, -1),
  c(feat(0.619, 1), feat(0.6, -1)),
  pch = 17,
  cex = 2,
  col = "ivory4"
)
curve(
  feat(0.519, x),
  from = -1,
  to = 1,
  ylim = c(0, 2),
  col = "steelblue",
  lty = 3,
  add = TRUE
)
points(
  c(1, -1),
  c(feat(0.519, 1), feat(0.51, -1)),
  pch = 15,
  cex = 2,
  col = "steelblue"
)
abline(h = 1, lty = 2, lwd = 1.5, col = "gray30")
legend(
  "topright",
  legend = c(
    expression(a[t] * " = 0.719"),
    expression(a[t] * " = 0.619"),
    expression(a[t] * " = 0.519")
  ),
  col = c("black", "ivory4", "steelblue"),
  lty = c(1, 1, 1),
  pch = c(20, 17, 15),
  lwd = 2
)
axis(1, at = c(-1, 1))
```

#### Figure 8

```{r}
n_25 <- 25
scaling <- 175
plot(
  c(1, n_25),
  c(0, D1[1] + 0.02),
  type = "n",
  xaxt = "n",
  xlab = "Data point",
  ylab = "Weight"
)
axis(1, at = 1:n_25, labels = paste(1:25), cex = 0.5)
for (i in 1:n_25) {
  lines(c(i, i), c(D1[i], D2[i]), lty = 3)
}

points(D1[1:n_25], pch = 20, col = "ivory4", cex = D1[1] * scaling)
points(
  D1[1:n_25],
  pch = 20,
  col = ifelse((y_train == yretro), "green", "red"),
  cex = D1[1] + 0.2
)
points(D2[1:n_25], pch = 20, col = "steelblue", cex = D2[1:n_25] * scaling)
legend(
  "topright",
  legend = c(expression(D[1]), expression(D[2])),
  col = c("ivory4", "steelblue"),
  pch = c(20, 20)
)
```

#### Listing 24: Code to loop through the algorithmic steps. Note that the code can take several seconds to complete.

```{r}
# Hyperparameter setup
treehypar <- rpart.control(maxdepth = 1)

# Number of trees
T <- 1000

# Learning rate
eta <- 1

# Lists for models and weights
fit <- vector("list", T)

# Initial observation weights
D <- rep(1, m) / m

# Loop through steps
for (t in seq_len(T)) {
  h <- rpart(
    replicate ~ .,
    data = train[, voinms],
    method = "class",
    weights = D,
    control = treehypar
  )
  yretro <- predict(object = h, newdata = train[, prednms], type = "class")
  e <- sum(D * (yretro != y_train))
  a <- 0.5 * log((1 - e) / e) * eta
  chi <- (yretro == y_train) * 1 + (yretro != y_train) * -1
  D_raw <- (D * exp(-a * chi))
  D <- D_raw / sum(D_raw)
  fit[[t]] <- list("h" = h, "a" = a)
}
names(fit) <- paste0("t", seq_len(T))

# Show first model weight
cat("Model weight 1:", fit$t1$a)
```

#### Listing 25: Code to generate ensemble retrodictions.

```{r}
# Boosted classification stumps
h <- lapply(fit, "[[", "h")

# Model weights
a <- sapply(fit, "[[", "a")

# Retrodictions from stumps
yretro_stumps <- sapply(h, predict, newdata = train[, prednms], type = "vector")

# Transformed values (-1,1)
yretro_stumps <- 2 * yretro_stumps - 3

# Retrodictions from full model
yretro_ada <- sign(a %*% t(yretro_stumps))
```

#### Listing 26: Code to replace matrix multiplication with a loop.

```{r}
#for (m in seq_len(m)) {
#yretro_ada[m] <- sign(sum(a * yretro_stumps[m, ]))
#}
```

#### Figure 9

```{r}
# Outcome vector
yobs <- train[, outnme]
# Make 0/1 factor into something ADA can handle (-1/1)
yobs_ml <- 2 * as.numeric(train[, outnme]) - 3
# Number of classes
k <- length(unique(yobs_ml))
# Range(s) as list
Xobs <- train[, c("power.o", "n.o")]
rng <- lapply(Xobs, range, na.rm = TRUE)
# Grid resolution
resolution <- 1e3
# Colors
col1 <- "cornflowerblue"
col2 <- "ivory4"
# Transparency
alpha.f <- 0.03
col1_transp <- adjustcolor(col1, alpha.f)
col2_transp <- adjustcolor(col2, alpha.f)
# First feature sequence
x1_seq <- seq(rng$power.o[1], rng$power.o[2], length.out = resolution)
# Second feature sequence
x2_seq <- seq(rng$n.o[1], rng$n.o[2], length.out = resolution)
# Make grid from sequences
grid <- expand.grid("power.o" = x1_seq, "n.o" = x2_seq)
# Make grid
# Tree hyperparameter setting
treehypar <- rpart::rpart.control(
  # Slightly different hyperparameters to improve visibility of boundary
  maxdepth = 2,
  minsplit = 5,
  minbucket = round(5 / 3),
  cp = 1e-2,
  split = "gini"
)
# Fit AdaBoost
fit <- trainAda(
  replicate ~ .,
  data = train[, c("replicate", "power.o", "n.o")],
  T = 5e2,
  eta = 1,
  treehypar = treehypar,
  verbose = FALSE,
  input_checks = FALSE
)
# Predict ADAboost
ypred_ml <- testAda(fit, grid, input_checks = FALSE, verbose = FALSE)
# Make prediction 0/1 instead of -1/1
ypred <- as.factor((ypred_ml + 1) / 2)
# Contour matrix
z_mat <- matrix(as.integer(ypred), nrow = resolution)
# "0" means "not replicate", "1" means "replicate"
symbol_colors <- ifelse(yobs == "failure", col1, col2)
grid_colors <- ifelse(ypred == 0, col1_transp, col2_transp)
plot(
  NULL,
  xlim = range(x1_seq),
  ylim = range(x2_seq),
  xlab = "Power (original study)",
  ylab = "Sample size (original study)"
)
points(grid, col = grid_colors, pch = ".")
z_mat <- matrix(as.integer(ypred), nrow = resolution)
# For the black line
contour(
  x1_seq,
  x2_seq,
  z_mat,
  add = TRUE,
  drawlabels = FALSE,
  lwd = 2,
  col = "black",
  levels = (1:(k - 1)) + .5
)
points(Xobs, col = symbol_colors, pch = as.integer(yobs) + 15)
legend(
  "topleft",
  c("success", "failure"),
  pch = c(17, 16),
  col = c(col2, col1),
  cex = 0.7
)
```

#### Listing 27: Code to evaluate the retrodictive performance of the ensemble.

```{r}
# Confusion matrix
table(y_train, yretro_ada) # Perfect retrodictions
```

#### Listing 28: Code to generate ensemble predictions.

```{r}
# Outcome variable (test set)
y_test <- test[, outnme]

# Predictions (test set)
ypred_stumps <- sapply(h, predict, newdata = test[, prednms], type = "vector")
ypred_stumps <- 2 * ypred_stumps - 3
ypred_ada <- sign(a %*% t(ypred_stumps))
```

#### Listing 29: Code to evaluate the predictive performance of the ensemble.

```{r}
# Confusion matrix
table(y_test, ypred_ada) # Inperfect predictions
```

# Module 4: Leave-Project-Out Cross-Validation and Hyperparameter Optimization

#### Listing 30: Code introducing the wrapper functions. Note that we reduced the number of trees to speed up runtime in the upcoming sections.

```{r}
# Boosted stumps and weights
fit <- trainAda(
  replicate ~ .,
  data = train[, voinms],
  # Boosting hyperparameter
  T = 100,
  eta = 1,
  # Tree hyperparameter
  treehypar = treehypar
)

# Predictions on test set
ypred <- testAda(fit, data = test[, prednms])
```

#### Listing 31: Code showing different measures to evaluate the predictive performance of the ensemble.

```{r}
# Confusion matrix
confusion <- table(ypred, y_test)

# Accuracy
sum(diag(confusion)) / sum(confusion)

# Load package
library(AUC)

# Area under the curve
auc(roc(ypred, y_test))

# Base rates
df$replicate_01 <- as.integer(df$replicate) - 1
baserates <- aggregate(replicate_01 ~ pid, df, mean)

# Average base rate
mean(baserates$replicate_01)
```

#### Listing 32: Code to do a leave-project-out cross-validation. Note that this ran for approximately 2 minutes on our laptops.

```{r}
# Partition data by project ID
df_split <- split(df, df$pid)

# List for AUC results
auc <- list()

runtime <- system.time({
  # LPO-CV
  for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
    message("\nThe ", paste0(rp), " model.\n")
    # Training and test set
    test <- df_split[[rp]]
    train <- df[!(df$pid %in% rp), ]
    # Fit model to data
    fit <- trainAda(
      replicate ~ .,
      data = train[, voinms],
      T = 1000,
      eta = 1,
      treehypar = treehypar
    )
    # Predictions (left-out project)
    ypred <- testAda(fit, data = test[, prednms])
    # Performance evaluation
    auc[[rp]] <- auc(roc(ypred, test$replicate))
  }
})

# Waiting time
message("Run time: ", runtime["elapsed"], " s\n")

# Results
c(unlist(auc), "mean" = mean(unlist(auc)))
```

#### Listing 33: Code to define a hypergrid.

```{r}
# Define hypergrid
hgrid <- expand.grid(
  # Tree-specific:
  # Tree depth
  "depth" = 1:3,
  # Boosting specific:
  # Number of trees
  "T" = c(10, 100, 1000),
  # Learning rate
  "eta" = c(1, 0.1),
  # Container for fold performance
  "eerp" = NA,
  "ml1" = NA,
  "ml3" = NA,
  "rpp" = NA,
  "ssrp" = NA,
  "auc.mean" = NA,
  "runtime" = NA
)
```

#### Listing 34: Code to do hyperparameter optimization using a leave-project-out cross-validation strategy. Note that this ran for approximately 3 minutes on our laptops.

```{r}
# Number of rows
hgrid_rows <- nrow(hgrid)

# Partition data by project ID
df_split <- split(df, df$pid)

# Hyperparameter optimization
for (i in seq_len(hgrid_rows)) {
  # Iteration time
  runtime <- system.time({
    # Leave-project-out CV
    for (rp in c("eerp", "ml1", "ml3", "rpp", "ssrp")) {
      # Progress report
      cat(
        "\rProgress: ",
        round(i / hgrid_rows * 100),
        "%",
        " | T: ",
        hgrid$T[i],
        " | eta: ",
        hgrid$eta[i],
        " | depth: ",
        hgrid$depth[i],
        " | model: ",
        rp
      )
      # Training and test set
      test <- df_split[[rp]]
      train <- df[!(df$pid %in% rp), ]
      # Fit model to data
      fit <- trainAda(
        replicate ~ .,
        data = train[, voinms],
        T = hgrid$T[i],
        eta = hgrid$eta[i],
        treehypar = rpart.control(maxdepth = hgrid$depth[i]),
        verbose = FALSE
      )
      # Predictions
      ypred <- testAda(fit, data = test[, prednms], verbose = FALSE)
      # Performance evaluation
      hgrid[[rp]][i] <- auc(roc(ypred, test$replicate))
    }
    # Average performance
    hgrid$auc.mean[i] <- mean(as.numeric(hgrid[i, 4:8]))
  })
  hgrid$runtime[i] <- runtime[["elapsed"]]
} #

# Hypergrid sorted by average AUC
hgrid[order(-hgrid$auc.mean), ]
```

#### Listing 35: Code to fit the final model to the data.

```{r}
# Fit optimized model
fit_final <- trainAda(
  replicate ~ .,
  data = df[, voinms],
  T = 10,
  eta = 1,
  treehypar = rpart.control(maxdepth = 1)
)
```
